### 2024-07-05 paper/paper.tex
updated

\documentclass[a4paper]{article}

% Packages
\usepackage{graphicx}
\usepackage[margin=1in]{geometry}
\usepackage[backend=bibtex]{biblatex}
\usepackage{comment}

\addbibresource{references.bib}

\title{Evaluation of Machine Translation in Languages}
\author{Stefan Liemawan Adji}
\date{\today}

\begin{document}

\maketitle

\section{Objectives}

This paper aims to evaluate existing techniques and circumstances around the task of machine translation across different languages.

\section{Related Work}

\section{Methodology}

Take the list of most spoken language by population according to Wikipedia \cite{wikipedia-list-languages}, then take one for each branch.

\begin{table}[htbp]
    \small
    \centering
    \begin{tabular}{|l|l|r|l|l|}
        \hline
        \textbf{No.} & \textbf{Language} & \textbf{Native speakers} & \textbf{Language family} & \textbf{Branch}   \\
                     &                   & \textbf{(in millions)}   &                          &                   \\
        \hline
        1            & Egyptian Arabic   & 78                       & Afroasiatic              & Semitic           \\
        2            & German            & 76                       & Indo-European            & Germanic          \\
        3            & Hausa             & 54                       & Afroasiatic              & Chadic            \\
        4            & Hindi             & 345                      & Indo-European            & Indo-Aryan        \\
        6            & Japanese          & 123                      & Japonic                  & Japanese          \\
        7            & Javanese          & 68                       & Austronesian             & Malayo-Polynesian \\
        8            & Korean            & 81                       & Koreanic                 & —                 \\
        9            & Mandarin Chinese  & 941                      & Sino-Tibetan             & Sinitic           \\
        5            & Persian           & 62                       & Indo-European            & Iranian           \\
        10           & Russian           & 148                      & Indo-European            & Balto-Slavic      \\
        11           & Spanish           & 486                      & Indo-European            & Romance           \\
        13           & Telugu            & 83                       & Dravidian                & South-Central     \\
        12           & Turkish           & 84                       & Turkic                   & Oghuz             \\
        14           & Vietnamese        & 85                       & Austroasiatic            & Vietic            \\
        \hline
    \end{tabular}
    \caption{List of most spoken languages per branch}
\end{table}

Egyptian Arabic, Javanese, and Telugu might be discarded due to low examples in dataset.

\printbibliography
\end{document}

### 2024-07-05 german/bert.py

@@ -1,0 +1,5 @@
+# Load model directly
+from transformers import AutoModelForMaskedLM, AutoTokenizer
+
+tokenizer = AutoTokenizer.from_pretrained("dbmdz/bert-base-german-uncased")
+model = AutoModelForMaskedLM.from_pretrained("dbmdz/bert-base-german-uncased")

### 2024-07-05 mbart.py

@@ -1,0 +1,7 @@
+# Load model directly
+from transformers import AutoModelForSeq2SeqLM, AutoTokenizer
+
+tokenizer = AutoTokenizer.from_pretrained("facebook/mbart-large-50-many-to-many-mmt")
+model = AutoModelForSeq2SeqLM.from_pretrained(
+    "facebook/mbart-large-50-many-to-many-mmt"
+)

### 2024-07-06 paper/paper.tex
updated

@@
 \usepackage{comment}
+\usepackage{hyperref}
 
@@
 
-\section{Objectives}
+\section{Introduction}
 
-This paper aims to evaluate existing techniques and circumstances around the task of machine translation across different languages.
+According to Ethnologue \cite{ethnologue-2024}, 7,164 languages currently exist and in use today, with 40\% of them considered endangered.
+
+As of July 2024, 243 languages and languages varities are supported by Google Translate (according to Wikipedia \cite{wikipedia-google-translate})
+
+Machine translation (MT) has emerged as a pivotal technology in facilitating cross-cultural communication and global information exchange.
+With the exponential growth of digital content in diverse languages, the demand for accurate and efficient translation systems has never been greater.
+This paper aims to evaluate and compare the performance of various machine translation models across multiple languages, shedding light on their effectiveness and limitations in real-world applications.
+
+Effective machine translation hinges on the ability of algorithms to comprehend and accurately render meaning from source text in one language to target text in another.
+Traditionally, statistical methods dominated the landscape, employing linguistic rules and alignment techniques.
+However, the advent of neural machine translation (NMT) has revolutionized the field, leveraging deep learning architectures to achieve remarkable improvements in translation quality.
+
+Statistical machine translation, which previously dominated MT research for many years with its reliance on various count-based models, has largely been surpassed by neural machine translation (NMT) \cite{stahlberg-2020-nmt-review}.
+
+The languages chosen for evaluation in this study represent a wide overview, striking a balance between resources and diversity.
+
+The evaluation criteria encompass standard metrics such as BLEU (Bilingual Evaluation Understudy) \cite{papieni-2002-bleu}, METEOR (Metric for Evaluation of Translation with Explicit ORdering) \cite{lavie-2007-meteor} to capture nuances in translation quality.
+These metrics not only quantify the fidelity of translations but also offer insights into the models' adaptability and robustness across different linguistic pairs.
+
+Through simple experimentation and analysis, this paper aims to evaluate existing techniques across different languages
+and contribute valuable insights into the current state of machine translation technology.
+By comparing metrics among languages, I hope to inform future researchers of the effectiveness of machine translation on each individual languages that are included in this paper.
+
+Evaluating languages enhances understanding of linguistic structures, phonetics, syntax, semantics, and other linguistic components.
+This knowledge contributes to the broader field of linguistics and helps in developing theories about language evolution, language families, and the cognitive processes involved in language acquisition and use.
+
+Machine translation can be tracked to as early as 1949 \cite{weaver-1999}
 
 \section{Related Work}
+
+As is the case of most fields within Natural Language Processing (NLP), the current state-of-the-art (SOTA) in machine translation (MT) is characterised by the significant advancements brought by large language models (LLMs) like OpenAI's GPT-4.
+These models have surpassed traditional neural machine translation (NMT) systems in several areas, offering more nuanced and accurate translations, especially in handling context and idiomatic expressions.
+
+mBART \cite{liu-2020-mbart} is considered one of the State-of-the-Art as of this moment for multilingual translation.
+
+T5 \cite{raffel-2023-t5}
+
+\section{Tatoeba Dataset}
+
+\subsection{Overview}
+
+Tatoeba is a vast, continuously expanding database consisting sentences and their translations, built through the contributions of thousands of volunteers, offering a tool that allows users to see examples of how words are used in sentences \cite{tatoeba}.
+They currently have 12,132,349 sentences and 423 supported languages, with around one to two thousand new sentences added daily, on average
+
+\begin{figure}[htbp]
+    \centering
+    \includegraphics[width=0.9\linewidth]{images/tatoeba_languages.png}
+    \caption{Tatoeba languages repository with 10,000+ sentences and 100,000+ sentences \cite{tatoeba}}
+    \label{fig:tatoeba_languages}
+\end{figure}
+
+Tatoeba English sentence dataset contains 1,905,089 sentences, the largest one in their repository, with Russian in the second place with 1,066,633 sentences.
+Some of the languages supported in the website is shown in Figure \ref{fig:tatoeba_languages} and Figure \ref{fig:tatoeba_top_bottom_languages}, sorted from the biggest corpus.
+Low-resource languages such as Rendille, Southern Haida, and Cuyonon can be seen at the bottom of the list, having only a single sentence example.
+Ancestor languages such as Old Saxon and Old Turkish can also be seen in the list, subsequently with low number of examples.
+
+\begin{figure}[htbp]
+    \centering
+    \includegraphics[width=0.5\linewidth]{images/tatoeba_top_20_lang.png}
+    \includegraphics[width=0.46\linewidth]{images/tatoeba_bottom_20_lang.png}
+    \caption{Tatoeba top 20 and bottom 20 languages based on sentences count \cite{tatoeba}}
+    \label{fig:tatoeba_top_bottom_languages}
+\end{figure}
+
+\subsection{Analysis}
+
+Sentences typically consist of everyday phrases such as 'I have to go to sleep', 'That is intriguing', and 'Where do you live?' They may also include single-word exclamations like 'Speak!' or 'Look!' Additionally, multiple sentences such as 'You may write in any language you want.
+On Tatoeba, all languages are considered equal', and 'Guns don't kill people.
+People kill people' can be found inside the corpus.
+A few of them also include human names, 'Compare your answer with Tom's', 'Muiriel is 20 now'.
+All of the sentences are straightforward and literal, without the use of linguistic devices such as metaphors or sarcasm.
+Therefore, machine translation process should be straightforward on this level.
 
@@
 
+Sentences in English are downloaded, 1,898,494 sentences (it is unclear why it is less than the number stated in the Tatoeba website).
+Then for each langauges, download sentence pairs compared to English.
+Merge every language sentences into one big dataframe, only keep where sentences exist for every language.
+
+Target language will be English, with machine translation task defined as Many-to-English.
+
 \begin{table}[htbp]
-    \small
     \centering
-    \begin{tabular}{|l|l|r|l|l|}
+    \begin{tabular}{|l|l|}
         \hline
-        \textbf{No.} & \textbf{Language} & \textbf{Native speakers} & \textbf{Language family} & \textbf{Branch}   \\
-                     &                   & \textbf{(in millions)}   &                          &                   \\
+        \textbf{No.} & \textbf{Language} \\
         \hline
-        1            & Egyptian Arabic   & 78                       & Afroasiatic              & Semitic           \\
-        2            & German            & 76                       & Indo-European            & Germanic          \\
-        3            & Hausa             & 54                       & Afroasiatic              & Chadic            \\
-        4            & Hindi             & 345                      & Indo-European            & Indo-Aryan        \\
-        6            & Japanese          & 123                      & Japonic                  & Japanese          \\
-        7            & Javanese          & 68                       & Austronesian             & Malayo-Polynesian \\
-        8            & Korean            & 81                       & Koreanic                 & —                 \\
-        9            & Mandarin Chinese  & 941                      & Sino-Tibetan             & Sinitic           \\
-        5            & Persian           & 62                       & Indo-European            & Iranian           \\
-        10           & Russian           & 148                      & Indo-European            & Balto-Slavic      \\
-        11           & Spanish           & 486                      & Indo-European            & Romance           \\
-        13           & Telugu            & 83                       & Dravidian                & South-Central     \\
-        12           & Turkish           & 84                       & Turkic                   & Oghuz             \\
-        14           & Vietnamese        & 85                       & Austroasiatic            & Vietic            \\
+        1            & Dutch             \\
+        2            & Finnish           \\
+        3            & French            \\
+        4            & German            \\
+        5            & Hebrew            \\
+        6            & Hungarian         \\
+        7            & Italian           \\
+        8            & Japanese          \\
+        9            & Mandarin Chinese  \\
+        10           & Polish            \\
+        11           & Portuguese        \\
+        12           & Russian           \\
+        13           & Spanish           \\
+        14           & Turkish           \\
+        15           & Ukrainian         \\
         \hline
     \end{tabular}
-    \caption{List of most spoken languages per branch}
+    \caption{List of chosen languages for evaluation}
 \end{table}
 
-Egyptian Arabic, Javanese, and Telugu might be discarded due to low examples in dataset.
+Sentences dataset from Tatoeba is used \footnote{\url{https://tatoeba.org/}}.
+Languages that has more than fifty thousand sentences are selected.
+Accordingly, languages that are available for mbart is also selected based on the list here \footnote{\url{https://dl-translate.readthedocs.io/en/latest/available_languages/}}
+
+For translation, all languages are translated into English as the target language. Then compare the true English sentence and the predicted one, calculated BLEU.
+
+\section{Evaluation}
+
+\section{Conclusion}
 

### 2024-07-06 analyse/analyse_sentences.py

@@ -1,0 +1,14 @@
+from collections import Counter
+
+import matplotlib.pyplot as plt
+import nltk
+import pandas as pd
+import spacy
+from textblob import TextBlob
+from wordcloud import WordCloud
+
+sentences_df = pd.read_csv("../dataset/sentences_v1.csv", index_col=0)
+
+print(sentences_df)
+
+english = sentences_df["english"].tolist()

### 2024-07-06 german/bert.py

@@ -1,5 +1,0 @@
-# Load model directly
-from transformers import AutoModelForMaskedLM, AutoTokenizer
-
-tokenizer = AutoTokenizer.from_pretrained("dbmdz/bert-base-german-uncased")
-model = AutoModelForMaskedLM.from_pretrained("dbmdz/bert-base-german-uncased")

### 2024-07-06 handle_tatoeba.py

@@ -1,0 +1,110 @@
+import pandas as pd
+from tqdm import tqdm
+
+
+def read_df(language, tsv_path):
+    df = pd.read_csv(
+        tsv_path,
+        sep="\t",
+        header=None,
+        on_bad_lines="skip",
+    )
+    df.columns = ["index", "english", "index_target", language]
+    df.drop(columns="index_target", inplace=True)
+
+    return df
+
+
+path = "original_dataset/tatoeba"
+
+language_list = [
+    (
+        "dutch",
+        f"{path}/Sentence pairs in English-Dutch - 2024-07-05.tsv",
+    ),
+    (
+        "finnish",
+        f"{path}/Sentence pairs in English-Finnish - 2024-07-05.tsv",
+    ),
+    (
+        "french",
+        f"{path}/Sentence pairs in English-French - 2024-07-05.tsv",
+    ),
+    (
+        "german",
+        f"{path}/Sentence pairs in English-German - 2024-07-05.tsv",
+    ),
+    (
+        "hebrew",
+        f"{path}/Sentence pairs in English-Hebrew - 2024-07-05.tsv",
+    ),
+    (
+        "hungarian",
+        f"{path}/Sentence pairs in English-Hungarian - 2024-07-05.tsv",
+    ),
+    (
+        "italian",
+        f"{path}/Sentence pairs in English-Italian - 2024-07-05.tsv",
+    ),
+    (
+        "japanese",
+        f"{path}/Sentence pairs in English-Japanese - 2024-07-05.tsv",
+    ),
+    (
+        "mandarin",
+        f"{path}/Sentence pairs in English-Mandarin Chinese - 2024-07-05.tsv",
+    ),
+    (
+        "polish",
+        f"{path}/Sentence pairs in English-Polish - 2024-07-05.tsv",
+    ),
+    (
+        "portuguese",
+        f"{path}/Sentence pairs in English-Portuguese - 2024-07-05.tsv",
+    ),
+    (
+        "russian",
+        f"{path}/Sentence pairs in English-Russian - 2024-07-05.tsv",
+    ),
+    (
+        "spanish",
+        f"{path}/Sentence pairs in English-Spanish - 2024-07-05.tsv",
+    ),
+    (
+        "turkish",
+        f"{path}/Sentence pairs in English-Turkish - 2024-07-05.tsv",
+    ),
+    (
+        "ukrainian",
[…]

### 2024-07-06 mbart.py

@@ -1,7 +1,65 @@
-# Load model directly
-from transformers import AutoModelForSeq2SeqLM, AutoTokenizer
+import os
 
-tokenizer = AutoTokenizer.from_pretrained("facebook/mbart-large-50-many-to-many-mmt")
-model = AutoModelForSeq2SeqLM.from_pretrained(
-    "facebook/mbart-large-50-many-to-many-mmt"
-)
+import pandas as pd
+import torch
+from tqdm import tqdm
+from transformers import MBart50Tokenizer, MBartForConditionalGeneration, MBartTokenizer
+
+os.environ["PYTORCH_ENABLE_MPS_FALLBACK"] = "1"
+
+MODEL_NAME = "facebook/mbart-large-50-many-to-many-mmt"
+
+MBART50_LANG_CODES = {
+    # "dutch": "nl_XX",
+    # "finnish": "fi_FI",
+    "french": "fr_XX",
+    "german": "de_DE",
+    "hebrew": "he_IL",
+    "italian": "it_IT",
+    "japanese": "ja_XX",
+    "mandarin": "zh_CN",
+    "polish": "pl_PL",
+    "portuguese": "pt_XX",
+    "russian": "ru_RU",
+    "spanish": "es_XX",
+    "turkish": "tr_TR",
+    "ukrainian": "uk_UA",
+}
+
+tqdm.pandas()
+
+model = MBartForConditionalGeneration.from_pretrained(MODEL_NAME)
+tokeniser = MBart50Tokenizer.from_pretrained(MODEL_NAME)
+
+df = pd.read_csv("dataset/sentences_v1.csv", index_col=0)
+
+print(df)
+
+
+def translate_batch(batch):
+    inputs = tokeniser(batch, return_tensors="pt", padding=True, truncation=True)
+    with torch.no_grad():
+        translated_tokens = model.generate(
+            **inputs, forced_bos_token_id=tokeniser.lang_code_to_id["en_XX"]
+        )
+    translations = tokeniser.batch_decode(translated_tokens, skip_special_tokens=True)
+
+    return translations
+
+
+batch_size = 8
+translated_df = df[["index", "english"]].copy()
+
+for src_lang in MBART50_LANG_CODES.keys():
+    print(f"Translating {src_lang}...")
+    translated_texts = []
+
+    for i in tqdm(range(0, len(df), batch_size)):
+        batch = df[src_lang][i : i + batch_size].tolist()
+        translations = translate_batch(batch)
[…]

### 2024-07-06 merge.py

@@ -1,0 +1,10 @@
+import pandas as pd
+
+dutch_df = pd.read_csv("result/v1/mbart50/translated_from_dutch.csv")
+finnish_df = pd.read_csv("result/v1/mbart50/translated_from_finnish.csv")
+df = finnish_df.merge(dutch_df)
+df.drop(columns=["Unnamed: 0"], inplace=True)
+
+print(df)
+
+df.to_csv("result/v1/mbart50/translated_1.csv")

### 2024-07-06 utils.py

@@ -1,0 +1,2 @@
+def tokenise():
+    ...

### 2024-07-06 paper/paper.tex
updated

@@
 
-Machine translation (MT) has emerged as a pivotal technology in facilitating cross-cultural communication and global information exchange.
-With the exponential growth of digital content in diverse languages, the demand for accurate and efficient translation systems has never been greater.
-This paper aims to evaluate and compare the performance of various machine translation models across multiple languages, shedding light on their effectiveness and limitations in real-world applications.
-
-Effective machine translation hinges on the ability of algorithms to comprehend and accurately render meaning from source text in one language to target text in another.
-Traditionally, statistical methods dominated the landscape, employing linguistic rules and alignment techniques.
-However, the advent of neural machine translation (NMT) has revolutionized the field, leveraging deep learning architectures to achieve remarkable improvements in translation quality.
+Machine translation (MT) can be traced down to as early as 1949 \cite{weaver-1999}, recently growing as a popular research field within Natural Language Processing (NLP).
 
 Statistical machine translation, which previously dominated MT research for many years with its reliance on various count-based models, has largely been surpassed by neural machine translation (NMT) \cite{stahlberg-2020-nmt-review}.
-
-The languages chosen for evaluation in this study represent a wide overview, striking a balance between resources and diversity.
-
-The evaluation criteria encompass standard metrics such as BLEU (Bilingual Evaluation Understudy) \cite{papieni-2002-bleu}, METEOR (Metric for Evaluation of Translation with Explicit ORdering) \cite{lavie-2007-meteor} to capture nuances in translation quality.
-These metrics not only quantify the fidelity of translations but also offer insights into the models' adaptability and robustness across different linguistic pairs.
+Nowadays, Large Language Models (LLMs) such as mBart \cite{liu-2020-mbart}, M2M-100 \cite{fan-2020-m2m100}, NLLB-200 \cite{nllb200}, T5 \cite{raffel-2023-t5}, and GPT-4 \cite{openai-2024-gpt4-technical-report} dominated the field of machine translation.
+DeepL \cite{deepl} currently proclaimed itself as 'The world most accurate translator'.
 
@@
 By comparing metrics among languages, I hope to inform future researchers of the effectiveness of machine translation on each individual languages that are included in this paper.
-
-Evaluating languages enhances understanding of linguistic structures, phonetics, syntax, semantics, and other linguistic components.
-This knowledge contributes to the broader field of linguistics and helps in developing theories about language evolution, language families, and the cognitive processes involved in language acquisition and use.
-
-Machine translation can be tracked to as early as 1949 \cite{weaver-1999}
 
@@
 
-mBART \cite{liu-2020-mbart} is considered one of the State-of-the-Art as of this moment for multilingual translation.
-
-T5 \cite{raffel-2023-t5}
-
-\section{Tatoeba Dataset}
+\section{Dataset}
 
@@
 They currently have 12,132,349 sentences and 423 supported languages, with around one to two thousand new sentences added daily, on average
+
+To build the dataset,
+
+Sentences in English are downloaded, 1,898,494 sentences (it is unclear why it is less than the number stated in the Tatoeba website).
+Then for each langauges, download sentence pairs compared to English.
+Merge every language sentences into one big dataframe, only keep where sentences exist for every language
 
@@
 
-\section{Methodology}
+\section{Methodology}.
 
-Take the list of most spoken language by population according to Wikipedia \cite{wikipedia-list-languages}, then take one for each branch.
-
-Sentences in English are downloaded, 1,898,494 sentences (it is unclear why it is less than the number stated in the Tatoeba website).
-Then for each langauges, download sentence pairs compared to English.
-Merge every language sentences into one big dataframe, only keep where sentences exist for every language.
-
-Target language will be English, with machine translation task defined as Many-to-English.
+The languages chosen for evaluation in this study represent a wide overview, striking a balance between resources and diversity.
+Each language will then be translated into English (Many-to-English) and evaluated.
+The evaluation criteria encompass standard metrics such as BLEU (Bilingual Evaluation Understudy) \cite{papieni-2002-bleu}, METEOR (Metric for Evaluation of Translation with Explicit ORdering) \cite{lavie-2007-meteor} to capture nuances in translation quality.
+These metrics not only quantify the fidelity of translations but also offer insights into the models' adaptability and robustness across different linguistic pairs.
 

### 2024-07-06 mbart.py

@@ -63,3 +63,5 @@ for src_lang in MBART50_LANG_CODES.keys():
 
     print(translated_df)
     translated_df.to_csv("result/v1/mbart50/translated.csv")
+
+# seems like some language still translated to non_english?

### 2024-07-06 merge.py

@@ -1,9 +1,11 @@
 import pandas as pd
 
-dutch_df = pd.read_csv("result/v1/mbart50/translated_from_dutch.csv")
-finnish_df = pd.read_csv("result/v1/mbart50/translated_from_finnish.csv")
-df = finnish_df.merge(dutch_df)
-df.drop(columns=["Unnamed: 0"], inplace=True)
+langs_df = pd.read_csv("result/v1/mbart50/translated.csv")
+dutch_fin_df = pd.read_csv("result/v1/mbart50/translated_dutch_finnish.csv")
+print(langs_df)
+df = dutch_fin_df.merge(langs_df, how="outer", on=["index", "english"])
+
+df.drop(columns=["Unnamed: 0_x", "Unnamed: 0_y"], inplace=True)
 
 print(df)
 

### 2024-07-06 utils.py

@@ -1,2 +1,41 @@
-def tokenise():
-    ...
+import nltk
+import sacrebleu
+
+# Ensure you have the METEOR data for nltk
+nltk.download("wordnet")
+nltk.download("omw-1.4")
+nltk.download("punkt")
+
+from nltk.translate.meteor_score import meteor_score
+
+# Example translations and references
+references_list = [
+    ["The cat is on the mat.", "There is a cat on the mat."],
+    ["Hello world!", "Hi world!"],
+]
+
+candidates = ["The cat is on the mat.", "Hello world!"]
+
+# Calculate BLEU Score
+bleu_scores = []
+for candidate, references in zip(candidates, references_list):
+    bleu = sacrebleu.corpus_bleu([candidate], [[ref] for ref in references])
+    bleu_scores.append(bleu.score)
+
+average_bleu = sum(bleu_scores) / len(bleu_scores)
+print(f"Average BLEU score: {average_bleu}")
+
+# Calculate METEOR Score
+
+tokenized_references = [
+    [nltk.word_tokenize(ref) for ref in refs] for refs in references_list
+]
+tokenized_candidates = [nltk.word_tokenize(candidate) for candidate in candidates]
+
+meteor_scores = []
+for candidate, references in zip(tokenized_candidates, tokenized_references):
+    meteor = meteor_score(references, candidate)
+    meteor_scores.append(meteor)
+
+average_meteor = sum(meteor_scores) / len(meteor_scores)
+print(f"Average METEOR score: {average_meteor}")

### 2024-07-13 paper/paper.tex
updated paper

@@
 
+Machine translation (MT) is the task of automatically translating from one language to another.
+This can be done through text or audio.
+Ideally, machine translation should be able to conserve the contextual, idiomatic and pragmatic nuances of both languages.
+
+Machine translation can be traced down to as early as 1949, \cite{weaver-1999}, with the first public demonstration of an MT system happened on 7th January 1954 through a collaboration with IBM \cite{hutchins-2006-first-mt}.
+A meticulously chosen set of 49 Russian sentences was translated into English, utilising a highly limited vocabulary of 250 words and only 6 grammar rules.
+While this had minimal scientific impact, it was impressive enough to encourage substantial funding for MT research in the USA.
+
+The decade of high expectation and disillusion, 1956-1966 \cite{hutchins-2001-mt-50-years}.
+
+The 1960s, particularly 1967-1976 is dubbed as 'the quiet decade' for machine translation by Hutchins \cite{hutchins-2001-mt-50-years}
+
+Rule-based MT relied on extensive linguistic rules and bilingual dictionaries. They were based on syntactic and grammatical rules to parse and generate translations.
+
+Since 1989, the dominance of the rule-based approach has been challenged by the rise of new methods and strategies, collectively referred to as ‘corpus-based’ methods \cite{hutchins-1994-research-methods-mt,hutchins-1998-development-mt}.
+Subsequently, statistics-based approaches for MT re-emerged, bolstered by the recent success of probabilistic techniques in
+speech recognition.
+
+1980s: Statistical Machine Translation (SMT) Emerges, use statistical models, probability
+
+Late 2010s: Rise of Neural Networks
+Neural Machine Translation (NMT): Starting around 2014, neural networks began to dominate MT.
+NMT systems, like those using the sequence-to-sequence (Seq2Seq) model with attention mechanisms (e.g., the Transformer model), demonstrated significant improvements in translation quality.
+Advantages Over SMT: NMT models often produce more fluent and coherent translations as they consider the entire sentence context, rather than breaking it down into smaller units like phrases.
+
+...recently growing as a popular research field within Natural Language Processing (NLP).
+Statistical machine translation \cite{lopez-2008-smt}, which previously dominated for many years in the early 2000s with its reliance on various count-based models, has largely been surpassed by neural machine translation (NMT) in (year).
+Since the introduction of Transformers in 2017 \cite{vaswani-2017-attention}, Natural Language Processing (NLP) and machine translation in particular reached a giant milestone.
+The following years saw the birth of Large Language Models (LLMs) such as BERT \cite{devlin-2019-bert} and GPT \cite{openai-2024-gpt4}.
+
+Difference between LLMs and MT engines?
+MT engines are specifically designed to translate text from one language to another while preserving the meaning and context as accurately as possible.
+MT engines can also be based on transformer architectures, but they are usually fine-tuned and optimized for the task of translation.
+They often incorporate additional components like alignment models and post-processing steps to ensure translation quality.
+
+Primary Function: LLMs focus on generating and understanding natural language text, whereas MT engines focus exclusively on translating text between languages.
+Training Data: LLMs are trained on diverse text data from various sources, while MT engines are trained specifically on parallel corpora (texts in two or more languages that are translations of each other).
+Output: LLMs generate coherent text in the same language as the input, often in a more creative or conversational style.
+MT engines produce translated text in a different language, aiming for accuracy and fidelity to the original meaning.
+
 According to Ethnologue \cite{ethnologue-2024}, 7,164 languages currently exist and in use today, with 40\% of them considered endangered.
 
-As of July 2024, 243 languages and languages varities are supported by Google Translate (according to Wikipedia \cite{wikipedia-google-translate})
+As of July 2024, 243 languages are supported by Google Translate (according to Wikipedia \cite{wikipedia-google-translate})
 
-Machine translation (MT) can be traced down to as early as 1949 \cite{weaver-1999}, recently growing as a popular research field within Natural Language Processing (NLP).
+Many language models are trained on predominantly English NLP tasks, before fine-tuned for downstream tasks in other languages.
 
-Statistical machine translation, which previously dominated MT research for many years with its reliance on various count-based models, has largely been surpassed by neural machine translation (NMT) \cite{stahlberg-2020-nmt-review}.
-Nowadays, Large Language Models (LLMs) such as mBart \cite{liu-2020-mbart}, M2M-100 \cite{fan-2020-m2m100}, NLLB-200 \cite{nllb200}, T5 \cite{raffel-2023-t5}, and GPT-4 \cite{openai-2024-gpt4-technical-report} dominated the field of machine translation.
-DeepL \cite{deepl} currently proclaimed itself as 'The world most accurate translator'.
+There is little insights on how the current state of machine translation is across different language. Therefore...
+
+There does not seem to be a lot of pre-trained models that focuses on a single language. This could be because multilingual LLMs perform better? find proof
+
+Focus on single-language models?
 
@@
 
-\section{Related Work}
+\section{Literature Review}
 
-As is the case of most fields within Natural Language Processing (NLP), the current state-of-the-art (SOTA) in machine translation (MT) is characterised by the significant advancements brought by large language models (LLMs) like OpenAI's GPT-4.
-These models have surpassed traditional neural machine translation (NMT) systems in several areas, offering more nuanced and accurate translations, especially in handling context and idiomatic expressions.
+DeepL \cite{deepl} is a machine translation service that has gained widespread acclaim, currently proclaimed itself as the world's most accurate translator.
+There is not much information on how the model or techniques as it is essentially a private business.
+(irrelevant?)
 
-\section{Dataset}
+Instruction Learning and decoder-based models...
 
-\subsection{Overview}
+GPT \cite{openai-2024-gpt4}
+
+Multilingual Language Models are on the rise for the task of machine translation that involves several languages
+
+A major drawback within most multilingual LMs nowadays is that they are either trained, tested, prioritised on English first.
+
+mBERT \cite{devlin-2019-bert} is the multilingual version of BERT, introduced in the same paper, trained on 104 different languages.
+However, the model been shown to suffer on low-resource languages \cite{wu-2020-mbert-are-all}.
+
+M-BERT
+does create multilingual representations,
+
+T5 \cite{raffel-2023-t5} is a Text-to-Text Transfer Transformer model trained on English.
+
+mt5 \cite{xue-2021-mt5} is a multilingual variant of T5 \cite{raffel-2023-t5}, covering a total of 101 languages.
+
+PolyLM \cite{wei-2023-polylm}
+
+mBart \cite{liu-2020-mbart}
+
+NLLB \cite{nllb200}
+
+MistralAI \cite{jiang-2023-mistral-7b}
+
+Massive Parallel Corpora: The availability of large parallel corpora (datasets containing aligned text pairs in different languages) has been crucial for training effective translation models.
+Datasets like Europarl, OpenSubtitles, and Common Crawl provide a wealth of parallel text data.
+Unsupervised Translation: Techniques have been developed to perform translation without the need for large parallel corpora, using monolingual data to train models that can translate between languages.
+This is especially useful for low-resource languages.
+
+Multilingual and Zero-Shot Translation:
+
+Multilingual NMT Models: Models like Google's multilingual NMT and Facebook's M2M-100 can translate directly between many language pairs without relying on an intermediate language (like English).
+These models are trained on datasets containing multiple languages simultaneously.
+Zero-Shot Translation: These models can sometimes translate between language pairs they were not explicitly trained on, leveraging knowledge from related languages or shared linguistic features.
+
+\section{Experiments}
+
+\subsection{Dataset}
 
@@
 
-\subsection{Analysis}
-
 Sentences typically consist of everyday phrases such as 'I have to go to sleep', 'That is intriguing', and 'Where do you live?' They may also include single-word exclamations like 'Speak!' or 'Look!' Additionally, multiple sentences such as 'You may write in any language you want.
@@
 Therefore, machine translation process should be straightforward on this level.
-
-\section{Methodology}.
 
@@
 
+GPT-4 \cite{openai-2024-gpt4} try GPT and see
+
+READ THIS \cite{intento-2020} (exactly what this paper should do, maybe compare results and find insights)
+
 \section{Evaluation}

### 2024-07-13 nllb.py
updated code

@@ -3,48 +3,47 @@ import os
 import pandas as pd
 import torch
 from tqdm import tqdm
-from transformers import MBart50Tokenizer, MBartForConditionalGeneration, MBartTokenizer
+from transformers import AutoModelForSeq2SeqLM, AutoTokenizer
 
 os.environ["PYTORCH_ENABLE_MPS_FALLBACK"] = "1"
 
-MODEL_NAME = "facebook/mbart-large-50-many-to-many-mmt"
+MODEL_NAME = "facebook/nllb-200-3.3B"
 
 MBART50_LANG_CODES = {
-    "dutch": "nl_XX",
-    "finnish": "fi_FI",
-    "french": "fr_XX",
-    "german": "de_DE",
+    "dutch": "nld_Latn",
+    "finnish": "fin_Latn",
+    "french": "fra_Latn",
+    "german": "deu_Latn",
     "hebrew": "he_IL",
-    "italian": "it_IT",
-    "japanese": "ja_XX",
-    "chinese": "zh_CN",
-    "polish": "pl_PL",
-    "portuguese": "pt_XX",
-    "russian": "ru_RU",
-    "spanish": "es_XX",
-    "turkish": "tr_TR",
-    "ukrainian": "uk_UA",
+    "italian": "ita_Latn",
+    "japanese": "jpn_Jpan",
+    "chinese": "zho_Hans",
+    "polish": "pol_Latn",
+    "portuguese": "por_Latn",
+    "russian": "rus_Cyrl",
+    "spanish": "spa_Latn",
+    "turkish": "tur_Latn",
+    "ukrainian": "ukr_Cyrl",
 }
 
 tqdm.pandas()
 
-model = MBartForConditionalGeneration.from_pretrained(MODEL_NAME)
+tokeniser = AutoTokenizer.from_pretrained(MODEL_NAME)
+model = AutoModelForSeq2SeqLM.from_pretrained(MODEL_NAME)
 
 df = pd.read_csv("dataset/sentences_v1.csv", index_col=0)
 
 
-def translate_batch(tokeniser, batch_input):
-    inputs = tokeniser(
-        batch_input,
-        return_tensors="pt",
-        padding=True,
-        truncation=True,
-    )
+def translate_batch(batch):
+    inputs = tokeniser(batch, return_tensors="pt", padding=True, truncation=True)
     with torch.no_grad():
         translated_tokens = model.generate(
-            **inputs, forced_bos_token_id=tokeniser.lang_code_to_id["en_XX"]
[…]

### 2024-07-29 paper/paper.tex
updated code
updated, added opus

--- 2024-07-06 paper/paper.tex
@@
 
-\title{Evaluation of Machine Translation in Languages}
+\title{Evaluation of Pre-Trained Models for Translation in Different Languages}
 \author{Stefan Liemawan Adji}
@@
 According to Ethnologue \cite{ethnologue-2024}, 7,164 languages currently exist and in use today, with 40\% of them considered endangered.
+As of July 2024, 243 languages are supported by Google Translate (according to Wikipedia \cite{wikipedia-google-translate}).
+In modern times, the need for translation services has surged due to the growing exchange of information across different regions that speak various languages \cite{okpor-2014-machine-ta}.
 
-As of July 2024, 243 languages and languages varities are supported by Google Translate (according to Wikipedia \cite{wikipedia-google-translate})
+Machine translation (MT) is the task of automatically translating from one language to another.
+This can be done through text or audio.
+It can be traced back to 1949 \cite{weaver-1999}, with the first public demonstration of an MT system on January 7, 1954, in collaboration with IBM, where 49 Russian sentences were translated into English using a limited vocabulary of 250 words and 6 grammar rules \cite{hutchins-2006-first-mt}.
+However, over the next several decades, growth were limited for machine translation, with 1956-1966 considered the decade of high expectation and disillusion, and 1967-1976 dubbed 'the quiet decade' \cite{hutchins-2001-mt-50-years}.
+Then in 1989, the dominance of the rule-based approach has been challenged by the rise of new methods and strategies, collectively referred to as ‘corpus-based’ methods (data-driven) \cite{hutchins-1994-research-methods-mt,hutchins-1998-development-mt}.
+Subsequently, statistics-based approaches for MT re-emerged, bolstered by the recent success of probabilistic techniques in speech recognition.
+Statistical machine translation \cite{lopez-2008-smt} dominated the domain between late 1990s through the early 2010s, before largely being surpassed by neural machine translation (NMT) \cite{cho-2014-properties,sutskever-2014-seq2seq}.
+Finally, since the introduction of Transformers in 2017 \cite{vaswani-2017-attention}, Natural Language Processing (NLP) and machine translation in particular reached a giant milestone.
+The following years saw the birth of Large Language Models (LLMs) such as BERT \cite{devlin-2019-bert}, GPT \cite{openai-2024-gpt4}, and T5 \cite{raffel-2023-t5}, which revolutionised both MT and the whole field of NLP.
+Soon after, several pre-trained models that are specifically designed for machine translation emerged, namely mBART \cite{liu-2020-mbart}, mT5 \cite{xue-2021-mt5}, and NLLB \cite{nllb200}.
 
-Machine translation (MT) can be traced down to as early as 1949 \cite{weaver-1999}, recently growing as a popular research field within Natural Language Processing (NLP).
+Since LLMs require vast amounts of training data, the availability of parallel corpora is crucial for building effective models.
+The lack of extensive parallel corpora, especially for low-resource languages, leads to suboptimal performance in NMT techniques compared to their high-resource counterparts \cite{ranathunga-2023-nmt-low-res}.
 
-Statistical machine translation, which previously dominated MT research for many years with its reliance on various count-based models, has largely been surpassed by neural machine translation (NMT) \cite{stahlberg-2020-nmt-review}.
-Nowadays, Large Language Models (LLMs) such as mBart \cite{liu-2020-mbart}, M2M-100 \cite{fan-2020-m2m100}, NLLB-200 \cite{nllb200}, T5 \cite{raffel-2023-t5}, and GPT-4 \cite{openai-2024-gpt4-technical-report} dominated the field of machine translation.
-DeepL \cite{deepl} currently proclaimed itself as 'The world most accurate translator'.
+OPUS \cite{opus} is a comprehensive collection of open-source parallel corpora used extensively in the field of machine translation (MT).
+It includes corpora for 744 languages and contains over 1,210 different datasets, amassing a total of 45,945,946,108 sentence pairs.
+Tatoeba \cite{tatoeba} is another prominent resource in the field of MT and NLP, known for its extensive collection of translated sentences.
+As of July 2024, it contains 12,186,207 sentences over 423 supported languages, growing daily through volunteer contributions.
 
-Through simple experimentation and analysis, this paper aims to evaluate existing techniques across different languages
+pre-trained models...
+
+Through simple experimentation and analysis, this paper aims to evaluate existing pre-trained models across different languages
 and contribute valuable insights into the current state of machine translation technology.
-By comparing metrics among languages, I hope to inform future researchers of the effectiveness of machine translation on each individual languages that are included in this paper.
 
-\section{Related Work}
+\section{Literature Review}
 
-As is the case of most fields within Natural Language Processing (NLP), the current state-of-the-art (SOTA) in machine translation (MT) is characterised by the significant advancements brought by large language models (LLMs) like OpenAI's GPT-4.
-These models have surpassed traditional neural machine translation (NMT) systems in several areas, offering more nuanced and accurate translations, especially in handling context and idiomatic expressions.
+\subsection{Neural Machine Translation (NMT)}
 
-\section{Dataset}
+The encoder-decoder approach \cite{cho-2014-properties} remains as the foundation architecture for many sequence-to-sequence models in natural language processing, especially machine translation.
 
-\subsection{Overview}
+\subsection{One-to-one Models}
+
+One-to-One Translation refers to a translation approach where a model is specifically trained to translate between one source language and one target language.
+This setup is characterised by having a dedicated model for each unique language pair.
+
+The OPUS-MT by Helsinki-NLP \cite{tiedemann-2020-opus-mt} is a good example of one-to-one translation project, they provide over 1,000 pre-trained translation models for many language pairs, allowing users to perform high-quality translations without needing to train models from scratch.
+
+OpenNMT \cite{klein-etal-2017-opennmt} is another framework that allows for one-to-one translation, although their pre-trained models only support English and German languages.
+
+\subsection{Many-to-many Models}
+
+With the rise of LLMs and pre-trained language models, multilingual machine translation models have received popularity, particularly many-to-many translation.
+
+Multilingual neural machine translation (NMT) allows for the training of a single model capable of translating between multiple source and target languages \cite{aharoni-2019-massively-multilingual}.
+
+Many-to-many translation refers to the capability of a model to translate text between multiple language pairs bidirectionally.
+This means the model can handle translations from any of the supported source languages to any of the supported target languages, not limited to a single source or target language.
+
+mBart \cite{liu-2020-mbart} (50 languages), NLLB-200 \cite{nllb200} (200 language), PolyLM \cite{wei-2023-polylm} (32 languages), and mt5 \cite{xue-2021-mt5} (101 languages) are some of the most popular many-to-many models.
+
+\subsection{Sequence-to-Sequence Models}
+
+GPT \cite{openai-2024-gpt4}
+
+mBERT \cite{devlin-2019-bert} is the multilingual version of BERT, introduced in the same paper, trained on 104 different languages.
+However, the model been shown to suffer on low-resource languages \cite{wu-2020-mbert-are-all}.
+
+T5 \cite{raffel-2023-t5} is a Text-to-Text Transfer Transformer model trained on English.
+
+MistralAI \cite{jiang-2023-mistral-7b}
+
+\section{Experiments}
+
+\subsection{Dataset}
 
@@
 
+Tatoeba is chosen due to its sentence having more daily life sentences...?
+
 To build the dataset,
@@
 Sentences in English are downloaded, 1,898,494 sentences (it is unclear why it is less than the number stated in the Tatoeba website).
-Then for each langauges, download sentence pairs compared to English.
+Then for each languages, download sentence pairs compared to English.
 Merge every language sentences into one big dataframe, only keep where sentences exist for every language

### 2024-07-29 handle_tatoeba.py

@@ -19,6 +19,10 @@ path = "original_dataset/tatoeba"
 
 language_list = [
     (
+        "chinese",
+        f"{path}/Sentence pairs in English-Mandarin Chinese - 2024-07-05.tsv",
+    ),
+    (
         "dutch",
         f"{path}/Sentence pairs in English-Dutch - 2024-07-05.tsv",
     ),
@@ -50,10 +54,10 @@ language_list = [
         "japanese",
         f"{path}/Sentence pairs in English-Japanese - 2024-07-05.tsv",
     ),
-    (
-        "mandarin",
-        f"{path}/Sentence pairs in English-Mandarin Chinese - 2024-07-05.tsv",
-    ),
+    # {
+    #     "lithuanian",
+    #     f"{path}/Sentence pairs in English-Lithuanian - 2024-07-29.tsv",
+    # },
     (
         "polish",
         f"{path}/Sentence pairs in English-Polish - 2024-07-05.tsv",
@@ -107,4 +111,4 @@ for language, tsv_path in tqdm(language_list):
 main_df.dropna(inplace=True)
 print(main_df)
 
-main_df.to_csv(f"dataset/sentences_v1_1.csv", index=False)
+main_df.to_csv(f"dataset/sentences_v3.csv", index=False)

### 2024-07-29 mbart.py

@@ -10,6 +10,7 @@ os.environ["PYTORCH_ENABLE_MPS_FALLBACK"] = "1"
 MODEL_NAME = "facebook/mbart-large-50-many-to-many-mmt"
 
 MBART50_LANG_CODES = {
+    "chinese": "zh_CN",
     "dutch": "nl_XX",
     "finnish": "fi_FI",
     "french": "fr_XX",
@@ -17,9 +18,8 @@ MBART50_LANG_CODES = {
     "hebrew": "he_IL",
     "italian": "it_IT",
     "japanese": "ja_XX",
-    "chinese": "zh_CN",
     "polish": "pl_PL",
-    "portuguese": "pt_XX",
+    # "portuguese": "pt_XX",
     "russian": "ru_RU",
     "spanish": "es_XX",
     "turkish": "tr_TR",
@@ -49,7 +49,7 @@ def translate_batch(tokeniser, batch_input):
     return translations
 
 
-batch_size = 8
+batch_size = 4
 translated_df = df[["index", "english"]].copy()
 
 for src_lang, src_lang_code in MBART50_LANG_CODES.items():

### 2024-07-29 nllb.py

@@ -48,7 +48,8 @@ def translate_batch(batch):
     return translations
 
 
-batch_size = 8
+batch_size = 1
+# batch_size = 8
 translated_df = df[["index", "english"]].copy()
 
 for src_lang, src_lang_code in MBART50_LANG_CODES.items():

### 2024-07-29 opus.py

@@ -1,0 +1,62 @@
+import os
+
+import pandas as pd
+import torch
+from tqdm import tqdm
+from transformers import MarianMTModel, MarianTokenizer
+
+MODEL_DICT = {
+    "chinese": "Helsinki-NLP/opus-mt-zh-en",
+    "dutch": "Helsinki-NLP/opus-mt-nl-en",
+    "finnish": "Helsinki-NLP/opus-mt-fi-en",
+    "french": "Helsinki-NLP/opus-mt-fr-en",
+    "german": "Helsinki-NLP/opus-mt-de-en",
+    "hebrew": "tiedeman/opus-mt-he-en",
+    "italian": "Helsinki-NLP/opus-mt-it-en",
+    "japanese": "Helsinki-NLP/opus-mt-ja-en",
+    "polish": "Helsinki-NLP/opus-mt-pl-en",
+    # "portuguese": "Helsinki-NLP/opus-mt-pt-en", # not found
+    "russian": "Helsinki-NLP/opus-mt-ru-en",
+    "spanish": "Helsinki-NLP/opus-mt-es-en",
+    "turkish": "Helsinki-NLP/opus-mt-tr-en",
+    "ukrainian": "Helsinki-NLP/opus-mt-uk-en",
+}
+
+tqdm.pandas()
+
+df = pd.read_csv("dataset/sentences_v1.csv", index_col=0)
+
+
+def translate_batch(tokeniser, batch_input):
+    inputs = tokeniser(
+        batch_input,
+        return_tensors="pt",
+        padding=True,
+        truncation=True,
+    )
+    with torch.no_grad():
+        translated_tokens = model.generate(**inputs)
+    translations = tokeniser.batch_decode(translated_tokens, skip_special_tokens=True)
+
+    return translations
+
+
+batch_size = 8
+translated_df = df[["index", "english"]].copy()
+
+
+for src_lang, model_name in MODEL_DICT.items():
+    model = MarianMTModel.from_pretrained(model_name)
+    tokeniser = MarianTokenizer.from_pretrained(model_name)
+
+    translated_texts = []
+
+    for i in tqdm(range(0, len(df), batch_size)):
+        batch_input = df[src_lang][i : i + batch_size].tolist()
+        translations = translate_batch(tokeniser, batch_input)
+        translated_texts.extend(translations)
+
+    translated_df[f"translated_from_{src_lang}"] = translated_texts
+
+    print(translated_df)
+    translated_df.to_csv(f"result/opus.csv")

### 2024-07-30 paper/paper.tex
updated

@@
 Statistical machine translation \cite{lopez-2008-smt} dominated the domain between late 1990s through the early 2010s, before largely being surpassed by neural machine translation (NMT) \cite{cho-2014-properties,sutskever-2014-seq2seq}.
-Finally, since the introduction of Transformers in 2017 \cite{vaswani-2017-attention}, Natural Language Processing (NLP) and machine translation in particular reached a giant milestone.
+
+Since the introduction of Transformers in 2017 \cite{vaswani-2017-attention}, Natural Language Processing (NLP) and machine translation in particular reached a giant milestone.
 The following years saw the birth of Large Language Models (LLMs) such as BERT \cite{devlin-2019-bert}, GPT \cite{openai-2024-gpt4}, and T5 \cite{raffel-2023-t5}, which revolutionised both MT and the whole field of NLP.
-Soon after, several pre-trained models that are specifically designed for machine translation emerged, namely mBART \cite{liu-2020-mbart}, mT5 \cite{xue-2021-mt5}, and NLLB \cite{nllb200}.
+Then in early 2020s, several pre-trained models (PTM) that are specifically designed for machine translation emerged, namely mBART \cite{liu-2020-mbart}, mT5 \cite{xue-2021-mt5}, NLLB \cite{nllb200-2020}, M2M \cite{fan-2020-m2m100}, and PolyLM \cite{wei-2023-polylm}.
+Most of these models are multilingually trained, able to translate between any of the supported pair of languages.
+This allows the models to generalise over shared lexical and linguistic among languages, and have been shown to increase performance compared to one-to-one translation models \cite{liu-2020-mbart}.
 
-Since LLMs require vast amounts of training data, the availability of parallel corpora is crucial for building effective models.
+Pre-trained models are often not evaluated under the same datasets or benchmarks \cite{liu-2020-mbart,nllb200-2020,fan-2020-m2m100,wei-2023-polylm}.
+Therefore, the performance of these models across different languages seem to be yet unclear.
+
+Through simple experimentation, this paper aims to evaluate the performance of existing pre-trained models (PTMs) across various languages.
+Without any training, the study involves running inference for translations from different languages into English.
+A dataset curated from the Tatoeba repository \cite{tatoeba}, containing 1,241 sentence pairs across 14 source languages, is utilised.
+The evaluation includes a one-to-one PTM, OPUS-MT \cite{tiedemann-2020-opus-mt}, and multilingual PTMs, such as mBART-50 \cite{liu-2020-mbart}, NLLB-200 \cite{nllb200-2020}, and M2M-100 \cite{fan-2020-m2m100}.
+The performance of these models is compared using the BLEU score \cite{papieni-2002-bleu}.
+
+\section{Literature Review}
+
+\subsection{Parallel Corpora}
+
+Since LLMs and pre-trained models require vast amounts of training data, the availability of parallel corpora is crucial for building effective models.
 The lack of extensive parallel corpora, especially for low-resource languages, leads to suboptimal performance in NMT techniques compared to their high-resource counterparts \cite{ranathunga-2023-nmt-low-res}.
-
 OPUS \cite{opus} is a comprehensive collection of open-source parallel corpora used extensively in the field of machine translation (MT).
@@
 
-pre-trained models...
+\subsection{Pre-Trained Models for Machine Translation}
 
-Through simple experimentation and analysis, this paper aims to evaluate existing pre-trained models across different languages
-and contribute valuable insights into the current state of machine translation technology.
+The encoder-decoder approach \cite{cho-2014-properties} remains as the foundation architecture for many sequence-to-sequence models in machine translation.
 
-\section{Literature Review}
-
-\subsection{Neural Machine Translation (NMT)}
-
-The encoder-decoder approach \cite{cho-2014-properties} remains as the foundation architecture for many sequence-to-sequence models in natural language processing, especially machine translation.
-
-\subsection{One-to-one Models}
+In terms of pre-trained models (PTMs) for machine translation, it can be divided into two categories: one-to-one models and many-to-many models (multilingual).
 
@@
 This setup is characterised by having a dedicated model for each unique language pair.
+An example of this setup is OPUS-MT by Helsinki-NLP \cite{tiedemann-2020-opus-mt}, which provides over 1,000 pre-trained models for translation between numerous language pairs.
 
-The OPUS-MT by Helsinki-NLP \cite{tiedemann-2020-opus-mt} is a good example of one-to-one translation project, they provide over 1,000 pre-trained translation models for many language pairs, allowing users to perform high-quality translations without needing to train models from scratch.
+With the advent of large language models and pre-trained language models, multilingual machine translation has gained prominence.
+This approach enables many-to-many translation, where a single model can translate between multiple source and target languages \cite{aharoni-2019-massively-multilingual}.
+mBART \cite{liu-2020-mbart} is a sequence-to-sequence denoising auto-encoder model specifically designed for multilingual tasks.
+The mBART-50 variant supports many-to-many translations for over 50 languages.
+M2M-100 \cite{fan-2020-m2m100} is designed to perform direct translation between 100 languages without relying on English as an intermediate language.
+NLLB-200: NLLB-200 \cite{nllb200-2020} is built to handle translation tasks across a broad spectrum of languages, including many that are low-resource or underrepresented in existing datasets.
+It supports translations for 200 languages, encompassing numerous underrepresented languages.
+mBART, M2M-100, and NLLB-200 were all developed by Meta AI (formerly Facebook AI), showcasing the organisation's significant impact in the machine translation field through pre-trained models (PTMs).
+These models represent a substantial advancement in multilingual translation capabilities.
 
-OpenNMT \cite{klein-etal-2017-opennmt} is another framework that allows for one-to-one translation, although their pre-trained models only support English and German languages.
-
-\subsection{Many-to-many Models}
-
-With the rise of LLMs and pre-trained language models, multilingual machine translation models have received popularity, particularly many-to-many translation.
-
-Multilingual neural machine translation (NMT) allows for the training of a single model capable of translating between multiple source and target languages \cite{aharoni-2019-massively-multilingual}.
-
-Many-to-many translation refers to the capability of a model to translate text between multiple language pairs bidirectionally.
-This means the model can handle translations from any of the supported source languages to any of the supported target languages, not limited to a single source or target language.
-
-mBart \cite{liu-2020-mbart} (50 languages), NLLB-200 \cite{nllb200} (200 language), PolyLM \cite{wei-2023-polylm} (32 languages), and mt5 \cite{xue-2021-mt5} (101 languages) are some of the most popular many-to-many models.
+\begin{comment}
+PolyLM \cite{wei-2023-polylm} (32 languages), and
+mT5 \cite{xue-2021-mt5} (101 languages).
 
@@
 
+\end{comment}
+
 \section{Experiments}
@@
 Tatoeba is a vast, continuously expanding database consisting sentences and their translations, built through the contributions of thousands of volunteers, offering a tool that allows users to see examples of how words are used in sentences \cite{tatoeba}.
-They currently have 12,132,349 sentences and 423 supported languages, with around one to two thousand new sentences added daily, on average
-
-Tatoeba is chosen due to its sentence having more daily life sentences...?
-
-To build the dataset,
-
-Sentences in English are downloaded, 1,898,494 sentences (it is unclear why it is less than the number stated in the Tatoeba website).
-Then for each languages, download sentence pairs compared to English.
-Merge every language sentences into one big dataframe, only keep where sentences exist for every language
+They currently have 12,132,349 sentences and 423 supported languages, with around one to two thousand new sentences added daily, on average.
+The English sentence dataset contains 1,905,089 sentences, the largest one in their repository, with Russian in the second place with 1,066,633 sentences.
+Some languages supported on the website is shown in Figure \ref{fig:tatoeba_languages} and Figure \ref{fig:tatoeba_top_bottom_languages}, sorted from the biggest corpus.
 
@@
     \includegraphics[width=0.9\linewidth]{images/tatoeba_languages.png}
-    \caption{Tatoeba languages repository with 10,000+ sentences and 100,000+ sentences \cite{tatoeba}}
+    \caption{Tatoeba's languages repository with 10,000+ sentences and 100,000+ sentences \cite{tatoeba}}
     \label{fig:tatoeba_languages}
 \end{figure}
-
-Tatoeba English sentence dataset contains 1,905,089 sentences, the largest one in their repository, with Russian in the second place with 1,066,633 sentences.
-Some of the languages supported in the website is shown in Figure \ref{fig:tatoeba_languages} and Figure \ref{fig:tatoeba_top_bottom_languages}, sorted from the biggest corpus.
-Low-resource languages such as Rendille, Southern Haida, and Cuyonon can be seen at the bottom of the list, having only a single sentence example.
-Ancestor languages such as Old Saxon and Old Turkish can also be seen in the list, subsequently with low number of examples.
 
@@
 \end{figure}
-
-Sentences typically consist of everyday phrases such as 'I have to go to sleep', 'That is intriguing', and 'Where do you live?' They may also include single-word exclamations like 'Speak!' or 'Look!' Additionally, multiple sentences such as 'You may write in any language you want.
-On Tatoeba, all languages are considered equal', and 'Guns don't kill people.
-People kill people' can be found inside the corpus.
-A few of them also include human names, 'Compare your answer with Tom's', 'Muiriel is 20 now'.
-All of the sentences are straightforward and literal, without the use of linguistic devices such as metaphors or sarcasm.
-Therefore, machine translation process should be straightforward on this level.
-
-The languages chosen for evaluation in this study represent a wide overview, striking a balance between resources and diversity.
-Each language will then be translated into English (Many-to-English) and evaluated.
-The evaluation criteria encompass standard metrics such as BLEU (Bilingual Evaluation Understudy) \cite{papieni-2002-bleu}, METEOR (Metric for Evaluation of Translation with Explicit ORdering) \cite{lavie-2007-meteor} to capture nuances in translation quality.
-These metrics not only quantify the fidelity of translations but also offer insights into the models' adaptability and robustness across different linguistic pairs.
 
@@
         10           & Polish            \\
-        11           & Portuguese        \\
-        12           & Russian           \\
-        13           & Spanish           \\
-        14           & Turkish           \\
-        15           & Ukrainian         \\
+        11           & Russian           \\
+        12           & Spanish           \\
+        13           & Turkish           \\
+        14           & Ukrainian         \\
         \hline
@@
     \caption{List of chosen languages for evaluation}
+    \label{table:eval_languages}
 \end{table}
 
-Sentences dataset from Tatoeba is used \footnote{\url{https://tatoeba.org/}}.
-Languages that has more than fifty thousand sentences are selected.
-Accordingly, languages that are available for mbart is also selected based on the list here \footnote{\url{https://dl-translate.readthedocs.io/en/latest/available_languages/}}
+Table \ref{table:eval_languages} show the 14 languages selected for this project.
+Languages are chosen based on its resources' availability in Tatoeba, as well as considering supported languages in most PTMs models.
+To build the dataset, sentences in English are first downloaded, containing 1,898,494 sentences (it is unclear why it is less than the number stated in the Tatoeba website).
+Then for each language, sentence pairs between English and source languages are downloaded individually and compiled.
+The result is a single Dataframe containing 1241 sentences in all 14 languages, this will be treated as a test set to evaluate the models performance on each language.
 
-For translation, all languages are translated into English as the target language. Then compare the true English sentence and the predicted one, calculated BLEU.
-
-GPT-4 \cite{openai-2024-gpt4} try GPT and see
+Sentences typically consist of everyday phrases such as 'I have to go to sleep', 'That is intriguing', and 'Where do you live?' They may also include single-word exclamations like 'Speak!' or 'Look!' Additionally, multiple sentences such as 'You may write in any language you want.
+On Tatoeba, all languages are considered equal', and 'Guns don't kill people.
+People kill people' can be found inside the corpus.
+A few of them also include human names, 'Compare your answer with Tom's', 'Muiriel is 20 now'.
+All of the sentences are straightforward and literal, without the use of linguistic devices such as metaphors or sarcasm.
+Therefore, machine translation process should be straightforward on this level.
 

### 2024-07-30 mbart.py

@@ -49,7 +49,7 @@ def translate_batch(tokeniser, batch_input):
     return translations
 
 
-batch_size = 4
+batch_size = 8
 translated_df = df[["index", "english"]].copy()
 
 for src_lang, src_lang_code in MBART50_LANG_CODES.items():

### 2024-07-30 nllb.py

@@ -7,9 +7,11 @@ from transformers import AutoModelForSeq2SeqLM, AutoTokenizer
 
 os.environ["PYTORCH_ENABLE_MPS_FALLBACK"] = "1"
 
-MODEL_NAME = "facebook/nllb-200-3.3B"
+# MODEL_NAME = "facebook/nllb-200-3.3B"
+MODEL_NAME = "facebook/nllb-200-distilled-600M"
 
 MBART50_LANG_CODES = {
+    "chinese": "zho_Hans",
     "dutch": "nld_Latn",
     "finnish": "fin_Latn",
     "french": "fra_Latn",
@@ -17,9 +19,8 @@ MBART50_LANG_CODES = {
     "hebrew": "he_IL",
     "italian": "ita_Latn",
     "japanese": "jpn_Jpan",
-    "chinese": "zho_Hans",
     "polish": "pol_Latn",
-    "portuguese": "por_Latn",
+    # "portuguese": "por_Latn",
     "russian": "rus_Cyrl",
     "spanish": "spa_Latn",
     "turkish": "tur_Latn",
@@ -41,15 +42,11 @@ def translate_batch(batch):
             **inputs
         )
     translations = tokeniser.batch_decode(translated_tokens, skip_special_tokens=True)
-    print(batch)
-    print(translations)
-    asd
 
     return translations
 
 
-batch_size = 1
-# batch_size = 8
+batch_size = 8
 translated_df = df[["index", "english"]].copy()
 
 for src_lang, src_lang_code in MBART50_LANG_CODES.items():
@@ -70,26 +67,3 @@ for src_lang, src_lang_code in MBART50_LANG_CODES.items():
 
     print(translated_df)
     translated_df.to_csv(f"result/{MODEL_NAME.split("/")[1]}.csv")
-
-# [
-#     "Ik moet gaan slapen.",
-#     "Muiriel is nu 20 jaar oud.",
-#     'Het wachtwoord is "Muiriel".',
-#     "Ik ben zo terug.",
-#     "Ik heb er geen woorden voor. | Woorden schieten me tekort.",
-#     "Hier komt nooit een eind aan. | Dit zal nooit eindigen.",
-#     "Ik weet gewoon niet wat ik moet zeggen... | Ik weet eenvoudig niet wat te zeggen...",
-#     "Ik was in de bergen.",
-# ]
-# [
-#     "Lo egin beharko nuke.",
-#     "Muiriel nun estas dudekjara.",
-#     'La pasvorto estas "Muiriel".',
[…]

### 2024-08-03 paper/paper.tex
nllb done

@@
 
-Pre-trained models are often not evaluated under the same datasets or benchmarks \cite{liu-2020-mbart,nllb200-2020,fan-2020-m2m100,wei-2023-polylm}.
-Therefore, the performance of these models across different languages seem to be yet unclear.
+many-to-one? ... text from multiple source languages into a single target language (so this is many-to-one)
+
+(Using many-to-many models for many-to-one translation is a practical approach, as these models are designed to handle multiple languages and can efficiently translate between any pair of supported languages.
+Many-to-many models leverage shared multilingual representations, making them well-suited for translating text from various source languages into a single target language.)
+
+Despite these advancements, pre-trained models are often evaluated using different datasets and benchmarks \cite{liu-2020-mbart,nllb200-2020,fan-2020-m2m100,wei-2023-polylm}, making it difficult to gauge their relative effectiveness across various languages.
+
+Only works by META AI are benchmarked thoroughly with popular datasets across various sources.
+
+There does not seem to be any works done on comparing or benchmarking pre-trained models in machine translation.
 
@@
 
+(Multilinguality has been explored in the supervised NMT literature, where it has been shown to enable information sharing among related languages.) \cite{garcia-2020-multilingual}
+
+(NMT systems have lower translation quality on very long sentences) \cite{koehn-2017-challenges}
+(they often require large amounts of parallel (source, target) sentence pairs to train) \cite{koehn-2017-challenges}
+
+Despite fine-tuning pre-trained multilingual models shows improved performance... \cite{cooper-stickland-2021-recipes}
+
+% Relevance and Impact of Machine Translation:
+% Discuss the broader implications of accurate machine translation beyond just convenience, such as its impact on global communication, business, and cross-cultural exchange.
+
+% Challenges and Limitations:
+% Briefly mention some common challenges in machine translation, such as handling idiomatic expressions, preserving context, and managing low-resource languages.
+
+% Novelty and Contributions:
+% Emphasize what your paper contributes to the existing body of research.
+For instance, is your evaluation filling a gap in the current literature?
+Are there specific insights or implications for future research or practical applications?
+
+% Paper Organization:
+% Briefly outline the structure of the paper to guide the reader on what to expect in subsequent sections.
+
 \section{Literature Review}
+
+\subsection{Evaluation Dataset}
+
+IWSLT \cite{agarwal-iwstl-2023}
+
+TED \cite{ye-2018-word-embeddings-ted}
+
+(Flores-101 evaluation benchmark, consisting of 3001 sentences extracted from English Wikipedia and covering a variety of different topics and domains.
+These sentences have been translated in 101 languages by professional translators through a carefully controlled process.
+The resulting dataset enables better assessment of model quality on the long tail of low-resource languages, including the evaluation of many-to-many multilingual translation systems, as all translations are fully aligned) \cite{goyal-2022-flores}
 
@@
 
-\begin{comment}
-PolyLM \cite{wei-2023-polylm} (32 languages), and
-mT5 \cite{xue-2021-mt5} (101 languages).
+mRASP2 \cite{pan-2021-mrasp2}
 
-\subsection{Sequence-to-Sequence Models}
+\subsection{Evaluation Metrics}
 
-GPT \cite{openai-2024-gpt4}
+% Introduce the concept of evaluation metrics for MT, explaining why BLEU is chosen for this study and if there are other metrics that could complement it (like METEOR or ROUGE).
 
-mBERT \cite{devlin-2019-bert} is the multilingual version of BERT, introduced in the same paper, trained on 104 different languages.
+% \begin{comment}
+% PolyLM \cite{wei-2023-polylm} (32 languages), and
+% mT5 \cite{xue-2021-mt5} (101 languages).
+
+% \subsection{Sequence-to-Sequence Models}
+
+% GPT \cite{openai-2024-gpt4}
+
+% mBERT \cite{devlin-2019-bert} is the multilingual version of BERT, introduced in the same paper, trained on 104 different languages.
 However, the model been shown to suffer on low-resource languages \cite{wu-2020-mbert-are-all}.
 
-T5 \cite{raffel-2023-t5} is a Text-to-Text Transfer Transformer model trained on English.
+% T5 \cite{raffel-2023-t5} is a Text-to-Text Transfer Transformer model trained on English.
 
-MistralAI \cite{jiang-2023-mistral-7b}
+% MistralAI \cite{jiang-2023-mistral-7b}
 
-\end{comment}
+% \end{comment}
 

### 2024-08-03 bleu.py

@@ -1,0 +1,45 @@
+import nltk
+from nltk.translate.bleu_score import corpus_bleu, sentence_bleu
+
+# Ensure you've downloaded the necessary NLTK data
+nltk.download("punkt")
+
+
+def calculate_bleu(reference_sentences, candidate_sentences):
+    """
+    Calculate BLEU score for a list of reference and candidate sentences.
+
+    :param reference_sentences: List of reference sentences (list of lists of words)
+    :param candidate_sentences: List of candidate sentences (list of words)
+    :return: BLEU score
+    """
+    # Tokenize sentences
+    reference_tokenized = [nltk.word_tokenize(sent) for sent in reference_sentences]
+    candidate_tokenized = [nltk.word_tokenize(sent) for sent in candidate_sentences]
+
+    # Calculate BLEU scores
+    individual_bleu_scores = [
+        sentence_bleu([ref], cand)
+        for ref, cand in zip(reference_tokenized, candidate_tokenized)
+    ]
+    corpus_bleu_score = corpus_bleu(
+        [[ref] for ref in reference_tokenized], candidate_tokenized
+    )
+
+    return individual_bleu_scores, corpus_bleu_score
+
+
+# Example usage
+reference_sentences = [
+    "The quick brown fox jumps over the lazy dog.",
+    "A fast brown fox leaps over a sleepy dog.",
+]
+candidate_sentences = [
+    "The quick brown fox jumps over the lazy dog.",
+    "A quick brown fox jumps over the lazy dog.",
+]
+
+individual_scores, overall_score = calculate_bleu(
+    reference_sentences, candidate_sentences
+)
+prin

### 2024-08-03 m2m.py

@@ -1,0 +1,45 @@
+from transformers import M2M100ForConditionalGeneration, M2M100Tokenizer
+
+# Load pre-trained model and tokenizer
+model_name = "facebook/m2m100_418M"
+tokeniser = M2M100Tokenizer.from_pretrained(model_name)
+model = M2M100ForConditionalGeneration.from_pretrained(model_name)
+
+LANG_CODES = {
+    "chinese": "zh",
+    "dutch": "nl",
+    "finnish": "fi",
+    "french": "fr",
+    "german": "de",
+    "hebrew": "he",
+    "italian": "it",
+    "japanese": "ja",
+    "polish": "pl",
+    # "portugues": "pt",
+    "russian": "ru",
+    "spanish": "es",
+    "turkish": "tr",
+    "ukrainian": "uk",
+}
+
+
+def translate_to_english(text, src_lang="en", tgt_lang="en"):
+    # Encode the source text
+    tokeniser.src_lang = src_lang
+    encoded_input = tokeniser(text, return_tensors="pt")
+
+    # Generate translation
+    generated_tokens = model.generate(
+        **encoded_input, forced_bos_token_id=tokeniser.get_lang_id(tgt_lang)
+    )
+
+    # Decode the generated tokens
+    translated_text = tokeniser.batch_decode(generated_tokens, skip_special_tokens=True)
+
+    return translated_text[0]
+
+
+# Example usage
+source_text = "Bonjour tout le monde"  # French for "Hello everyone"
+translated_text = translate_to_english(source_text, src_lang="fr", tgt_lang="en")
+print(translated_text)

### 2024-08-03 mbart.py

@@ -9,7 +9,7 @@ os.environ["PYTORCH_ENABLE_MPS_FALLBACK"] = "1"
 
 MODEL_NAME = "facebook/mbart-large-50-many-to-many-mmt"
 
-MBART50_LANG_CODES = {
+LANG_CODES = {
     "chinese": "zh_CN",
     "dutch": "nl_XX",
     "finnish": "fi_FI",
@@ -52,7 +52,7 @@ def translate_batch(tokeniser, batch_input):
 batch_size = 8
 translated_df = df[["index", "english"]].copy()
 
-for src_lang, src_lang_code in MBART50_LANG_CODES.items():
+for src_lang, src_lang_code in LANG_CODES.items():
     print(f"Translating {src_lang}...")
     tokeniser = MBart50Tokenizer.from_pretrained(
         MODEL_NAME,

### 2024-08-03 mrasp2.py

@@ -7,8 +7,7 @@ from transformers import AutoModelForSeq2SeqLM, AutoTokenizer
 
 os.environ["PYTORCH_ENABLE_MPS_FALLBACK"] = "1"
 
-# MODEL_NAME = "facebook/nllb-200-3.3B"
-MODEL_NAME = "facebook/nllb-200-distilled-600M"
+MODEL_NAME = "chinmaydan/mRASP2"
 
 LANG_CODES = {
     "chinese": "zho_Hans",
@@ -29,20 +28,17 @@ LANG_CODES = {
 
 tqdm.pandas()
 
-tokeniser = AutoTokenizer.from_pretrained(MODEL_NAME)
+tokenizer = AutoTokenizer.from_pretrained(MODEL_NAME)
 model = AutoModelForSeq2SeqLM.from_pretrained(MODEL_NAME)
 
 df = pd.read_csv("dataset/sentences_v1.csv", index_col=0)
 
 
 def translate_batch(batch):
-    inputs = tokeniser(batch, return_tensors="pt", padding=True, truncation=True)
+    inputs = tokenizer(batch, return_tensors="pt", padding=True, truncation=True)
     with torch.no_grad():
-        translated_tokens = model.generate(
-            **inputs
-        )
-    translations = tokeniser.batch_decode(translated_tokens, skip_special_tokens=True)
-
+        translated_tokens = model.generate(**inputs)
+    translations = tokenizer.batch_decode(translated_tokens, skip_special_tokens=True)
     return translations
 
 
@@ -51,7 +47,7 @@ translated_df = df[["index", "english"]].copy()
 
 for src_lang, src_lang_code in LANG_CODES.items():
     print(f"Translating {src_lang}...")
-    tokeniser = AutoTokenizer.from_pretrained(
+    tokenizer = AutoTokenizer.from_pretrained(
         MODEL_NAME,
         src_lang=src_lang_code,
         tgt_lang="eng_Latn",
@@ -66,4 +62,4 @@ for src_lang, src_lang_code in LANG_CODES.items():
     translated_df[f"translated_from_{src_lang}"] = translated_texts
 
     print(translated_df)
-    translated_df.to_csv(f"result/{MODEL_NAME.split("/")[1]}.csv")
+    translated_df.to_csv(f"result/{MODEL_NAME.split('/')[1]}.csv")

### 2024-08-04 paper/paper.tex
updated

--- 2024-07-30 paper/paper.tex
@@
 \usepackage{hyperref}
+\usepackage{amsmath}
 
@@
 
-\title{Evaluation of Pre-Trained Models for Translation in Different Languages}
+\title{Evaluation of Pre-Trained Models for Many-to-English Translation}
 \author{Stefan Liemawan Adji}
@@
 Then in early 2020s, several pre-trained models (PTM) that are specifically designed for machine translation emerged, namely mBART \cite{liu-2020-mbart}, mT5 \cite{xue-2021-mt5}, NLLB \cite{nllb200-2020}, M2M \cite{fan-2020-m2m100}, and PolyLM \cite{wei-2023-polylm}.
-Most of these models are multilingually trained, able to translate between any of the supported pair of languages.
+Most of these models are multilingually trained, allowing for many-to-many translation: able to translate between any of the supported pair of languages.
 This allows the models to generalise over shared lexical and linguistic among languages, and have been shown to increase performance compared to one-to-one translation models \cite{liu-2020-mbart}.
 
-Pre-trained models are often not evaluated under the same datasets or benchmarks \cite{liu-2020-mbart,nllb200-2020,fan-2020-m2m100,wei-2023-polylm}.
-Therefore, the performance of these models across different languages seem to be yet unclear.
+Intento published 'The State of Machine Translation 2024' \cite{intento-2020} providing an in-depth evaluation of popular MT engines and LLMs.
+However, the biggest drawback in this report is that the selected LLMs are general LLMs such as GPT, LLaMa, Mistral, instead of MT-specific LLMs.
 
-Through simple experimentation, this paper aims to evaluate the performance of existing pre-trained models (PTMs) across various languages.
-Without any training, the study involves running inference for translations from different languages into English.
-A dataset curated from the Tatoeba repository \cite{tatoeba}, containing 1,241 sentence pairs across 14 source languages, is utilised.
-The evaluation includes a one-to-one PTM, OPUS-MT \cite{tiedemann-2020-opus-mt}, and multilingual PTMs, such as mBART-50 \cite{liu-2020-mbart}, NLLB-200 \cite{nllb200-2020}, and M2M-100 \cite{fan-2020-m2m100}.
-The performance of these models is compared using the BLEU score \cite{papieni-2002-bleu}.
+Despite these advancements, pre-trained models are often evaluated using different set of benchmarks \cite{liu-2020-mbart,nllb200-2020,fan-2020-m2m100,wei-2023-polylm}, making it difficult to gauge their relative effectiveness across various languages.
+Nevertheless, there does not seem to be much work on comparing or benchmarking different pre-trained models in machine translation.
+
+Through simple experimentations, this paper aims to evaluate the performance of existing pre-trained models (PTMs) on many-to-English translation across 14 source languages.
+Although fine-tuning multilingual PTMs has been proven to increase model performance \cite{cooper-stickland-2021-recipes}, no pre-training or fine-tuning is performed in this study for simplicity reasons.
+A dataset is curated from the Tatoeba repository \cite{tatoeba}, containing 1,241 parallel sentence pairs across source and target languages.
+The models includes a one-to-one PTMs: OPUS-MT \cite{tiedemann-2020-opus-mt}, and multilingual PTMs: such as mBART-50 \cite{liu-2020-mbart}, NLLB-200 \cite{nllb200-2020}, and M2M-100 \cite{fan-2020-m2m100}.
+The performance of these models is evaluated using the BLEU score \cite{papieni-2002-bleu}.
+
+% (Multilinguality has been explored in the supervised NMT literature, where it has been shown to enable information sharing among related languages.) \cite{garcia-2020-multilingual}
+% (NMT systems have lower translation quality on very long sentences) \cite{koehn-2017-challenges}
 
@@
 
-Since LLMs and pre-trained models require vast amounts of training data, the availability of parallel corpora is crucial for building effective models.
+Since Neural Machine Translation (NMT) systems require vast amounts of training data, the availability of parallel corpora is crucial for building effective models \cite{koehn-2017-challenges}.
 The lack of extensive parallel corpora, especially for low-resource languages, leads to suboptimal performance in NMT techniques compared to their high-resource counterparts \cite{ranathunga-2023-nmt-low-res}.
@@
 As of July 2024, it contains 12,186,207 sentences over 423 supported languages, growing daily through volunteer contributions.
+
+The Tatoeba Challenge \cite{tiedemann-2020-tatoeba-challenge}, IWSLT \cite{agarwal-iwstl-2023}, TED \cite{ye-2018-word-embeddings-ted}, Flores \cite{goyal-2022-flores}, and WMT \cite{barrault-2020-wmt} are among the biggest and most popular parallel corpora datasets, allowing for many-to-many translations.
+
+Despite being used diligently by Meta AI papers for benchmarking \cite{fan-2020-m2m100,nllb200-2020}, these datasets are not seen used
 
@@
 
-\begin{comment}
-PolyLM \cite{wei-2023-polylm} (32 languages), and
-mT5 \cite{xue-2021-mt5} (101 languages).
+mRASP2 \cite{pan-2021-mrasp2}
 
-\subsection{Sequence-to-Sequence Models}
+\subsection{Evaluation Metrics}
 
-GPT \cite{openai-2024-gpt4}
+Bilingual Evaluation Understudy (BLEU) \cite{papieni-2002-bleu} is the most commonly used metrics for machine translation (MT).
+It assesses how well a candidate translation matches the reference translation using precision metrics for n-grams and incorporates a brevity penalty to prevent overly short translations from achieving high scores.
 
-mBERT \cite{devlin-2019-bert} is the multilingual version of BERT, introduced in the same paper, trained on 104 different languages.
+The n-gram precision, as presented in the original BLEU paper \cite{papieni-2002-bleu}, is calculated as:
+
+\begin{equation}
+    p_n = \frac{\sum_{C \in \{Candidates\}} \sum_{\text{n-gram} \in C} \text{Count}_{\text{clip}}(\text{n-gram})}{\sum_{C \in \{Candidates\}} \sum_{\text{n-gram} \in C} \text{Count}(\text{n-gram})}
+\end{equation}
+
+Where:
+\begin{itemize}
+    \item \( p_n \) is the precision for n-grams.
+    \item \( \sum_{C \in \{Candidates\}} \) denotes the summation over all candidate translations.
+    \item \( \sum_{\text{n-gram} \in C} \) denotes the summation over all n-grams in a candidate translation \( C \).
+    \item \( \text{Count}_{\text{clip}}(\text{n-gram}) \) is the clipped count of the n-gram, which is the count of the n-gram in the candidate translation limited by the maximum count of that n-gram in any reference translation.
+    \item \( \text{Count}(\text{n-gram}) \) is the count of the n-gram in the candidate translation.
+\end{itemize}
+
+Thus, the BLEU score is calculated as:
+
+\begin{equation}
+    \text{BLEU} = BP \cdot \exp \left( \sum_{n=1}^{N} w_n \log p_n \right)
+\end{equation}
+
+Where:
+\begin{itemize}
+    \item \( BP \) is the brevity penalty.
+    \item \( p_n \) is the precision for n-grams.
+    \item \( w_n \) is the weight for each n-gram (often uniformly distributed, so \( w_n = \frac{1}{N} \)).
+\end{itemize}
+
+The brevity penalty (BP) is calculated as:
+
+\begin{equation}
+    BP = \begin{cases}
+        1                     & \text{if } c > r    \\
+        e^{(1 - \frac{r}{c})} & \text{if } c \leq r
+    \end{cases}
+\end{equation}
+
+Where:
+\begin{itemize}
+    \item \( c \) is the length of the candidate translation.
+    \item \( r \) is the effective reference length.
+\end{itemize}
+
+While other metrics exist for machine translation such as METEOR \cite{lavie-2007-meteor} and COMET \cite{rei-2020-comet}, it is far more practical to implement BLEU due to its popular usage in other works.
+
+% Introduce the concept of evaluation metrics for MT, explaining why BLEU is chosen for this study and if there are other metrics that could complement it (like METEOR or ROUGE).
+
+% \begin{comment}
+% PolyLM \cite{wei-2023-polylm} (32 languages), and
+% mT5 \cite{xue-2021-mt5} (101 languages).
+
+% \subsection{Sequence-to-Sequence Models}
+
+% GPT \cite{openai-2024-gpt4}
+
+% mBERT \cite{devlin-2019-bert} is the multilingual version of BERT, introduced in the same paper, trained on 104 different languages.
 However, the model been shown to suffer on low-resource languages \cite{wu-2020-mbert-are-all}.
 
-T5 \cite{raffel-2023-t5} is a Text-to-Text Transfer Transformer model trained on English.
+% T5 \cite{raffel-2023-t5} is a Text-to-Text Transfer Transformer model trained on English.
 
-MistralAI \cite{jiang-2023-mistral-7b}
+% MistralAI \cite{jiang-2023-mistral-7b}
 
-\end{comment}
+% \end{comment}
 
@@
 
-READ THIS \cite{intento-2020} (exactly what this paper should do, maybe compare results and find insights)
+\begin{figure}[htbp]
+    \centering
+    \includegraphics[width=1\linewidth]{figures/word_count_box.png}
+    \caption{Dataset word count distribution, consisting of mostly short sentences.}
+    \label{fig:word_count_box}
+\end{figure}
+
+Figure \ref{fig:word_count_box} shows a box plot of sentences word count. It can be seen that the majority of sentences are short.
+
+Chinese and Japanese are counted per letter.
+
+While most languages cluster around 4-6 words per sentence, there are notable exceptions like Japanese, which exhibits much less variability.
+This analysis can be useful for understanding language-specific characteristics in sentence structure, which could inform tasks like translation, text processing, or linguistic studies.
 
@@
 
+% READ THIS \cite{intento-2020} (exactly what this paper should do, maybe compare results and find insights)
+
 \printbibliography

### 2024-08-04 analyse/analyse_sentences.py

@@ -1,14 +1,148 @@
+import os
 from collections import Counter
 
 import matplotlib.pyplot as plt
 import nltk
 import pandas as pd
+import seaborn as sns
 import spacy
+from sentence_transformers import SentenceTransformer
+from sklearn.manifold import TSNE
 from textblob import TextBlob
 from wordcloud import WordCloud
 
-sentences_df = pd.read_csv("../dataset/sentences_v1.csv", index_col=0)
+os.environ["TOKENIZERS_PARALLELISM"] = "false"
+RANDOM_STATE = 42
 
-print(sentences_df)
-
+sentences_df = pd.read_csv("../dataset/sentences_v2.csv", index_col=0)
 english = sentences_df["english"].tolist()
+
+LANG = [
+    "english",
+    "chinese",
+    "dutch",
+    "finnish",
+    "french",
+    "german",
+    "hebrew",
+    "italian",
+    "japanese",
+    "polish",
+    "russian",
+    "spanish",
+    "turkish",
+    "ukrainian",
+]
+
+
+def calculate_len(sentence):
+    if "|" in sentence:
+        sentence = sentence.split("|")
+        return min([len(nltk.word_tokenize(x)) for x in sentence])
+    else:
+        return len(nltk.word_tokenize(sentence))
+
+
+def analyse_sentence_length(df):
+    plt.clf()
+    plt.figure(figsize=(10, 6))
+
+    for lang in LANG:
+        if lang in ["chinese", "japanese"]:
+            df[f"{lang}_length"] = df[lang].apply(len)
+        else:
+            df[f"{lang}_length"] = df[lang].apply(calculate_len)
+
+        plt.hist(df[f"{lang}_length"], bins=20, alpha=0.3, label=lang)
+
+    plt.legend(loc="upper right")
+    plt.title("Word Count Distribution")
+    plt.ylabel("Word Count")
+    plt.ylabel("Frequency")
+
+    plt.savefig("./figures/word_count_hist.png")
+
+
+def analyse_sentence_length_2(df):
+    plt.clf()
+    plt.figure(figsize=(15, 6))
+
+    # Prepare data for box plot
+    lengths = []
+    languages = []
+    for lang in LANG:
+        if lang in ["chinese", "japanese"]:
+            df[f"{lang}_length"] = df[lang].apply(len)
+        else:
[…]

### 2024-08-04 handle_tatoeba.py

@@ -54,17 +54,9 @@ language_list = [
         "japanese",
         f"{path}/Sentence pairs in English-Japanese - 2024-07-05.tsv",
     ),
-    # {
-    #     "lithuanian",
-    #     f"{path}/Sentence pairs in English-Lithuanian - 2024-07-29.tsv",
-    # },
     (
         "polish",
         f"{path}/Sentence pairs in English-Polish - 2024-07-05.tsv",
-    ),
-    (
-        "portuguese",
-        f"{path}/Sentence pairs in English-Portuguese - 2024-07-05.tsv",
     ),
     (
         "russian",
@@ -96,9 +88,10 @@ main_df.drop(columns=["language"], inplace=True)
 for language, tsv_path in tqdm(language_list):
     lang_df = read_df(language, tsv_path)
 
-    lang_df_grouped = lang_df.groupby("index")[language].apply(
-        lambda x: " | ".join(x.dropna().astype(str))
-    )
+    # lang_df_grouped = lang_df.groupby("index")[language].apply(
+    #     lambda x: " | ".join(x.dropna().astype(str))
+    # )
+    lang_df_grouped = lang_df.groupby("index")[language].first()
 
     main_df = pd.merge(
         main_df,
@@ -111,4 +104,4 @@ for language, tsv_path in tqdm(language_list):
 main_df.dropna(inplace=True)
 print(main_df)
 
-main_df.to_csv(f"dataset/sentences_v3.csv", index=False)
+main_df.to_csv(f"dataset/sentences_v2.csv", index=True)

### 2024-08-04 inference/m2m.py

@@ -1,0 +1,75 @@
+import time
+
+import pandas as pd
+from tqdm import tqdm
+from transformers import M2M100ForConditionalGeneration, M2M100Tokenizer
+
+MODEL_NAME = "facebook/m2m100_418M"
+tokeniser = M2M100Tokenizer.from_pretrained(MODEL_NAME)
+model = M2M100ForConditionalGeneration.from_pretrained(MODEL_NAME)
+
+df = pd.read_csv("../dataset/sentences_v2.csv", index_col=0)
+tqdm.pandas()
+
+LANG_CODES = {
+    "chinese": "zh",
+    "dutch": "nl",
+    "finnish": "fi",
+    "french": "fr",
+    "german": "de",
+    "hebrew": "he",
+    "italian": "it",
+    "japanese": "ja",
+    "polish": "pl",
+    "russian": "ru",
+    "spanish": "es",
+    "turkish": "tr",
+    "ukrainian": "uk",
+}
+
+
+def translate_batch(tokeniser, batch_input):
+    inputs = tokeniser(
+        batch_input,
+        return_tensors="pt",
+        padding=True,
+        truncation=True,
+    )
+    translated_tokens = model.generate(
+        **inputs, forced_bos_token_id=tokeniser.get_lang_id("en")
+    )
+    translations = tokeniser.batch_decode(translated_tokens, skip_special_tokens=True)
+
+    return translations
+
+
+batch_size = 4
+translated_df = df[["index", "english"]].copy()
+timing_df = pd.DataFrame()
+
+for src_lang, src_lang_code in LANG_CODES.items():
+    print(f"Translating {src_lang}...")
+    start = time.time()
+
+    tokeniser = M2M100Tokenizer.from_pretrained(
+        MODEL_NAME,
+        src_lang=src_lang_code,
+        tgt_lang="en",
+    )
+    translated_texts = []
+
+    for i in tqdm(range(0, len(df), batch_size)):
+        batch_input = df[src_lang][i : i + batch_size].tolist()
+        translations = translate_batch(tokeniser, batch_input)
+        translated_texts.extend(translations)
+
+    translated_df[f"translated_from_{src_lang}"] = translated_texts
+
+    end = time.time()
+    timing_df[f"translated_from_{src_lang}"] = [end - start]
+
+    translated_df.to_csv(f"result/{MODEL_NAME.split("/")[1]}.csv")
+    timing_df.to_csv(f"result/{MODEL_NAME.split("/")[1]}_timings.csv")
+
+
+# use vsc

### 2024-08-04 inference/mbart.py

@@ -1,4 +1,5 @@
 import os
+import time
 
 import pandas as pd
 import torch
@@ -19,18 +20,17 @@ LANG_CODES = {
     "italian": "it_IT",
     "japanese": "ja_XX",
     "polish": "pl_PL",
-    # "portuguese": "pt_XX",
     "russian": "ru_RU",
     "spanish": "es_XX",
     "turkish": "tr_TR",
     "ukrainian": "uk_UA",
 }
 
-tqdm.pandas()
 
 model = MBartForConditionalGeneration.from_pretrained(MODEL_NAME)
 
-df = pd.read_csv("dataset/sentences_v1.csv", index_col=0)
+df = pd.read_csv("../dataset/sentences_v2.csv", index_col=0)
+tqdm.pandas()
 
 
 def translate_batch(tokeniser, batch_input):
@@ -49,11 +49,14 @@ def translate_batch(tokeniser, batch_input):
     return translations
 
 
-batch_size = 8
+batch_size = 4
 translated_df = df[["index", "english"]].copy()
+timing_df = pd.DataFrame()
 
 for src_lang, src_lang_code in LANG_CODES.items():
     print(f"Translating {src_lang}...")
+    start = time.time()
+
     tokeniser = MBart50Tokenizer.from_pretrained(
         MODEL_NAME,
         src_lang=src_lang_code,
@@ -68,26 +71,9 @@ for src_lang, src_lang_code in LANG_CODES.items():
 
     translated_df[f"translated_from_{src_lang}"] = translated_texts
 
-    print(translated_df)
-    translated_df.to_csv(f"result/{MODEL_NAME.split("/")[1]}.csv")
+    end = time.time()
+    timing_df[f"translated_from_{src_lang}"] = [end - start]
 
-# [
-#     "Ik moet gaan slapen.",
-#     "Muiriel is nu 20 jaar oud.",
-#     'Het wachtwoord is "Muiriel".',
-#     "Ik ben zo terug.",
-#     "Ik heb er geen woorden voor. | Woorden schieten me tekort.",
-#     "Hier komt nooit een eind aan. | Dit zal nooit eindigen.",
-#     "Ik weet gewoon niet wat ik moet zeggen... | Ik weet eenvoudig niet wat te zeggen...",
-#     "Ik was in de bergen.",
-# ]
-# [
-#     "I have to go to sleep.",
-#     "Muiriel is now 20 years old.",
-#     'The password is "Muiriel."',
-#     "I'm so back.",
[…]

### 2024-08-04 inference/mrasp2.py

@@ -1,4 +1,5 @@
 import os
+import time
 
 import pandas as pd
 import torch
@@ -7,8 +8,7 @@ from transformers import AutoModelForSeq2SeqLM, AutoTokenizer
 
 os.environ["PYTORCH_ENABLE_MPS_FALLBACK"] = "1"
 
-# MODEL_NAME = "facebook/nllb-200-3.3B"
-MODEL_NAME = "facebook/nllb-200-distilled-600M"
+MODEL_NAME = "chinmaydan/mRASP2"
 
 LANG_CODES = {
     "chinese": "zho_Hans",
@@ -20,7 +20,6 @@ LANG_CODES = {
     "italian": "ita_Latn",
     "japanese": "jpn_Jpan",
     "polish": "pol_Latn",
-    # "portuguese": "por_Latn",
     "russian": "rus_Cyrl",
     "spanish": "spa_Latn",
     "turkish": "tur_Latn",
@@ -29,29 +28,29 @@ LANG_CODES = {
 
 tqdm.pandas()
 
-tokeniser = AutoTokenizer.from_pretrained(MODEL_NAME)
+tokenizer = AutoTokenizer.from_pretrained(MODEL_NAME)
 model = AutoModelForSeq2SeqLM.from_pretrained(MODEL_NAME)
 
-df = pd.read_csv("dataset/sentences_v1.csv", index_col=0)
+df = pd.read_csv("../dataset/sentences_v2.csv", index_col=0)
 
 
 def translate_batch(batch):
-    inputs = tokeniser(batch, return_tensors="pt", padding=True, truncation=True)
+    inputs = tokenizer(batch, return_tensors="pt", padding=True, truncation=True)
     with torch.no_grad():
-        translated_tokens = model.generate(
-            **inputs
-        )
-    translations = tokeniser.batch_decode(translated_tokens, skip_special_tokens=True)
-
+        translated_tokens = model.generate(**inputs)
+    translations = tokenizer.batch_decode(translated_tokens, skip_special_tokens=True)
     return translations
 
 
-batch_size = 8
+batch_size = 4
 translated_df = df[["index", "english"]].copy()
+timing_df = pd.DataFrame()
 
 for src_lang, src_lang_code in LANG_CODES.items():
     print(f"Translating {src_lang}...")
-    tokeniser = AutoTokenizer.from_pretrained(
+    start = time.time()
+
+    tokenizer = AutoTokenizer.from_pretrained(
[…]

### 2024-08-04 inference/nllb.py

@@ -1,4 +1,5 @@
 import os
+import time
 
 import pandas as pd
 import torch
@@ -7,7 +8,8 @@ from transformers import AutoModelForSeq2SeqLM, AutoTokenizer
 
 os.environ["PYTORCH_ENABLE_MPS_FALLBACK"] = "1"
 
-MODEL_NAME = "chinmaydan/mRASP2"
+# MODEL_NAME = "facebook/nllb-200-3.3B"
+MODEL_NAME = "facebook/nllb-200-distilled-600M"
 
 LANG_CODES = {
     "chinese": "zho_Hans",
@@ -19,7 +21,6 @@ LANG_CODES = {
     "italian": "ita_Latn",
     "japanese": "jpn_Jpan",
     "polish": "pol_Latn",
-    # "portuguese": "por_Latn",
     "russian": "rus_Cyrl",
     "spanish": "spa_Latn",
     "turkish": "tur_Latn",
@@ -28,26 +29,32 @@ LANG_CODES = {
 
 tqdm.pandas()
 
-tokenizer = AutoTokenizer.from_pretrained(MODEL_NAME)
+tokeniser = AutoTokenizer.from_pretrained(MODEL_NAME)
 model = AutoModelForSeq2SeqLM.from_pretrained(MODEL_NAME)
 
-df = pd.read_csv("dataset/sentences_v1.csv", index_col=0)
+df = pd.read_csv("../dataset/sentences_v2.csv", index_col=0)
 
 
 def translate_batch(batch):
-    inputs = tokenizer(batch, return_tensors="pt", padding=True, truncation=True)
+    inputs = tokeniser(batch, return_tensors="pt", padding=True, truncation=True)
     with torch.no_grad():
-        translated_tokens = model.generate(**inputs)
-    translations = tokenizer.batch_decode(translated_tokens, skip_special_tokens=True)
+        translated_tokens = model.generate(
+            **inputs
+        )
+    translations = tokeniser.batch_decode(translated_tokens, skip_special_tokens=True)
+
     return translations
 
 
-batch_size = 8
+batch_size = 4
 translated_df = df[["index", "english"]].copy()
+timing_df = pd.DataFrame()
 
 for src_lang, src_lang_code in LANG_CODES.items():
     print(f"Translating {src_lang}...")
-    tokenizer = AutoTokenizer.from_pretrained(
+    start = time.time()
+
+    tokeniser = AutoTokenizer.from_pretrained(
[…]

### 2024-08-04 inference/opus.py

@@ -1,4 +1,4 @@
-import os
+import time
 
 import pandas as pd
 import torch
@@ -15,7 +15,6 @@ MODEL_DICT = {
     "italian": "Helsinki-NLP/opus-mt-it-en",
     "japanese": "Helsinki-NLP/opus-mt-ja-en",
     "polish": "Helsinki-NLP/opus-mt-pl-en",
-    # "portuguese": "Helsinki-NLP/opus-mt-pt-en", # not found
     "russian": "Helsinki-NLP/opus-mt-ru-en",
     "spanish": "Helsinki-NLP/opus-mt-es-en",
     "turkish": "Helsinki-NLP/opus-mt-tr-en",
@@ -24,7 +23,7 @@ MODEL_DICT = {
 
 tqdm.pandas()
 
-df = pd.read_csv("dataset/sentences_v1.csv", index_col=0)
+df = pd.read_csv("../dataset/sentences_v2.csv", index_col=0)
 
 
 def translate_batch(tokeniser, batch_input):
@@ -41,11 +40,14 @@ def translate_batch(tokeniser, batch_input):
     return translations
 
 
-batch_size = 8
+batch_size = 4
 translated_df = df[["index", "english"]].copy()
-
+timing_df = pd.DataFrame()
 
 for src_lang, model_name in MODEL_DICT.items():
+    print(f"Translating {src_lang}...")
+    start = time.time()
+
     model = MarianMTModel.from_pretrained(model_name)
     tokeniser = MarianTokenizer.from_pretrained(model_name)
 
@@ -58,5 +60,8 @@ for src_lang, model_name in MODEL_DICT.items():
 
     translated_df[f"translated_from_{src_lang}"] = translated_texts
 
-    print(translated_df)
+    end = time.time()
+    timing_df[f"translated_from_{src_lang}"] = [end - start]
+
     translated_df.to_csv(f"result/opus.csv")
+    timing_df.to_csv(f"result/opus_timings.csv")

### 2024-08-04 m2m.py

@@ -1,45 +1,0 @@
-from transformers import M2M100ForConditionalGeneration, M2M100Tokenizer
-
-# Load pre-trained model and tokenizer
-model_name = "facebook/m2m100_418M"
-tokeniser = M2M100Tokenizer.from_pretrained(model_name)
-model = M2M100ForConditionalGeneration.from_pretrained(model_name)
-
-LANG_CODES = {
-    "chinese": "zh",
-    "dutch": "nl",
-    "finnish": "fi",
-    "french": "fr",
-    "german": "de",
-    "hebrew": "he",
-    "italian": "it",
-    "japanese": "ja",
-    "polish": "pl",
-    # "portugues": "pt",
-    "russian": "ru",
-    "spanish": "es",
-    "turkish": "tr",
-    "ukrainian": "uk",
-}
-
-
-def translate_to_english(text, src_lang="en", tgt_lang="en"):
-    # Encode the source text
-    tokeniser.src_lang = src_lang
-    encoded_input = tokeniser(text, return_tensors="pt")
-
-    # Generate translation
-    generated_tokens = model.generate(
-        **encoded_input, forced_bos_token_id=tokeniser.get_lang_id(tgt_lang)
-    )
-
-    # Decode the generated tokens
-    translated_text = tokeniser.batch_decode(generated_tokens, skip_special_tokens=True)
-
-    return translated_text[0]
-
-
-# Example usage
-source_text = "Bonjour tout le monde"  # French for "Hello everyone"
-translated_text = translate_to_english(source_text, src_lang="fr", tgt_lang="en")
-print(translated_text)

### 2024-08-05 paper/paper.tex
updated and added result

@@
 \usepackage[margin=1in]{geometry}
+\usepackage[utf8]{inputenc}
+\usepackage[T1]{fontenc}
 \usepackage[backend=bibtex]{biblatex}
@@
 \usepackage{amsmath}
+
+\usepackage{CJKutf8}
 
@@
 Although fine-tuning multilingual PTMs has been proven to increase model performance \cite{cooper-stickland-2021-recipes}, no pre-training or fine-tuning is performed in this study for simplicity reasons.
-A dataset is curated from the Tatoeba repository \cite{tatoeba}, containing 1,241 parallel sentence pairs across source and target languages.
-The models includes a one-to-one PTMs: OPUS-MT \cite{tiedemann-2020-opus-mt}, and multilingual PTMs: such as mBART-50 \cite{liu-2020-mbart}, NLLB-200 \cite{nllb200-2020}, and M2M-100 \cite{fan-2020-m2m100}.
+A dataset is curated from the Tatoeba repository \cite{tatoeba}, containing 1,323 parallel sentence pairs across source and target languages.
+The models includes a one-to-one PTMs: OPUS-MT \cite{tiedemann-2023-democratizing,tiedemann-2020-opus-mt}, and multilingual PTMs: such as mBART-50 \cite{liu-2020-mbart}, NLLB-200 \cite{nllb200-2020}, and M2M-100 \cite{fan-2020-m2m100}.
 The performance of these models is evaluated using the BLEU score \cite{papieni-2002-bleu}.
@@
 \section{Literature Review}
-
-\subsection{Parallel Corpora}
-
-Since Neural Machine Translation (NMT) systems require vast amounts of training data, the availability of parallel corpora is crucial for building effective models \cite{koehn-2017-challenges}.
-The lack of extensive parallel corpora, especially for low-resource languages, leads to suboptimal performance in NMT techniques compared to their high-resource counterparts \cite{ranathunga-2023-nmt-low-res}.
-OPUS \cite{opus} is a comprehensive collection of open-source parallel corpora used extensively in the field of machine translation (MT).
-It includes corpora for 744 languages and contains over 1,210 different datasets, amassing a total of 45,945,946,108 sentence pairs.
-Tatoeba \cite{tatoeba} is another prominent resource in the field of MT and NLP, known for its extensive collection of translated sentences.
-As of July 2024, it contains 12,186,207 sentences over 423 supported languages, growing daily through volunteer contributions.
-
-The Tatoeba Challenge \cite{tiedemann-2020-tatoeba-challenge}, IWSLT \cite{agarwal-iwstl-2023}, TED \cite{ye-2018-word-embeddings-ted}, Flores \cite{goyal-2022-flores}, and WMT \cite{barrault-2020-wmt} are among the biggest and most popular parallel corpora datasets, allowing for many-to-many translations.
-
-Despite being used diligently by Meta AI papers for benchmarking \cite{fan-2020-m2m100,nllb200-2020}, these datasets are not seen used
 
@@
 This setup is characterised by having a dedicated model for each unique language pair.
-An example of this setup is OPUS-MT by Helsinki-NLP \cite{tiedemann-2020-opus-mt}, which provides over 1,000 pre-trained models for translation between numerous language pairs.
+An example of this setup is OPUS-MT by Helsinki-NLP \cite{tiedemann-2023-democratizing,tiedemann-2020-opus-mt}, which provides over 1,000 pre-trained models for translation between numerous language pairs.
 
@@
 
-mRASP2 \cite{pan-2021-mrasp2}
+\subsection{Parallel Corpora}
+
+Corpora are large and structured sets of texts used for linguistic research and analysis.
+Thus, parallel corpora are defined as sets of texts in a given source language along with their translations in another target language \cite{lefer-2020-parallel-corpora}.
+It can be bilingual or multilingual, and are crucial for training and evaluating machine translation.
+Languages without or with limited parallel corpora are referred as \textbf{low-resources language}.
+
+Since the early 2000s, English has often been used as an intermediary language, meaning texts originally written in languages are first translated into English, and then from English into other languages, resulting in most target texts being translations of translations \cite{lefer-2020-parallel-corpora}.
+This is called indirect translation, and posed for many problems in translations as terms can be ambiguous with varying or similar meanings \cite{rosa-2017-indirect-translation-problems}.
+
+Neural Machine Translation (NMT) systems require vast amounts of training data, and thus the availability of parallel corpora is crucial for building effective models \cite{koehn-2017-challenges}.
+The lack of extensive parallel corpora, especially for low-resource languages, leads to suboptimal performance in NMT techniques compared to their high-resource counterparts \cite{ranathunga-2023-nmt-low-res}.
+OPUS \cite{opus} is a comprehensive collection of open-source parallel corpora used extensively in the field of machine translation (MT).
+It includes corpora for 744 languages and contains over 1,210 different datasets, amassing a total of 45,945,946,108 sentence pairs.
+Tatoeba \cite{tatoeba} is another prominent resource in the field of MT and NLP, known for its extensive collection of translated sentences.
+As of July 2024, it contains 12,186,207 sentences over 423 supported languages, growing daily through volunteer contributions.
+
+Several datasets are often used for evaluation in machine translation.
+The Tatoeba Challenge \cite{tiedemann-2020-tatoeba-challenge} covers 487 languages in 4,024 language pairs, including 657 test sets sourced from Tatoeba website, covering 138 languages.
+The TED Talks dataset \cite{ye-2018-word-embeddings-ted} contains transcripts from TED talks for more than 50 languages.
+IWSLT \cite{agarwal-iwstl-2023} also contains TED talks data, but paired with English translations.
+Finally, WMT \cite{barrault-2020-wmt} is an annual event that organises tasks for machine translation and provides a collection of datasets for benchmarking and evaluating translation systems.
+
+While these datasets can be used to evaluate translation performance between any language pairs, they do not contain parallel corpora between different languages.
+Corpus between English and French for example, contains different text to the corpus between English and Spanish.
+Therefore, they do not support easy evaluation for many-to-English translations.
+
+Furthermore, papers often use different sets and different versions of datasets to evaluate their model performance, incurring unstraightforward comparison.
+
+mBart \cite{liu-2020-mbart} uses WMT19 for English-German translation and TED15 for Chinese-English translation.
+M2M-100 \cite{fan-2020-m2m100} uses 7 different datasets shared across languages.
+NLLB-200  \cite{nllb200-2020} uses FLORES dataset, designed for low-resources languages.
+OPUS-MT \cite{tiedemann-2023-democratizing,tiedemann-2020-opus-mt} models use datasets from Tatoeba challenges.
 
 \subsection{Evaluation Metrics}
+
+Other metrics for machine translation exist, but not used nearly as widely as BLEU.
+Crosslingual Optimized Metric for Evaluation of Translation (COMET) \cite{rei-2020-comet} is a recent metric that leverages neural-network-based models for deeper understanding and evaluation.
+BERTScore.
+
+In this project, a variant of BLEU called SacreBLEU and METEOR will be used due to their simplicity.
+
+\subsubsection{BLEU}
 
@@
 
-While other metrics exist for machine translation such as METEOR \cite{lavie-2007-meteor} and COMET \cite{rei-2020-comet}, it is far more practical to implement BLEU due to its popular usage in other works.
+The machine translation community's rely heavily BLEU score, however, it has several drawbacks.
+The metric has been reported to not correlate strongly with human judgement, showing variations in translation that could mean that a higher BLEU score does not necessarily indicate a true enhancement in translation quality \cite{callison-burch-2006-reevaluating-bleu}.
 
-% Introduce the concept of evaluation metrics for MT, explaining why BLEU is chosen for this study and if there are other metrics that could complement it (like METEOR or ROUGE).
+Furthermore, It is challenging to directly compare BLEU scores between paper \cite{post-2018-clarity-bleu}.
+Thus, the author proposed a variation called SacreBLEU \cite{post-2018-clarity-bleu}, which claimed to facilitate easy computation of BLEU scores that are shareable, comparable, and reproducible.
+This is done by including preprocessing steps and normalisation routines.
 
-% \begin{comment}
-% PolyLM \cite{wei-2023-polylm} (32 languages), and
-% mT5 \cite{xue-2021-mt5} (101 languages).
+\subsubsection{METEOR}
 
-% \subsection{Sequence-to-Sequence Models}
+Metric for Evaluation of Translation with Explicit ORdering (METEOR) \cite{lavie-2007-meteor} assesses a translation by calculating a score that reflects explicit word-to-word matches between the reference and a candidate translation \cite{agarwal-2008-meteor-mbleu-mter}.
+It is designed to address some limitations of the BLEU score, allowing matches between simple morphological variants and synonyms.
 
-% GPT \cite{openai-2024-gpt4}
+\begin{equation}
+    \text{METEOR} = (1 - \gamma \cdot \text{frag}) \cdot \frac{P \cdot R}{\alpha \cdot P + (1 - \alpha) \cdot R}
+\end{equation}
 
-% mBERT \cite{devlin-2019-bert} is the multilingual version of BERT, introduced in the same paper, trained on 104 different languages.
-However, the model been shown to suffer on low-resource languages \cite{wu-2020-mbert-are-all}.
-
-% T5 \cite{raffel-2023-t5} is a Text-to-Text Transfer Transformer model trained on English.
-
-% MistralAI \cite{jiang-2023-mistral-7b}
-
-% \end{comment}
+Where:
+\begin{itemize}
+    \item \(P\) is the precision,
+    \item \(R\) is the recall,
+    \item \(\text{frag}\) is the fragmentation penalty,
+    \item \(\gamma\) is a parameter that controls the weight of the fragmentation penalty, commonly 0.5,
+    \item \(\alpha\) is a parameter that controls the balance between precision and recall, commonly 0.9,
+\end{itemize}
 
@@
     \centering
-    \begin{tabular}{|l|l|}
+    \begin{tabular}{|l|l|r|}
         \hline
-        \textbf{No.} & \textbf{Language} \\
+        \textbf{No.} & \textbf{Language} & \textbf{Sentence Pairs Count} \\
         \hline
-        1            & Dutch             \\
-        2            & Finnish           \\
-        3            & French            \\
-        4            & German            \\
-        5            & Hebrew            \\
-        6            & Hungarian         \\
-        7            & Italian           \\
-        8            & Japanese          \\
-        9            & Mandarin Chinese  \\
-        10           & Polish            \\
-        11           & Russian           \\
-        12           & Spanish           \\
-        13           & Turkish           \\
-        14           & Ukrainian         \\
+        1            & Chinese           & 68,814                        \\
+        2            & Dutch             & 155,856                       \\
+        3            & Finnish           & 102,202                       \\
+        4            & French            & 405,088                       \\
+        5            & German            & 501,145                       \\
+        6            & Hebrew            & 172,082                       \\
+        7            & Hungarian         & 171,698                       \\
+        8            & Italian           & 624,160                       \\
+        9            & Japanese          & 270,116                       \\
+        10           & Polish            & 77,345                        \\
+        11           & Russian           & 722,837                       \\
+        12           & Spanish           & 265,253                       \\
+        13           & Turkish           & 710,279                       \\
+        14           & Ukrainian         & 214,244                       \\
         \hline
@@
 Then for each language, sentence pairs between English and source languages are downloaded individually and compiled.
-The result is a single Dataframe containing 1241 sentences in all 14 languages, this will be treated as a test set to evaluate the models performance on each language.
+The result is a single Dataframe containing 1,323 parallel sentences in all 14 languages, this will be treated as a test set to evaluate the models performance on each language.
 
-Sentences typically consist of everyday phrases such as 'I have to go to sleep', 'That is intriguing', and 'Where do you live?' They may also include single-word exclamations like 'Speak!' or 'Look!' Additionally, multiple sentences such as 'You may write in any language you want.
+Sentences typically consist of everyday phrases such as 'I have to go to sleep', 'That is intriguing', and 'Where do you live?'.
+They may also include single-word exclamations like 'Speak!' or 'Look!'.
+Additionally, multiple sentences such as 'You may write in any language you want.
 On Tatoeba, all languages are considered equal', and 'Guns don't kill people.
@@
 Therefore, machine translation process should be straightforward on this level.
+
+\begin{CJK}{UTF8}{gbsn}
+    \begin{table}[htbp]
+        \centering
+        \begin{tabular}{|l|l|l|}
+            \hline
+            \textbf{Language} & \textbf{Sentence 1}                   & \textbf{Sentence 2}                   \\
+            \hline
+            English           & I have to go to sleep.                & So what?                              \\
+            Chinese           & 我该去睡觉了。                               & 那又怎樣?                                 \\
+            Dutch             & Ik moet gaan slapen.                  & Dus?                                  \\
+            Finnish           & Minun täytyy mennä nukkumaan.         & Mitä sitten?                          \\
+            French            & Je dois aller dormir.                 & Et alors ?                            \\
+            German            & Ich muss jetzt schlafen.              & Na und?                               \\
+            Hebrew            & <hidden-due-to-latex-incompatibility> & <hidden-due-to-latex-incompatibility> \\
+            Hungarian         & Aludni kell mennem.                   & És akkor mi van?                      \\
+            Italian           & Devo andare a dormire.                & E allora?                             \\
+            Japanese          & 私は眠らなければなりませ                          & だから何？                                 \\
+            Polish            & Muszę iść spać.                       & No i co?                              \\
+            Russian           & Мне пора идти                         & Так что?                              \\
+            Spanish           & Tengo que irme a dormir.              & ¿Entonces qué?                        \\
+            Turkish           & Yatmaya gitmek zorundayım.            & Öyleyse ne yapmalı?                   \\
+            Ukrainian         & Маю піти спати.                       & Ну то що?                             \\
+            \hline
+        \end{tabular}
+        \caption{A snippet of the dataset}
+        \label{table:parallel_sentence}
+    \end{table}
+\end{CJK}
+
+Figure \ref{table:parallel_sentence} shows a few examples of parallel sentences in the final dataset.
 
@@
 
+\begin{table}[htbp]
+    \centering
+    \begin{minipage}{0.49\linewidth}
+        \footnotesize
+        \begin{tabular}{|l|r|r|r|}
+            \hline
+            \textbf{Language} & \textbf{BLEU} & \textbf{SacreBLEU} & \textbf{METEOR} \\
+            \hline
+            Chinese           & 0.0975        & 43.3490            & 0.6848          \\
+            Dutch             & 0.0976        & 52.6920            & 0.7448          \\
+            Finnish           & 0.0975        & 49.9149            & 0.7145          \\
+            French            & 0.0969        & 51.4090            & 0.7288          \\
+            German            & 0.0969        & 52.7877            & 0.7423          \\
+            Hebrew            & 0.0981        & 50.3287            & 0.7198          \\
+            Italian           & 0.0975        & 53.1644            & 0.7341          \\
+            Japanese          & 0.0959        & 41.6610            & 0.6511          \\
+            Polish            & 0.0984        & 52.5151            & 0.7301          \\
+            Russian           & 0.0976        & 48.2725            & 0.7027          \\
+            Spanish           & 0.0977        & 52.9323            & 0.7371          \\
+            Turkish           & 0.0978        & 51.3009            & 0.7262          \\
+            Ukrainian         & 0.0978        & 46.5555            & 0.6890          \\
+            \hline
+        \end{tabular}
+        \caption{M2M-100}
+    \end{minipage}
+    \begin{minipage}{0.49\linewidth}
+        \footnotesize
+        \begin{tabular}{|l|r|r|r|}
+            \hline
+            \textbf{Language} & \textbf{BLEU} & \textbf{SacreBLEU} & \textbf{METEOR} \\
+            \hline
+            Chinese           & 0.0962        & 54.7322            & 0.7600          \\
+            Dutch             & 0.0970        & 63.6482            & 0.8007          \\
+            Finnish           & 0.0963        & 47.2194            & 0.7067          \\
+            French            & 0.0968        & 57.2482            & 0.7598          \\
+            German            & 0.0967        & 63.1666            & 0.8022          \\
+            Hebrew            & 0.0972        & 58.0846            & 0.7590          \\
+            Italian           & 0.0972        & 65.9415            & 0.8068          \\
+            Japanese          & 0.0943        & 43.9547            & 0.7151          \\
+            Polish            & 0.0978        & 62.7550            & 0.7923          \\
+            Russian           & 0.0974        & 58.8820            & 0.7686          \\
+            Spanish           & 0.0883        & 35.1593            & 0.7306          \\
+            Turkish           & 0.0975        & 50.9377            & 0.6982          \\
+            Ukrainian         & 0.0964        & 55.8637            & 0.7496          \\
+            \hline
+        \end{tabular}
+        \caption{mBART50}
+    \end{minipage}
+\end{table}
+
+\begin{table}[htbp]
+    \centering
+    \begin{minipage}{0.49\linewidth}
+        \footnotesize
+        \begin{tabular}{|l|r|r|r|}
+            \hline
+            \textbf{Language} & \textbf{BLEU} & \textbf{SacreBLEU} & \textbf{METEOR} \\
+            \hline
+            Chinese           & 0.0949        & 50.4454            & 0.7180          \\
+            Dutch             & 0.0915        & 20.8988            & 0.2899          \\
+            Finnish           & 0.0921        & 32.3926            & 0.4006          \\
+            French            & 0.0950        & 32.1389            & 0.4338          \\
+            German            & 0.0903        & 7.1436             & 0.1528          \\
+            Hebrew            & 0.0897        & 2.3455             & 0.1412          \\
+            Italian           & 0.0926        & 24.2702            & 0.3059          \\
+            Japanese          & 0.0842        & 1.9569             & 0.0919          \\
+            Polish            & 0.0937        & 38.4395            & 0.4739          \\
+            Russian           & 0.0918        & 20.6734            & 0.2977          \\
+            Spanish           & 0.0929        & 26.3313            & 0.3578          \\
+            Turkish           & 0.0945        & 45.2721            & 0.6187          \\
+            Ukrainian         & 0.0914        & 9.0340             & 0.1613          \\
+            \hline
+        \end{tabular}
+        \caption{NLLB-200}
+    \end{minipage}
+    \begin{minipage}{0.49\linewidth}
+        \footnotesize
+        \begin{tabular}{|l|r|r|r|}
+            \hline
+            \textbf{Language} & \textbf{BLEU} & \textbf{SacreBLEU} & \textbf{METEOR} \\
+            \hline
+            Chinese           & 0.0971        & 59.5098            & 0.7953          \\
+            Dutch             & 0.0974        & 69.8803            & 0.8471          \\
+            Finnish           & 0.0965        & 66.6267            & 0.8296          \\
+            French            & 0.0967        & 69.8185            & 0.8357          \\
+            German            & 0.0969        & 69.7422            & 0.8419          \\
+            Hebrew            & 0.0977        & 66.5149            & 0.8229          \\
+            Italian           & 0.0979        & 74.1298            & 0.8584          \\
+            Japanese          & 0.0960        & 63.1435            & 0.7893          \\
+            Polish            & 0.0952        & 61.9026            & 0.8425          \\
+            Russian           & 0.0974        & 66.7046            & 0.8179          \\
+            Spanish           & 0.0979        & 71.4174            & 0.8463          \\
+            Turkish           & 0.0980        & 72.6551            & 0.8460          \\
+            Ukrainian         & 0.0978        & 75.4447            & 0.8667          \\
+            \hline
+        \end{tabular}
+        \caption{OPUS-MT}
+    \end{minipage}
+\end{table}
+
+\begin{figure}[htbp]
+    \centering
+    \includegraphics[width=1\linewidth]{figures/metrics_bar.png}
+    \caption{BLEU, SacreBLEU, and METEOR scores of each model on each language.}
+    \label{fig:result_visual}
+\end{figure}
+
+Figure \ref{fig:result_visual} shows the visualisation of all scores.
+
 \section{Conclusion}

### 2024-08-05 analyse/analyse_results.py

@@ -1,0 +1,76 @@
+import json
+import os
+from collections import Counter
+
+import matplotlib.pyplot as plt
+import nltk
+import numpy as np
+import pandas as pd
+import seaborn as sns
+import spacy
+from sentence_transformers import SentenceTransformer
+from sklearn.manifold import TSNE
+from textblob import TextBlob
+from wordcloud import WordCloud
+
+LANGS = [
+    "chinese",
+    "dutch",
+    "finnish",
+    "french",
+    "german",
+    "hebrew",
+    "italian",
+    "japanese",
+    "polish",
+    "russian",
+    "spanish",
+    "turkish",
+    "ukrainian",
+]
+
+MODELS = [
+    "opus",
+    "mbart-large-50-many-to-many-mmt",
+    "nllb-200-distilled-600M",
+    "m2m100_418M",
+]
+
+with open("../inference/result/metrics.json", "r") as f:
+    metrics = json.loads(f.read())
+
+
+def plot_scores(dataframe, metric_name, ax, add_legend=False):
+    bar_width = 0.2
+    index = np.arange(len(dataframe.index))
+
+    for i, model in enumerate(dataframe.columns):
+        ax.bar(index + i * bar_width, dataframe[model], bar_width, label=model)
+
+    ax.set_ylabel(f"{metric_name} Score")
+    ax.set_title(f"{metric_name} Scores by Language and Model")
+    ax.set_xticks(index + bar_width * (len(dataframe.columns) - 1) / 2)
+    ax.set_xticklabels(dataframe.index, rotation=45, ha="right")
+    if add_legend:
+        ax.legend(loc="upper right", fontsize="small")
+
+
+bleu_scores = pd.DataFrame(index=LANGS)
+sacre_bleu_scores = pd.DataFrame(index=LANGS)
+meteor_bleu_scores = pd.DataFrame(index=LANGS)
+
+for model in MODELS:
+    bleu_scores[model] = [metrics[model][lang]["bleu"] for lang in LANGS]
+    sacre_bleu_scores[model] = [metrics[model][lang]["sacre_bleu"] for lang in LANGS]
+    meteor_bleu_scores[model] = [metrics[model][lang]["meteor"] for lang in LANGS]
+
+fig, (ax1, ax2, ax3) = plt.subplots(3, 1, figsize=(12, 10), sharex=True)
+
[…]

### 2024-08-05 bleu.py

@@ -1,45 +1,0 @@
-import nltk
-from nltk.translate.bleu_score import corpus_bleu, sentence_bleu
-
-# Ensure you've downloaded the necessary NLTK data
-nltk.download("punkt")
-
-
-def calculate_bleu(reference_sentences, candidate_sentences):
-    """
-    Calculate BLEU score for a list of reference and candidate sentences.
-
-    :param reference_sentences: List of reference sentences (list of lists of words)
-    :param candidate_sentences: List of candidate sentences (list of words)
-    :return: BLEU score
-    """
-    # Tokenize sentences
-    reference_tokenized = [nltk.word_tokenize(sent) for sent in reference_sentences]
-    candidate_tokenized = [nltk.word_tokenize(sent) for sent in candidate_sentences]
-
-    # Calculate BLEU scores
-    individual_bleu_scores = [
-        sentence_bleu([ref], cand)
-        for ref, cand in zip(reference_tokenized, candidate_tokenized)
-    ]
-    corpus_bleu_score = corpus_bleu(
-        [[ref] for ref in reference_tokenized], candidate_tokenized
-    )
-
-    return individual_bleu_scores, corpus_bleu_score
-
-
-# Example usage
-reference_sentences = [
-    "The quick brown fox jumps over the lazy dog.",
-    "A fast brown fox leaps over a sleepy dog.",
-]
-candidate_sentences = [
-    "The quick brown fox jumps over the lazy dog.",
-    "A quick brown fox jumps over the lazy dog.",
-]
-
-individual_scores, overall_score = calculate_bleu(
-    reference_sentences, candidate_sentences
-)
-prin

### 2024-08-05 inference/metrics.py

@@ -1,0 +1,81 @@
+import json
+
+import nltk
+import numpy as np
+import pandas as pd
+import sacrebleu
+from nltk.translate import meteor_score
+from nltk.translate.bleu_score import SmoothingFunction, corpus_bleu
+from tqdm import tqdm
+
+# nltk.download("punkt_tab")
+# nltk.download("wordnet")
+
+SRC_LANGS = [
+    "chinese",
+    "dutch",
+    "finnish",
+    "french",
+    "german",
+    "hebrew",
+    "italian",
+    "japanese",
+    "polish",
+    "russian",
+    "spanish",
+    "turkish",
+    "ukrainian",
+]
+
+PATHS = [
+    "opus",
+    "mbart-large-50-many-to-many-mmt",
+    "nllb-200-distilled-600M",
+    "m2m100_418M",
+]
+
+
+def calculate_metric(reference, candidate):
+    ref_tokenised = [nltk.word_tokenize(ref) for ref in reference]
+    cand_tokenised = [nltk.word_tokenize(cand) for cand in candidate]
+
+    meteor = np.mean(
+        [
+            meteor_score.single_meteor_score(ref, cand)
+            for ref, cand in zip(ref_tokenised, cand_tokenised)
+        ]
+    )
+
+    sacre_bleu = sacrebleu.corpus_bleu(candidate, [reference]).score
+
+    bleu = corpus_bleu(
+        ref_tokenised, cand_tokenised, smoothing_function=SmoothingFunction().method7
+    )
+
+    return bleu, sacre_bleu, meteor
+
+
+metrics = {}
+
+for path in PATHS:
+    print(f"Calculating {path}...")
+    df = pd.read_csv(f"./result/{path}.csv", index_col=0)
+    metrics[path] = {}
+
+    reference = df["english"].tolist()
+
+    for src_lang in tqdm(SRC_LANGS):
+        candidate = df[f"translated_from_{src_lang}"].tolist()
+        bleu, sacre_blue, meteor = calculate_metric(reference, candidate)
+        metrics[path][src_lang] = {
+            "bleu": bleu,
+            "sacre_bleu": sacre_blue,
+            "meteor": meteor,
+        }
+
+
+print(metrics)
+
+
+with open("./result/metrics.json", "w") as f:
+    json.dump(metrics, f, indent=4, sort_keys=True)

### 2024-08-05 inference/mrasp2.py

@@ -12,18 +12,18 @@ MODEL_NAME = "chinmaydan/mRASP2"
 
 LANG_CODES = {
     "chinese": "zho_Hans",
-    "dutch": "nld_Latn",
-    "finnish": "fin_Latn",
-    "french": "fra_Latn",
-    "german": "deu_Latn",
-    "hebrew": "he_IL",
-    "italian": "ita_Latn",
-    "japanese": "jpn_Jpan",
-    "polish": "pol_Latn",
-    "russian": "rus_Cyrl",
-    "spanish": "spa_Latn",
-    "turkish": "tur_Latn",
-    "ukrainian": "ukr_Cyrl",
+    # "dutch": "nld_Latn",
+    # "finnish": "fin_Latn",
+    # "french": "fra_Latn",
+    # "german": "deu_Latn",
+    # "hebrew": "he_IL",
+    # "italian": "ita_Latn",
+    # "japanese": "jpn_Jpan",
+    # "polish": "pol_Latn",
+    # "russian": "rus_Cyrl",
+    # "spanish": "spa_Latn",
+    # "turkish": "tur_Latn",
+    # "ukrainian": "ukr_Cyrl",
 }
 
 tqdm.pandas()
@@ -61,6 +61,7 @@ for src_lang, src_lang_code in LANG_CODES.items():
         batch = df[src_lang][i : i + batch_size].tolist()
         translations = translate_batch(batch)
         translated_texts.extend(translations)
+        break
 
     translated_df[f"translated_from_{src_lang}"] = translated_texts
 

### 2024-08-08 paper/paper.tex
updated

@@
 
+% Language barriers
+
 \section{Literature Review}
@@
 
-In terms of pre-trained models (PTMs) for machine translation, it can be divided into two categories: one-to-one models and many-to-many models (multilingual).
+In terms of pre-trained models (PTMs) for machine translation, most PTMs support two kinds of translations: one-to-one translation and many-to-many translation (multilingual models).
 
@@
 This setup is characterised by having a dedicated model for each unique language pair.
-An example of this setup is OPUS-MT by Helsinki-NLP \cite{tiedemann-2023-democratizing,tiedemann-2020-opus-mt}, which provides over 1,000 pre-trained models for translation between numerous language pairs.
+OPUS-MT \cite{tiedemann-2020-opus-mt} and Facebook Fairseq \cite{ott-2019-fairseq} are examples of PTMs that support one-to-one translation.
 
@@
 This approach enables many-to-many translation, where a single model can translate between multiple source and target languages \cite{aharoni-2019-massively-multilingual}.
-mBART \cite{liu-2020-mbart} is a sequence-to-sequence denoising auto-encoder model specifically designed for multilingual tasks.
-The mBART-50 variant supports many-to-many translations for over 50 languages.
-M2M-100 \cite{fan-2020-m2m100} is designed to perform direct translation between 100 languages without relying on English as an intermediate language.
-NLLB-200: NLLB-200 \cite{nllb200-2020} is built to handle translation tasks across a broad spectrum of languages, including many that are low-resource or underrepresented in existing datasets.
-It supports translations for 200 languages, encompassing numerous underrepresented languages.
-mBART, M2M-100, and NLLB-200 were all developed by Meta AI (formerly Facebook AI), showcasing the organisation's significant impact in the machine translation field through pre-trained models (PTMs).
-These models represent a substantial advancement in multilingual translation capabilities.
+
+mBART \cite{liu-2020-mbart}, M2M \cite{fan-2020-m2m100}, NLLB-200 \cite{nllb200-2020}, are examples of PTMs that support many-to-many translation language.
+This domain seems to be dominated by META AI.
+
+% Back translation?
 
@@
 OPUS-MT \cite{tiedemann-2023-democratizing,tiedemann-2020-opus-mt} models use datasets from Tatoeba challenges.
+
+\section{Methodology}
+
+\subsection{Dataset}
+
+Tatoeba is a vast, continuously expanding database consisting sentences and their translations, built through the contributions of thousands of volunteers, offering a tool that allows users to see examples of how words are used in sentences \cite{tatoeba}.
+They currently have 12,132,349 sentences and 423 supported languages, with around one to two thousand new sentences added daily, on average.
+The English sentence dataset contains 1,905,089 sentences, the largest one in their repository, with Russian in the second place with 1,066,633 sentences.
+Some languages supported on the website is shown in Figure \ref{fig:tatoeba_languages} and Figure \ref{fig:tatoeba_top_bottom_languages}, sorted from the biggest corpus.
+
+\begin{figure}[htbp]
+    \centering
+    \includegraphics[width=0.9\linewidth]{images/tatoeba_languages.png}
+    \caption{Tatoeba's languages repository with 10,000+ sentences and 100,000+ sentences \cite{tatoeba}}
+    \label{fig:tatoeba_languages}
+\end{figure}
+
+\begin{figure}[htbp]
+    \centering
+    \includegraphics[width=0.5\linewidth]{images/tatoeba_top_20_lang.png}
+    \includegraphics[width=0.46\linewidth]{images/tatoeba_bottom_20_lang.png}
+    \caption{Tatoeba top 20 and bottom 20 languages based on sentences count \cite{tatoeba}}
+    \label{fig:tatoeba_top_bottom_languages}
+\end{figure}
+
+\begin{table}[htbp]
+    \centering
+    \begin{tabular}{|l|l|r|}
+        \hline
+        \textbf{No.} & \textbf{Language} & \textbf{Sentence Pairs Count} \\
+        \hline
+        1            & Chinese           & 68,814                        \\
+        2            & Dutch             & 155,856                       \\
+        3            & Finnish           & 102,202                       \\
+        4            & French            & 405,088                       \\
+        5            & German            & 501,145                       \\
+        6            & Hebrew            & 172,082                       \\
+        7            & Hungarian         & 171,698                       \\
+        8            & Italian           & 624,160                       \\
+        9            & Japanese          & 270,116                       \\
+        10           & Polish            & 77,345                        \\
+        11           & Russian           & 722,837                       \\
+        12           & Spanish           & 265,253                       \\
+        13           & Turkish           & 710,279                       \\
+        14           & Ukrainian         & 214,244                       \\
+        \hline
+    \end{tabular}
+    \caption{List of chosen languages for evaluation}
+    \label{table:eval_languages}
+\end{table}
+
+Table \ref{table:eval_languages} show the 14 languages selected for this project.
+Languages are chosen based on its resources' availability in Tatoeba, as well as considering supported languages in most PTMs models.
+To build the dataset, sentences in English are first downloaded, containing 1,898,494 sentences (it is unclear why it is less than the number stated in the Tatoeba website).
+Then for each language, sentence pairs between English and source languages are downloaded individually and compiled.
+The result is a single Dataframe containing 1,323 parallel sentences in all 14 languages, this will be treated as a test set to evaluate the models performance on each language.
+
+Sentences typically consist of everyday phrases such as 'I have to go to sleep', 'That is intriguing', and 'Where do you live?'.
+They may also include single-word exclamations like 'Speak!' or 'Look!'.
+Additionally, multiple sentences such as 'You may write in any language you want.
+On Tatoeba, all languages are considered equal', and 'Guns don't kill people.
+People kill people' can be found inside the corpus.
+A few of them also include human names, 'Compare your answer with Tom's', 'Muiriel is 20 now'.
+All of the sentences are straightforward and literal, without the use of linguistic features such as metaphors or sarcasm.
+Therefore, machine translation process should be straightforward on this level.
+
+\begin{CJK}{UTF8}{gbsn}
+    \begin{table}[htbp]
+        \centering
+        \begin{tabular}{|l|l|l|}
+            \hline
+            \textbf{Language} & \textbf{Sentence 1}                   & \textbf{Sentence 2}                   \\
+            \hline
+            English           & I have to go to sleep.                & So what?                              \\
+            Chinese           & 我该去睡觉了。                               & 那又怎樣?                                 \\
+            Dutch             & Ik moet gaan slapen.                  & Dus?                                  \\
+            Finnish           & Minun täytyy mennä nukkumaan.         & Mitä sitten?                          \\
+            French            & Je dois aller dormir.                 & Et alors ?                            \\
+            German            & Ich muss jetzt schlafen.              & Na und?                               \\
+            Hebrew            & <hidden-due-to-latex-incompatibility> & <hidden-due-to-latex-incompatibility> \\
+            Hungarian         & Aludni kell mennem.                   & És akkor mi van?                      \\
+            Italian           & Devo andare a dormire.                & E allora?                             \\
+            Japanese          & 私は眠らなければなりませ                          & だから何？                                 \\
+            Polish            & Muszę iść spać.                       & No i co?                              \\
+            Russian           & Мне пора идти                         & Так что?                              \\
+            Spanish           & Tengo que irme a dormir.              & ¿Entonces qué?                        \\
+            Turkish           & Yatmaya gitmek zorundayım.            & Öyleyse ne yapmalı?                   \\
+            Ukrainian         & Маю піти спати.                       & Ну то що?                             \\
+            \hline
+        \end{tabular}
+        \caption{A snippet of the dataset}
+        \label{table:parallel_sentence}
+    \end{table}
+\end{CJK}
+
+Figure \ref{table:parallel_sentence} shows two examples of parallel sentences in the final dataset, across 14 languages.
+
+\begin{figure}[htbp]
+    \centering
+    \includegraphics[width=1\linewidth]{figures/word_count_box.png}
+    \caption{Dataset word count distribution, consisting of mostly short sentences.}
+    \label{fig:word_count_box}
+\end{figure}
+
+Figure \ref{fig:word_count_box} shows a box plot of sentences word count.
+It can be seen that the majority of sentences are short.
+Chinese and Japanese are counted per letter in this figure, which accounts for the variability seen in the Figure/
+
+Most languages cluster around 4-6 words per sentence. while Chinese cluster around 6-8 letters and Japanese cluster around 8-12 letters.
+
+\subsection{Proposed Pre-Trained Models}
+
+\subsubsection{OPUS-MT}
+
+\textbf{OPUS-MT} \cite{tiedemann-2023-democratizing,tiedemann-2020-opus-mt} provides over 1,000 pre-trained models for translation between numerous language pairs.
+The architecture is based on MARIAN-NMT \cite{mariannmt}, based on a standard transformer setup: 6 self-attentive layers in both the encoder and decoder networks, each with 8 attention heads per layer \cite{tiedemann-2020-opus-mt}.
+
+While OPUS-MT support both monolingual and multilingual PTMs, the multilingual models seem to lack proper documentation and therefore excluded from this experiment.
+
+\subsubsection{mBART-50}
+
+\textbf{mBART} \cite{liu-2020-mbart} is a sequence-to-sequence denoising auto-encoder model specifically designed for multilingual tasks.
+The mBART-50 variant supports many-to-many translations for over 50 languages.
+
+\begin{figure}[htbp]
+    \centering
+    \includegraphics[width=0.9\linewidth]{images/mbart.png}
+    \caption{mBART framework for Multilingual Denoising Pre-training (left) and fine-tuning on downstream MT tasks
+        (right) \cite{liu-2020-mbart}}
+    \label{fig:mbart}
+\end{figure}
+
+Figure \ref{fig:mbart} shows the architecture of mBART's key featuresL multilingual denoising pre-training and fine-tuning.
+It is shown that the model achieved consistent performance gains through pre-training in low-to-medium resource sentence level MT \cite{liu-2020-mbart}.
+"Denoising" here refers to a pre-training technique used to improve the model's ability to generate coherent and accurate translations by training it to reconstruct original sentences from corrupted (noisy) versions.
+
+\subsubsection{M2M-100}
+
+\textbf{M2M-100} \cite{fan-2020-m2m100} is designed to perform direct translation between 100 languages without relying on English as an intermediate language.
+
+\begin{figure}[htbp]
+    \centering
+    \includegraphics[width=0.9\linewidth]{images/m2m.png}
+    \caption{Summary of M2M dataset and multilingual model.}
+    \label{fig:m2m}
+\end{figure}
+
+% (English-centric data (top left) includes only training data involving English, while our many-to-many multilingual setting (top right) encompasses direct translation between various languages, and our proposed M2M-100 model (bottom) integrates dense and sparse language-specific parameters for direct translation between languages.)
+
+% (Multilingual translation models factorize computation when translating to many
+% languages and share information between similar languages, which benefits low resource
+% directions (Arivazhagan et al., 2019)) \cite{arivazhagan-2019-massively-multilingual}
+
+\subsubsection{NLLB-200}
+
+\textbf{NLLB-200} \cite{nllb200-2020} is built to handle translation tasks across a broad spectrum of languages, including many that are low-resource or underrepresented in existing datasets.
+It supports translations for 200 languages, encompassing numerous underrepresented languages.
+mBART, M2M-100, and NLLB-200 were all developed by Meta AI (formerly Facebook AI), showcasing the organisation's significant impact in the machine translation field through pre-trained models (PTMs).
+These models represent a substantial advancement in multilingual translation capabilities.
+
+\begin{figure}[htbp]
+    \centering
+    \includegraphics[width=0.9\linewidth]{images/nllb.png}
+    \caption{NLLB pipeline \cite{nllb200-2020}}
+    \label{fig:nllb}
+\end{figure}
+
+% (No Language Left Behind: Our low-resource translation effort focuses on four
+% cornerstones. (1) We strive to understand the low-resource translation problem from the
+% perspective of native speakers. (2) We study how to automatically create training data to
+% move low-resource languages towards high-resource. (3) We utilize this data to create
+% state-of-the-art translation models. (4) We evaluate every language we aim to translate.)
+
+\subsection{Inference Details}
+
+OPUS-MT models are taken from the Helsinki-NLP repository.
+For mBART, the 'mbart-large-50-many-to-many-mmt' is used, 'm2m100\_418M', and 'nllb-200-distilled-600M'.
+All models are taken using the transformers \cite{wolf-etal-2020-transformers} library from HuggingFace \cite{huggingface}.
+
+Batch size of 4
+
+The translation are first generated and saved into a CSV file, before calculating the metrics score.
 
@@
 
-\section{Experiments}
-
-\subsection{Dataset}
-
-Tatoeba is a vast, continuously expanding database consisting sentences and their translations, built through the contributions of thousands of volunteers, offering a tool that allows users to see examples of how words are used in sentences \cite{tatoeba}.
-They currently have 12,132,349 sentences and 423 supported languages, with around one to two thousand new sentences added daily, on average.
-The English sentence dataset contains 1,905,089 sentences, the largest one in their repository, with Russian in the second place with 1,066,633 sentences.
-Some languages supported on the website is shown in Figure \ref{fig:tatoeba_languages} and Figure \ref{fig:tatoeba_top_bottom_languages}, sorted from the biggest corpus.
+\section{Evaluation}
 
@@
     \centering
-    \includegraphics[width=0.9\linewidth]{images/tatoeba_languages.png}
-    \caption{Tatoeba's languages repository with 10,000+ sentences and 100,000+ sentences \cite{tatoeba}}
-    \label{fig:tatoeba_languages}
+    \includegraphics[width=1\linewidth]{figures/metrics_bar.png}
+    \caption{BLEU, SacreBLEU, and METEOR scores of each model on each language.}
+    \label{fig:result_visual}
 \end{figure}
 
-\begin{figure}[htbp]
-    \centering
-    \includegraphics[width=0.5\linewidth]{images/tatoeba_top_20_lang.png}
-    \includegraphics[width=0.46\linewidth]{images/tatoeba_bottom_20_lang.png}
-    \caption{Tatoeba top 20 and bottom 20 languages based on sentences count \cite{tatoeba}}
-    \label{fig:tatoeba_top_bottom_languages}
-\end{figure}
+Figure \ref{fig:result_visual} shows the visualisation of all scores.
+
+OPUS-MT models outperform other models at every language on SacreBLEU and METEOR scores.
+
+mBart50 performed second best, being the best-performing multilingual models in this study.
+
+NLLB-200 performed the worse, as the model is more concerned with translations on low-resource languages.
 
@@
     \centering
-    \begin{tabular}{|l|l|r|}
-        \hline
-        \textbf{No.} & \textbf{Language} & \textbf{Sentence Pairs Count} \\
-        \hline
-        1            & Chinese           & 68,814                        \\
-        2            & Dutch             & 155,856                       \\
-        3            & Finnish           & 102,202                       \\
-        4            & French            & 405,088                       \\
-        5            & German            & 501,145                       \\
-        6            & Hebrew            & 172,082                       \\
-        7            & Hungarian         & 171,698                       \\
-        8            & Italian           & 624,160                       \\
-        9            & Japanese          & 270,116                       \\
-        10           & Polish            & 77,345                        \\
-        11           & Russian           & 722,837                       \\
-        12           & Spanish           & 265,253                       \\
-        13           & Turkish           & 710,279                       \\
-        14           & Ukrainian         & 214,244                       \\
-        \hline
-    \end{tabular}
-    \caption{List of chosen languages for evaluation}
-    \label{table:eval_languages}
-\end{table}
-
-Table \ref{table:eval_languages} show the 14 languages selected for this project.
-Languages are chosen based on its resources' availability in Tatoeba, as well as considering supported languages in most PTMs models.
-To build the dataset, sentences in English are first downloaded, containing 1,898,494 sentences (it is unclear why it is less than the number stated in the Tatoeba website).
-Then for each language, sentence pairs between English and source languages are downloaded individually and compiled.
-The result is a single Dataframe containing 1,323 parallel sentences in all 14 languages, this will be treated as a test set to evaluate the models performance on each language.
-
-Sentences typically consist of everyday phrases such as 'I have to go to sleep', 'That is intriguing', and 'Where do you live?'.
-They may also include single-word exclamations like 'Speak!' or 'Look!'.
-Additionally, multiple sentences such as 'You may write in any language you want.
-On Tatoeba, all languages are considered equal', and 'Guns don't kill people.
-People kill people' can be found inside the corpus.
-A few of them also include human names, 'Compare your answer with Tom's', 'Muiriel is 20 now'.
-All of the sentences are straightforward and literal, without the use of linguistic devices such as metaphors or sarcasm.
-Therefore, machine translation process should be straightforward on this level.
-
-\begin{CJK}{UTF8}{gbsn}
-    \begin{table}[htbp]
-        \centering
-        \begin{tabular}{|l|l|l|}
-            \hline
-            \textbf{Language} & \textbf{Sentence 1}                   & \textbf{Sentence 2}                   \\
-            \hline
-            English           & I have to go to sleep.                & So what?                              \\
-            Chinese           & 我该去睡觉了。                               & 那又怎樣?                                 \\
-            Dutch             & Ik moet gaan slapen.                  & Dus?                                  \\
-            Finnish           & Minun täytyy mennä nukkumaan.         & Mitä sitten?                          \\
-            French            & Je dois aller dormir.                 & Et alors ?                            \\
-            German            & Ich muss jetzt schlafen.              & Na und?                               \\
-            Hebrew            & <hidden-due-to-latex-incompatibility> & <hidden-due-to-latex-incompatibility> \\
-            Hungarian         & Aludni kell mennem.                   & És akkor mi van?                      \\
-            Italian           & Devo andare a dormire.                & E allora?                             \\
-            Japanese          & 私は眠らなければなりませ                          & だから何？                                 \\
-            Polish            & Muszę iść spać.                       & No i co?                              \\
-            Russian           & Мне пора идти                         & Так что?                              \\
-            Spanish           & Tengo que irme a dormir.              & ¿Entonces qué?                        \\
-            Turkish           & Yatmaya gitmek zorundayım.            & Öyleyse ne yapmalı?                   \\
-            Ukrainian         & Маю піти спати.                       & Ну то що?                             \\
-            \hline
-        \end{tabular}
-        \caption{A snippet of the dataset}
-        \label{table:parallel_sentence}
-    \end{table}
-\end{CJK}
-
-Figure \ref{table:parallel_sentence} shows a few examples of parallel sentences in the final dataset.
-
-\begin{figure}[htbp]
-    \centering
-    \includegraphics[width=1\linewidth]{figures/word_count_box.png}
-    \caption{Dataset word count distribution, consisting of mostly short sentences.}
-    \label{fig:word_count_box}
-\end{figure}
-
-Figure \ref{fig:word_count_box} shows a box plot of sentences word count. It can be seen that the majority of sentences are short.
-
-Chinese and Japanese are counted per letter.
-
-While most languages cluster around 4-6 words per sentence, there are notable exceptions like Japanese, which exhibits much less variability.
-This analysis can be useful for understanding language-specific characteristics in sentence structure, which could inform tasks like translation, text processing, or linguistic studies.
-
-\section{Evaluation}
-
-\begin{table}[htbp]
-    \centering
-    \begin{minipage}{0.49\linewidth}
-        \footnotesize
-        \begin{tabular}{|l|r|r|r|}
-            \hline
-            \textbf{Language} & \textbf{BLEU} & \textbf{SacreBLEU} & \textbf{METEOR} \\
-            \hline
-            Chinese           & 0.0975        & 43.3490            & 0.6848          \\
-            Dutch             & 0.0976        & 52.6920            & 0.7448          \\
-            Finnish           & 0.0975        & 49.9149            & 0.7145          \\
-            French            & 0.0969        & 51.4090            & 0.7288          \\
-            German            & 0.0969        & 52.7877            & 0.7423          \\
-            Hebrew            & 0.0981        & 50.3287            & 0.7198          \\
-            Italian           & 0.0975        & 53.1644            & 0.7341          \\
-            Japanese          & 0.0959        & 41.6610            & 0.6511          \\
-            Polish            & 0.0984        & 52.5151            & 0.7301          \\
-            Russian           & 0.0976        & 48.2725            & 0.7027          \\
-            Spanish           & 0.0977        & 52.9323            & 0.7371          \\
-            Turkish           & 0.0978        & 51.3009            & 0.7262          \\
-            Ukrainian         & 0.0978        & 46.5555            & 0.6890          \\
-            \hline
-        \end{tabular}
-        \caption{M2M-100}
-    \end{minipage}
     \begin{minipage}{0.49\linewidth}
@@
             Hebrew            & 0.0972          & 58.0846            & 0.7590          \\
-            Italian           & 0.0972        & 65.9415            & 0.8068          \\
+            Italian           & 0.0972          & \textbf{65.9415}   & \textbf{0.8068} \\
             Japanese          & 0.0943          & 43.9547            & 0.7151          \\
-            Polish            & 0.0978        & 62.7550            & 0.7923          \\
+            Polish            & \textbf{0.0978} & 62.7550            & 0.7923          \\
             Russian           & 0.0974          & 58.8820            & 0.7686          \\
-            Spanish           & 0.0883        & 35.1593            & 0.7306          \\
-            Turkish           & 0.0975        & 50.9377            & 0.6982          \\
+            Spanish           & \textit{0.0883} & \textit{35.1593}   & 0.7306          \\
+            Turkish           & 0.0975          & 50.9377            & \textit{0.6982} \\
             Ukrainian         & 0.0964          & 55.8637            & 0.7496          \\
@@
         \end{tabular}
-        \caption{mBART50}
+        \caption{mBART50 result.}
+        \label{table:mbart_result}
+    \end{minipage}
+    \begin{minipage}{0.49\linewidth}
+        \footnotesize
+        \begin{tabular}{|l|r|r|r|}
+            \hline
+            \textbf{Language} & \textbf{BLEU}   & \textbf{SacreBLEU} & \textbf{METEOR} \\
+            \hline
+            Chinese           & 0.0975          & 43.3490            & 0.6848          \\
+            Dutch             & 0.0976          & 52.6920            & \textbf{0.7448} \\
+            Finnish           & 0.0975          & 49.9149            & 0.7145          \\
+            French            & 0.0969          & 51.4090            & 0.7288          \\
+            German            & 0.0969          & 52.7877            & 0.7423          \\
+            Hebrew            & 0.0981          & 50.3287            & 0.7198          \\
+            Italian           & 0.0975          & \textbf{53.1644}   & 0.7341          \\
+            Japanese          & \textit{0.0959} & \textit{41.6610}   & \textit{0.6511} \\
+            Polish            & \textbf{0.0984} & 52.5151            & 0.7301          \\
+            Russian           & 0.0976          & 48.2725            & 0.7027          \\
+            Spanish           & 0.0977          & 52.9323            & 0.7371          \\
+            Turkish           & 0.0978          & 51.3009            & 0.7262          \\
+            Ukrainian         & 0.0978          & 46.5555            & 0.6890          \\
+            \hline
+        \end{tabular}
+        \caption{M2M-100 result.}
+        \label{table:m2m_result}
     \end{minipage}
@@
             \hline
-            Chinese           & 0.0949        & 50.4454            & 0.7180          \\
+            Chinese           & \textbf{0.0949} & \textbf{50.4454}   & \textbf{0.7180} \\
             Dutch             & 0.0915          & 20.8988            & 0.2899          \\
@@
             Italian           & 0.0926          & 24.2702            & 0.3059          \\
-            Japanese          & 0.0842        & 1.9569             & 0.0919          \\
+            Japanese          & \textit{0.0842} & \textit{1.9569}    & \textit{0.0919} \\
             Polish            & 0.0937          & 38.4395            & 0.4739          \\
@@
         \end{tabular}
-        \caption{NLLB-200}
+        \caption{NLLB-200 result.}
+        \label{table:nllb_result}
     \end{minipage}
@@
             \hline
-            Chinese           & 0.0971        & 59.5098            & 0.7953          \\
+            Chinese           & 0.0971          & \textit{59.5098}   & 0.7953          \\
             Dutch             & 0.0974          & 69.8803            & 0.8471          \\
@@
             Italian           & 0.0979          & 74.1298            & 0.8584          \\
-            Japanese          & 0.0960        & 63.1435            & 0.7893          \\
-            Polish            & 0.0952        & 61.9026            & 0.8425          \\
+            Japanese          & 0.0960          & 63.1435            & \textit{0.7893} \\
+            Polish            & \textit{0.0952} & 61.9026            & 0.8425          \\
             Russian           & 0.0974          & 66.7046            & 0.8179          \\
             Spanish           & 0.0979          & 71.4174            & 0.8463          \\
-            Turkish           & 0.0980        & 72.6551            & 0.8460          \\
-            Ukrainian         & 0.0978        & 75.4447            & 0.8667          \\
+            Turkish           & \textbf{0.0980} & 72.6551            & 0.8460          \\
+            Ukrainian         & 0.0978          & \textbf{75.4447}   & \textbf{0.8667} \\
             \hline
         \end{tabular}
-        \caption{OPUS-MT}
+        \caption{OPUS-MT result.}
+        \label{table:opus_result}
     \end{minipage}
@@
 
+Table \ref{table:mbart_result}, \ref{table:m2m_result}, \ref{table:nllb_result}, and \ref{table:opus_result} shows the result from all four PTMs.
+
+The standard BLEU score is very low in all cases due to the sentences being short.
+
+The mBART50 model achieved best SacreBLEU and METEOR score in Italian-to-English translation, while suffers the most in Spanish-to-English (SacreBLEU) and Turkish-to-English (METEOR) translation.
+
+Simialrly, SacreBLEU for Italian-to-English translation outperforms other languages with the M2M-100 model, although Dutch beat Italian in METEOR score.
+Japanese received the worse score in both SacreBLEU and METEOR for this model.
+
+NLLB-200 scores way lower in both SacreBLEU and METEOR in most languages.
+It scores highest in Chinese-to-English translation in all metrics.
+However, Japanese-to-English translation suffers with a low score of 1.95 SacreBLEU and 0.09 METEOR
+
+OPUS-MT model shows best performance in all languages, the highest being Ukrainian-to-English translation with 75.44 SacreBLEU and 0.86 METEOR.
+The lowest scores are within Chinese-to-English translation: 59.50 SacreBLEU and Japanese-to-English: 0.78 METEOR.
+
+Consistency Across Models:
+Japanese translations consistently receive lower scores across all models.
+
+Dutch and Italian translations generally score higher across all metrics, suggesting that these languages are well-supported by the pre-trained models.
+
+Performance Variability:
+The NLLB-200 model shows significant variability, especially for Japanese, where the SacreBLEU and METEOR scores are extremely low.
+This indicates poor performance for this language pair with this particular model.
+The mBART50 model displays a wide range in SacreBLEU scores, with Spanish and Chinese having much lower scores compared to others, highlighting potential inconsistencies.
+
+Highest Performing Models:
+The OPUS-MT model consistently shows high scores across all languages, with the highest SacreBLEU and METEOR scores overall, especially for Ukrainian and Dutch.
+
+Based on the provided tables, the OPUS-MT model appears to be the most robust and reliable across different languages, with consistently high scores.
+Japanese translations, however, pose significant challenges for all models, as indicated by the consistently low scores across all metrics.
+The variability observed in NLLB-200, especially for Japanese, suggests that this model might need further tuning or additional data for certain language pairs.
+
 \begin{figure}[htbp]
     \centering
-    \includegraphics[width=1\linewidth]{figures/metrics_bar.png}
-    \caption{BLEU, SacreBLEU, and METEOR scores of each model on each language.}
-    \label{fig:result_visual}
+    \includegraphics[width=0.49\linewidth]{figures/chinese_all_metrics.png}
+    \includegraphics[width=0.49\linewidth]{figures/dutch_all_metrics.png}
+    \includegraphics[width=0.49\linewidth]{figures/finnish_all_metrics.png}
+    \includegraphics[width=0.49\linewidth]{figures/french_all_metrics.png}
+    \includegraphics[width=0.49\linewidth]{figures/german_all_metrics.png}
+    \includegraphics[width=0.49\linewidth]{figures/hebrew_all_metrics.png}
+    \includegraphics[width=0.49\linewidth]{figures/italian_all_metrics.png}
+    \includegraphics[width=0.49\linewidth]{figures/japanese_all_metrics.png}
+    \includegraphics[width=0.49\linewidth]{figures/polish_all_metrics.png}
+    \includegraphics[width=0.49\linewidth]{figures/russian_all_metrics.png}
+    \includegraphics[width=0.49\linewidth]{figures/spanish_all_metrics.png}
+    \includegraphics[width=0.49\linewidth]{figures/turkish_all_metrics.png}
+    \includegraphics[width=0.49\linewidth]{figures/ukrainian_all_metrics.png}
+    \caption{Performance of every language.}
+    \label{fig:lang_metrics}
 \end{figure}
 
-Figure \ref{fig:result_visual} shows the visualisation of all scores.
+Figure \ref{fig:lang_metrics} shows the SacreBLEU and METEOR metrics of each language.
+
+\begin{figure}[htbp]
+    \centering
+    \includegraphics[width=0.9\linewidth]{figures/inference_timings.png}
+    \caption{Inference timings for every model.}
+    \label{fig:inference_timings}
+\end{figure}
+
+Figure \ref{fig:inference_timings} shows the time to translate for each language across experimented models.
+
+It is quite unclear why mBART50 Spanish-to-English translation suffers from significantly higher timings. Re-running the experiment shows that this timing is consistent and not an anomaly.
 
 \section{Conclusion}
+
+Specific languages model perform better than multilingual models in the said languages.
+
+Limitations, only the first translation of the same phrases are taken.
 

### 2024-08-08 analyse/analyse_results.py

@@ -1,17 +1,10 @@
 import json
 import os
-from collections import Counter
 
 import matplotlib.pyplot as plt
-import nltk
 import numpy as np
 import pandas as pd
 import seaborn as sns
-import spacy
-from sentence_transformers import SentenceTransformer
-from sklearn.manifold import TSNE
-from textblob import TextBlob
-from wordcloud import WordCloud
 
 LANGS = [
     "chinese",
@@ -31,10 +24,13 @@ LANGS = [
 
 MODELS = [
     "opus",
-    "mbart-large-50-many-to-many-mmt",
-    "nllb-200-distilled-600M",
-    "m2m100_418M",
+    "mbart-50",
+    "nllb-200",
+    "m2m-100",
 ]
+
+BAR_COLOURS = ["tab:blue", "tab:orange", "tab:green", "tab:red"]
+
 
 with open("../inference/result/metrics.json", "r") as f:
     metrics = json.loads(f.read())
@@ -52,7 +48,7 @@ def plot_scores(dataframe, metric_name, ax, add_legend=False):
     ax.set_xticks(index + bar_width * (len(dataframe.columns) - 1) / 2)
     ax.set_xticklabels(dataframe.index, rotation=45, ha="right")
     if add_legend:
-        ax.legend(loc="upper right", fontsize="small")
+        ax.legend(loc="lower right", fontsize="small", framealpha=0.5)
 
 
 bleu_scores = pd.DataFrame(index=LANGS)
@@ -74,3 +70,84 @@ plot_scores(meteor_bleu_scores, "METEOR", ax3)
 
 plt.tight_layout()
 plt.savefig(f"./figures/metrics_bar.png")
+
+plt.clf()
+
+# Compute average scores for each model
+average_bleu_scores = bleu_scores.mean()
+average_sacre_bleu_scores = sacre_bleu_scores.mean()
+average_meteor_scores = meteor_bleu_scores.mean()
+
+# Create a DataFrame for average scores
+average_scores = pd.DataFrame(
+    {
+        "BLEU": average_bleu_scores,
+        "SacreBLEU": average_sacre_bleu_scores,
+        "METEOR": average_meteor_scores,
+    }
+)
+
+fig, (ax4, ax5, ax6) = plt.subplots(3, 1, figsize=(8, 10))
+
+bar_width = 0.8
+index = np.arange(len(average_scores.index))
+
[…]

### 2024-08-08 analyse/analyse_timings.py

@@ -1,0 +1,59 @@
+import json
+import os
+from collections import Counter
+
+import matplotlib.pyplot as plt
+import nltk
+import numpy as np
+import pandas as pd
+import seaborn as sns
+import spacy
+from sentence_transformers import SentenceTransformer
+from sklearn.manifold import TSNE
+from textblob import TextBlob
+from wordcloud import WordCloud
+
+MODELS = [
+    "opus",
+    "mbart-50",
+    "nllb-200",
+    "m2m-100",
+]
+
+opus_timings = pd.read_csv("../inference/result/opus_timings.csv", index_col=0)
+mbart50_timings = pd.read_csv(
+    "../inference/result/mbart-large-50-many-to-many-mmt_timings.csv", index_col=0
+)
+m2m100_timings = pd.read_csv("../inference/result/m2m100_418M_timings.csv", index_col=0)
+nllb200_timings = pd.read_csv(
+    "../inference/result/nllb-200-distilled-600M_timing.csv", index_col=0
+)
+
+
+# Combine the data into a single DataFrame
+combined_timings = pd.DataFrame(
+    {
+        "opus": opus_timings.iloc[0],
+        "mbart50": mbart50_timings.iloc[0],
+        "m2m100": m2m100_timings.iloc[0],
+        "nllb": nllb200_timings.iloc[0],
+    }
+)
+
+# Plot the differences in inference timing for every language
+plt.figure(figsize=(14, 8))
+
+languages = combined_timings.index
+
+for model in combined_timings.columns:
+    plt.plot(languages, combined_timings[model], marker="o", label=model)
+
+plt.xlabel("Language")
+plt.ylabel("Inference Time (seconds)")
+plt.title("Inference Timing Comparison for Different Models")
+plt.xticks(rotation=45, ha="right")
+plt.legend()
+# plt.grid(True)
+plt.tight_layout()
+
+plt.savefig("./figures/inference_timings.png")

### 2024-08-08 inference/opus_multilingual.py

@@ -1,0 +1,77 @@
+import time
+
+import pandas as pd
+import torch
+from tqdm import tqdm
+from transformers import MarianMTModel, MarianTokenizer
+
+MODEL_NAME = "Helsinki-NLP/opus-mt-mul-en"
+
+
+LANGS = {
+    "chinese": "zh",
+    "dutch": "nl",
+    "finnish": "fi",
+    "french": "fr",
+    "german": "de",
+    "hebrew": "he",
+    "italian": "it",
+    "japanese": "ja",
+    "polish": "pl",
+    "russian": "ru",
+    "spanish": "es",
+    "turkish": "tr",
+    "ukrainian": "uk",
+}
+
+tqdm.pandas()
+
+df = pd.read_csv("../dataset/sentences_v2.csv", index_col=0)
+
+
+def translate_batch(tokeniser, batch_input):
+    inputs = tokeniser(
+        batch_input,
+        return_tensors="pt",
+        padding=True,
+        truncation=True,
+    )
+    with torch.no_grad():
+        translated_tokens = model.generate(**inputs)
+    translations = tokeniser.batch_decode(translated_tokens, skip_special_tokens=True)
+
+    return translations
+
+
+batch_size = 4
+translated_df = df[["index", "english"]].copy()
+timing_df = pd.DataFrame()
+
+for src_lang, src_lang_code in LANGS.items():
+    print(f"Translating {src_lang}...")
+    start = time.time()
+
+    model = MarianMTModel.from_pretrained(MODEL_NAME)
+    tokeniser = MarianTokenizer.from_pretrained(
+        MODEL_NAME,
+        src_lang=src_lang_code,
+        tgt_lang="en",
+    )
+
+    translated_texts = []
+
+    for i in tqdm(range(0, len(df), batch_size)):
+        batch_input = df[src_lang][i : i + batch_size].tolist()
+        translations = translate_batch(tokeniser, batch_input)
+        translated_texts.extend(translations)
+
+    translated_df[f"translated_from_{src_lang}"] = translated_texts
+
+    end = time.time()
+    timing_df[f"translated_from_{src_lang}"] = [end - start]
+
+    translated_df.to_csv(f"result/opus.csv")
+    timing_df.to_csv(f"result/opus_timings.csv")
+
+
+# not entirely sure how to use the mt model

### 2024-08-08 paper/paper.tex
updated

@@
 mBART \cite{liu-2020-mbart}, M2M \cite{fan-2020-m2m100}, NLLB-200 \cite{nllb200-2020}, are examples of PTMs that support many-to-many translation language.
-This domain seems to be dominated by META AI.
+
+mBART, M2M-100, and NLLB-200 were all developed by Meta AI (formerly Facebook AI), showcasing the organisation's significant impact in the machine translation field through pre-trained models (PTMs).
+These models represent a substantial advancement in multilingual translation capabilities.
 
@@
 Languages are chosen based on its resources' availability in Tatoeba, as well as considering supported languages in most PTMs models.
+Thus, the languages chosen here can be considered as moderate to high resources languages.
+
 To build the dataset, sentences in English are first downloaded, containing 1,898,494 sentences (it is unclear why it is less than the number stated in the Tatoeba website).
@@
 The result is a single Dataframe containing 1,323 parallel sentences in all 14 languages, this will be treated as a test set to evaluate the models performance on each language.
+Note that in the case of multiple reference translations are available for the same sentence in English, only the first sentence is taken while the rest is discarded.
+Therefore, there is only one reference for each candidate sentence.
 
 Sentences typically consist of everyday phrases such as 'I have to go to sleep', 'That is intriguing', and 'Where do you live?'.
-They may also include single-word exclamations like 'Speak!' or 'Look!'.
+They may also include single-word exclamations like 'Speak!', 'So what?', or 'Look!'.
 Additionally, multiple sentences such as 'You may write in any language you want.
@@
 Therefore, machine translation process should be straightforward on this level.
+Figure \ref{table:parallel_sentence} shows two examples of parallel sentences in the final dataset, across all 14 languages and the original English sentence.
 
@@
 
-Figure \ref{table:parallel_sentence} shows two examples of parallel sentences in the final dataset, across 14 languages.
-
 \begin{figure}[htbp]
@@
     \includegraphics[width=1\linewidth]{figures/word_count_box.png}
-    \caption{Dataset word count distribution, consisting of mostly short sentences.}
+    \caption{Dataset word count distribution per language, Chinese and Japanese sentences are counted per letter.}
     \label{fig:word_count_box}
@@
 Figure \ref{fig:word_count_box} shows a box plot of sentences word count.
-It can be seen that the majority of sentences are short.
-Chinese and Japanese are counted per letter in this figure, which accounts for the variability seen in the Figure/
-
-Most languages cluster around 4-6 words per sentence. while Chinese cluster around 6-8 letters and Japanese cluster around 8-12 letters.
+The median for Dutch, Finnish, French, German, Hebrew, Italian, Polish, Russian, Spanish, Turkish, Ukrainian seems relatively low, with a moderate spread.
+A few to moderate count of outlier exist in all languages, suggesting occasional longer sentences, with the highest variability seen in Japanese.
+Note that for Chinese and Japanese the count is based on each Chinese/Japanese letter, which explains their higher interquartile range (IQR) and larger numbers of outliers.
 
@@
 The architecture is based on MARIAN-NMT \cite{mariannmt}, based on a standard transformer setup: 6 self-attentive layers in both the encoder and decoder networks, each with 8 attention heads per layer \cite{tiedemann-2020-opus-mt}.
-
 While OPUS-MT support both monolingual and multilingual PTMs, the multilingual models seem to lack proper documentation and therefore excluded from this experiment.
@@
 
-Figure \ref{fig:mbart} shows the architecture of mBART's key featuresL multilingual denoising pre-training and fine-tuning.
+Figure \ref{fig:mbart} shows the architecture of mBART's key features multilingual denoising pre-training and fine-tuning.
 It is shown that the model achieved consistent performance gains through pre-training in low-to-medium resource sentence level MT \cite{liu-2020-mbart}.
-"Denoising" here refers to a pre-training technique used to improve the model's ability to generate coherent and accurate translations by training it to reconstruct original sentences from corrupted (noisy) versions.
+The "denoising" here refers to a pre-training technique used to improve the model's ability to generate coherent and accurate translations by training it to reconstruct original sentences from corrupted (noisy) versions.
 
@@
 
-% (English-centric data (top left) includes only training data involving English, while our many-to-many multilingual setting (top right) encompasses direct translation between various languages, and our proposed M2M-100 model (bottom) integrates dense and sparse language-specific parameters for direct translation between languages.)
-
-% (Multilingual translation models factorize computation when translating to many
-% languages and share information between similar languages, which benefits low resource
-% directions (Arivazhagan et al., 2019)) \cite{arivazhagan-2019-massively-multilingual}
+Figure \ref{fig:m2m} show the inner working of the M2M model.
+The English-centric dataset (top left) includes training data exclusively involving translations to and from English, while the many-to-many multilingual setting (top right) involves direct translation data among multiple language pairs \cite{fan-2020-m2m100}.
+The M2M-100 model then integrates dense and sparse language-specific parameters to enable direct translation between languages (bottom part of Figure \ref{fig:m2m}).
 
@@
 \textbf{NLLB-200} \cite{nllb200-2020} is built to handle translation tasks across a broad spectrum of languages, including many that are low-resource or underrepresented in existing datasets.
-It supports translations for 200 languages, encompassing numerous underrepresented languages.
-mBART, M2M-100, and NLLB-200 were all developed by Meta AI (formerly Facebook AI), showcasing the organisation's significant impact in the machine translation field through pre-trained models (PTMs).
-These models represent a substantial advancement in multilingual translation capabilities.
+It supports translations for 200 languages, including numerous underrepresented languages, and is currently one of the most extensive multilingual machine translation models.
+The model employs a Mixture of Experts (MoE) architecture and achieves state-of-the-art (SoTA) results across many language pairs, even surpassing Meta's previous model, M2M-100 \cite{nllb200-2020}.
 
@@
     \centering
-    \includegraphics[width=0.9\linewidth]{images/nllb.png}
+    \includegraphics[width=0.7\linewidth]{images/nllb.png}
     \caption{NLLB pipeline \cite{nllb200-2020}}
@@
 
-% (No Language Left Behind: Our low-resource translation effort focuses on four
-% cornerstones. (1) We strive to understand the low-resource translation problem from the
-% perspective of native speakers. (2) We study how to automatically create training data to
-% move low-resource languages towards high-resource. (3) We utilize this data to create
-% state-of-the-art translation models. (4) We evaluate every language we aim to translate.)
+Figure \ref{fig:nllb} shows the Dense Transformer and MoE Transformer layers implemented within the model.
+MoE \cite{masoudnia-2012-moe} utilises a gating mechanism to route different inputs to different subsets of experts (sub-networks), allowing the model to handle various linguistic phenomena efficiently.
+Additionally, the model is then trained on a diverse set of languages, complemented with back-translation and data augmentation to generate additional data for low-resource languages \cite{nllb200-2020}.
+
+% The original paper is 192 pages long and impossible to be dissected in just one section.
 
@@
 OPUS-MT models are taken from the Helsinki-NLP repository.
-For mBART, the 'mbart-large-50-many-to-many-mmt' is used, 'm2m100\_418M', and 'nllb-200-distilled-600M'.
-All models are taken using the transformers \cite{wolf-etal-2020-transformers} library from HuggingFace \cite{huggingface}.
-
-Batch size of 4
-
-The translation are first generated and saved into a CSV file, before calculating the metrics score.
+The 'mbart-large-50-many-to-many-mmt' is used, 'm2m100\_418M', and 'nllb-200-distilled-600M'.
+All models are taken using the transformers \cite{wolf-etal-2020-transformers} library from Hugging Face \cite{huggingface}.
+Batch size of 4 is used in all models.
+Translations are first generated and saved into CSV files, before calculating the metrics score.
 
@@
 
-Other metrics for machine translation exist, but not used nearly as widely as BLEU.
-Crosslingual Optimized Metric for Evaluation of Translation (COMET) \cite{rei-2020-comet} is a recent metric that leverages neural-network-based models for deeper understanding and evaluation.
-BERTScore.
+In this project, a variant of BLEU \cite{papieni-2002-bleu} called SacreBLEU \cite{post-2018-sacrebleu} and METEOR \cite{lavie-2007-meteor} will be used in this experiment due to their popularity, simplicity, and ease-of-use.
 
-In this project, a variant of BLEU called SacreBLEU and METEOR will be used due to their simplicity.
+While other metrics such as Crosslingual Optimized Metric for Evaluation of Translation (COMET) \cite{rei-2020-comet} and BERTSCore \cite{zhang-2020-bertscore} exist, they involve using a deep neural network or a transformer to evaluate, which naturally increase computational cost and slow down calculation greatly.
 
@@
 
-Furthermore, It is challenging to directly compare BLEU scores between paper \cite{post-2018-clarity-bleu}.
-Thus, the author proposed a variation called SacreBLEU \cite{post-2018-clarity-bleu}, which claimed to facilitate easy computation of BLEU scores that are shareable, comparable, and reproducible.
+Furthermore, It is challenging to directly compare BLEU scores between paper \cite{post-2018-sacrebleu}.
+Thus, the author proposed a variation called SacreBLEU \cite{post-2018-sacrebleu}, which claimed to facilitate easy computation of BLEU scores that are shareable, comparable, and reproducible.
 This is done by including preprocessing steps and normalisation routines.
@@
 
-\begin{figure}[htbp]
-    \centering
-    \includegraphics[width=1\linewidth]{figures/metrics_bar.png}
-    \caption{BLEU, SacreBLEU, and METEOR scores of each model on each language.}
-    \label{fig:result_visual}
-\end{figure}
-
-Figure \ref{fig:result_visual} shows the visualisation of all scores.
-
-OPUS-MT models outperform other models at every language on SacreBLEU and METEOR scores.
-
-mBart50 performed second best, being the best-performing multilingual models in this study.
-
-NLLB-200 performed the worse, as the model is more concerned with translations on low-resource languages.
-
 \begin{table}[htbp]
     \centering
+    \begin{minipage}{0.49\linewidth}
+        \footnotesize
+        \begin{tabular}{|l|r|r|r|}
+            \hline
+            \textbf{Language} & \textbf{BLEU}   & \textbf{SacreBLEU} & \textbf{METEOR} \\
+            \hline
+            Chinese           & 0.0971          & \textit{59.5098}   & 0.7953          \\
+            Dutch             & 0.0974          & 69.8803            & 0.8471          \\
+            Finnish           & 0.0965          & 66.6267            & 0.8296          \\
+            French            & 0.0967          & 69.8185            & 0.8357          \\
+            German            & 0.0969          & 69.7422            & 0.8419          \\
+            Hebrew            & 0.0977          & 66.5149            & 0.8229          \\
+            Italian           & 0.0979          & 74.1298            & 0.8584          \\
+            Japanese          & 0.0960          & 63.1435            & \textit{0.7893} \\
+            Polish            & \textit{0.0952} & 61.9026            & 0.8425          \\
+            Russian           & 0.0974          & 66.7046            & 0.8179          \\
+            Spanish           & 0.0979          & 71.4174            & 0.8463          \\
+            Turkish           & \textbf{0.0980} & 72.6551            & 0.8460          \\
+            Ukrainian         & 0.0978          & \textbf{75.4447}   & \textbf{0.8667} \\
+            \hline
+        \end{tabular}
+        \caption{OPUS-MT result.}
+        \label{table:opus_result}
+    \end{minipage}
     \begin{minipage}{0.49\linewidth}
@@
     \end{minipage}
+\end{table}
+
+\begin{table}[htbp]
+    \centering
     \begin{minipage}{0.49\linewidth}
@@
     \end{minipage}
-\end{table}
-
-\begin{table}[htbp]
-    \centering
     \begin{minipage}{0.49\linewidth}
@@
     \end{minipage}
-    \begin{minipage}{0.49\linewidth}
-        \footnotesize
-        \begin{tabular}{|l|r|r|r|}
-            \hline
-            \textbf{Language} & \textbf{BLEU}   & \textbf{SacreBLEU} & \textbf{METEOR} \\
-            \hline
-            Chinese           & 0.0971          & \textit{59.5098}   & 0.7953          \\
-            Dutch             & 0.0974          & 69.8803            & 0.8471          \\
-            Finnish           & 0.0965          & 66.6267            & 0.8296          \\
-            French            & 0.0967          & 69.8185            & 0.8357          \\
-            German            & 0.0969          & 69.7422            & 0.8419          \\
-            Hebrew            & 0.0977          & 66.5149            & 0.8229          \\
-            Italian           & 0.0979          & 74.1298            & 0.8584          \\
-            Japanese          & 0.0960          & 63.1435            & \textit{0.7893} \\
-            Polish            & \textit{0.0952} & 61.9026            & 0.8425          \\
-            Russian           & 0.0974          & 66.7046            & 0.8179          \\
-            Spanish           & 0.0979          & 71.4174            & 0.8463          \\
-            Turkish           & \textbf{0.0980} & 72.6551            & 0.8460          \\
-            Ukrainian         & 0.0978          & \textbf{75.4447}   & \textbf{0.8667} \\
-            \hline
-        \end{tabular}
-        \caption{OPUS-MT result.}
-        \label{table:opus_result}
-    \end{minipage}
 \end{table}
@@
 Table \ref{table:mbart_result}, \ref{table:m2m_result}, \ref{table:nllb_result}, and \ref{table:opus_result} shows the result from all four PTMs.
-
-The standard BLEU score is very low in all cases due to the sentences being short.
-
-The mBART50 model achieved best SacreBLEU and METEOR score in Italian-to-English translation, while suffers the most in Spanish-to-English (SacreBLEU) and Turkish-to-English (METEOR) translation.
-
-Simialrly, SacreBLEU for Italian-to-English translation outperforms other languages with the M2M-100 model, although Dutch beat Italian in METEOR score.
+The BLEU score is very low in all cases due to the sentences being short (find cite).
+The mBART-50 model achieved best SacreBLEU and METEOR score in Italian-to-English translation, while suffers the most in Spanish-to-English (SacreBLEU) and Turkish-to-English (METEOR) translation.
+In addition, it slightly outperforms OPUS-MT in Polish-to-English translation on SacreBLEU score.
+Similarly, SacreBLEU for Italian-to-English translation outperforms other languages with the M2M-100 model, although Dutch beat Italian in METEOR score.
 Japanese received the worse score in both SacreBLEU and METEOR for this model.
-
-NLLB-200 scores way lower in both SacreBLEU and METEOR in most languages.
+NLLB-200 scores much lower in both SacreBLEU and METEOR in most languages.
 It scores highest in Chinese-to-English translation in all metrics.
-However, Japanese-to-English translation suffers with a low score of 1.95 SacreBLEU and 0.09 METEOR
-
+However, Japanese-to-English translation suffers with a low score of 1.95 SacreBLEU and 0.09 METEOR.
 OPUS-MT model shows best performance in all languages, the highest being Ukrainian-to-English translation with 75.44 SacreBLEU and 0.86 METEOR.
@@
 
-Consistency Across Models:
-Japanese translations consistently receive lower scores across all models.
+\begin{figure}[htbp]
+    \centering
+    \includegraphics[width=1\linewidth]{figures/metrics_bar.png}
+    \caption{BLEU, SacreBLEU, and METEOR scores of each model on each language.}
+    \label{fig:result_visual}
+\end{figure}
 
-Dutch and Italian translations generally score higher across all metrics, suggesting that these languages are well-supported by the pre-trained models.
+Figure \ref{fig:result_visual} shows the visualisation of all scores.
 
-Performance Variability:
+All models perform similarly on BLEU score across all languages, with NLLB-200 suffering a bit more on Japanese translation.
+The OPUS-MT model consistently shows high scores across all languages, with the highest SacreBLEU and METEOR scores overall, especially for Ukrainian and Italian.
+
+The OPUS-MT model Generally performs consistently across most languages but slightly lower compared to other models in a few cases.
+The mBART-50 model achieved the best-performing multilingual model, showing competitive performance, often achieving the second-highest scores across several languages.
+M2M-100 performs worse in SacreBLEU compared to mBART-50, but comparably similar in METEOR scores.
 The NLLB-200 model shows significant variability, especially for Japanese, where the SacreBLEU and METEOR scores are extremely low.
-This indicates poor performance for this language pair with this particular model.
-The mBART50 model displays a wide range in SacreBLEU scores, with Spanish and Chinese having much lower scores compared to others, highlighting potential inconsistencies.
 
-Highest Performing Models:
-The OPUS-MT model consistently shows high scores across all languages, with the highest SacreBLEU and METEOR scores overall, especially for Ukrainian and Dutch.
-
-Based on the provided tables, the OPUS-MT model appears to be the most robust and reliable across different languages, with consistently high scores.
-Japanese translations, however, pose significant challenges for all models, as indicated by the consistently low scores across all metrics.
-The variability observed in NLLB-200, especially for Japanese, suggests that this model might need further tuning or additional data for certain language pairs.
+Dutch and Italian translations generally score higher across all metrics, suggesting that these languages are well-supported by the pre-trained models, while Japanese translations consistently receive lower scores across all models.
 
@@
 
-Figure \ref{fig:lang_metrics} shows the SacreBLEU and METEOR metrics of each language.
+Figure \ref{fig:lang_metrics} shows the SacreBLEU and METEOR score of each language by models.
+The OPUS-MT and mBART-50 models are the clear winners in almost all languages.
+Although M2M-100 performs competitively to mBART-50 in most languages and beating mBART-50 in Spanish-to-English translation.
+Again, NLLB-200 performs the weakest, with extremely low scores on German, Hebrew, and Japanese.
 
@@
 
-Figure \ref{fig:inference_timings} shows the time to translate for each language across experimented models.
-
-It is quite unclear why mBART50 Spanish-to-English translation suffers from significantly higher timings. Re-running the experiment shows that this timing is consistent and not an anomaly.
+Figure \ref{fig:inference_timings} shows the time taken to translate for each language across experimented models.
+The OPUS-MT model is the fastest among the other three multilingual models, which perform similarly in timings.
+It is quite unclear why mBART50 Spanish-to-English translation suffers from a significantly higher timing.
+Re-running the experiment shows that this timing is consistent and not an anomaly.
 
@@
 
-% READ THIS \cite{intento-2020} (exactly what this paper should do, maybe compare results and find insights)
+We show that one-to-one PTMs outperform multilingual PTMs on sentence translation.
+
+mBART50 performed better than both M2M-100 and NLLB-200 in these 14 languages.
+
+the OPUS-MT model appears to be the most robust and reliable across different languages, with consistently high scores.
+Japanese translations, however, pose significant challenges for all models, as indicated by the consistently low scores across all metrics.
+The variability observed in NLLB-200, especially for Japanese, suggests that this model might need further tuning or additional data for certain language pairs.
 

### 2024-08-08 paper/paper.tex
updated

@@
 
+\begin{abstract}
+    Machine translation (MT) has become an essential tool for overcoming language barriers in an increasingly globalised world.
+Today, there are pre-trained models (PTMs) specifically designed for automatic machine translation.
+This paper evaluates the performance of various pre-trained models (PTMs) for many-to-English translation.
+The models assessed include OPUS-MT \cite{tiedemann-2020-opus-mt}, mBART-50 \cite{liu-2020-mbart}, NLLB-200 \cite{nllb200-2020}, and M2M-100 \cite{fan-2020-m2m100}.
+Using a dataset curated from the Tatoeba repository, comprising 1,323 parallel sentence pairs across 14 source languages, translation quality is measured using BLEU, SacreBLEU, and METEOR scores.
+The results indicate that OPUS-MT, a one-to-one PTM, outperforms multilingual models in one-to-English translation tasks, with mBART-50 leading among the multilingual PTMs.
+These findings highlight the strengths and limitations of both specific language and multilingual PTMs in machine translation.
+\end{abstract}
+
 \section{Introduction}
@@
 The following years saw the birth of Large Language Models (LLMs) such as BERT \cite{devlin-2019-bert}, GPT \cite{openai-2024-gpt4}, and T5 \cite{raffel-2023-t5}, which revolutionised both MT and the whole field of NLP.
-Then in early 2020s, several pre-trained models (PTM) that are specifically designed for machine translation emerged, namely mBART \cite{liu-2020-mbart}, mT5 \cite{xue-2021-mt5}, NLLB \cite{nllb200-2020}, M2M \cite{fan-2020-m2m100}, and PolyLM \cite{wei-2023-polylm}.
-Most of these models are multilingually trained, allowing for many-to-many translation: able to translate between any of the supported pair of languages.
+Then in early 2020s, several pre-trained models (PTM) that are specifically designed for machine translation emerged, OPUS-MT \cite{tiedemann-2020-opus-mt} produced numerous PTMs designed for one-to-one translations.
+Additionally, multilingual models: mBART-50 \cite{liu-2020-mbart}, NLLB-200 \cite{nllb200-2020}, and M2M-100 \cite{fan-2020-m2m100} have also been published, growing popularity as multilinguality has been shown to allow for sharing information among languages \cite{garcia-2020-multilingual}.
+Most of these models are trained on diverse set of languages, allowing for many-to-many translation: able to translate between any of the supported pair of languages.
 This allows the models to generalise over shared lexical and linguistic among languages, and have been shown to increase performance compared to one-to-one translation models \cite{liu-2020-mbart}.
-
-Intento published 'The State of Machine Translation 2024' \cite{intento-2020} providing an in-depth evaluation of popular MT engines and LLMs.
-However, the biggest drawback in this report is that the selected LLMs are general LLMs such as GPT, LLaMa, Mistral, instead of MT-specific LLMs.
 
@@
 Nevertheless, there does not seem to be much work on comparing or benchmarking different pre-trained models in machine translation.
+Intento published 'The State of Machine Translation 2024' \cite{intento-2020} providing an in-depth evaluation of popular MT engines and LLMs.
+However, the biggest drawback in this report is that the selected LLMs are general LLMs such as GPT, LLaMa, Mistral, instead of MT-specific LLMs.
 
@@
 A dataset is curated from the Tatoeba repository \cite{tatoeba}, containing 1,323 parallel sentence pairs across source and target languages.
-The models includes a one-to-one PTMs: OPUS-MT \cite{tiedemann-2023-democratizing,tiedemann-2020-opus-mt}, and multilingual PTMs: such as mBART-50 \cite{liu-2020-mbart}, NLLB-200 \cite{nllb200-2020}, and M2M-100 \cite{fan-2020-m2m100}.
-The performance of these models is evaluated using the BLEU score \cite{papieni-2002-bleu}.
+The models include a one-to-one PTMs: OPUS-MT \cite{tiedemann-2023-democratizing,tiedemann-2020-opus-mt}, and multilingual PTMs: such as mBART-50 \cite{liu-2020-mbart}, NLLB-200 \cite{nllb200-2020}, and M2M-100 \cite{fan-2020-m2m100}.
+The performance of these models is evaluated using the BLEU \cite{papieni-2002-bleu}, SacreBLEU \cite{post-2018-sacrebleu}, and METEOR \cite{lavie-2007-meteor}.
 
-% (Multilinguality has been explored in the supervised NMT literature, where it has been shown to enable information sharing among related languages.) \cite{garcia-2020-multilingual}
-% (NMT systems have lower translation quality on very long sentences) \cite{koehn-2017-challenges}
-
-% Language barriers
+The result indicates that OPUS-MT outperforms all other multilingual models in one-to-English translation, highlighting the superiority of specific languages PTMs compared to multilingual PTMs.
+mBART-50 shows the best performance among other multilingual PTMs: M2M-100 and NLLB-200.
 
@@
 The encoder-decoder approach \cite{cho-2014-properties} remains as the foundation architecture for many sequence-to-sequence models in machine translation.
+Attention mechanism has been used for Neural Machine Translation (NMT) \cite{bahdanau-2016-nmt-jointly, luong-2015-effective-attention} before the release of the transformer \cite{vaswani-2017-attention}
 
-In terms of pre-trained models (PTMs) for machine translation, most PTMs support two kinds of translations: one-to-one translation and many-to-many translation (multilingual models).
+Currently, Pre-Trained-Models (PTMs) have become a cornerstone in machine translation, leveraging large datasets and powerful architectures to achieve state-of-the-art performance \cite{han-2021-ptms}.
+Most PTMs support two kinds of translations: one-to-one translation and many-to-many translation (multilingual models).
 
@@
 OPUS-MT \cite{tiedemann-2020-opus-mt} and Facebook Fairseq \cite{ott-2019-fairseq} are examples of PTMs that support one-to-one translation.
+Multilingual models have also gained significant prominence, enabling many-to-many translation, where a single model can translate between multiple source and target languages \cite{aharoni-2019-massively-multilingual}.
 
-With the advent of large language models and pre-trained language models, multilingual machine translation has gained prominence.
-This approach enables many-to-many translation, where a single model can translate between multiple source and target languages \cite{aharoni-2019-massively-multilingual}.
-
-mBART \cite{liu-2020-mbart}, M2M \cite{fan-2020-m2m100}, NLLB-200 \cite{nllb200-2020}, are examples of PTMs that support many-to-many translation language.
-
-mBART, M2M-100, and NLLB-200 were all developed by Meta AI (formerly Facebook AI), showcasing the organisation's significant impact in the machine translation field through pre-trained models (PTMs).
-These models represent a substantial advancement in multilingual translation capabilities.
-
-% Back translation?
+OPUS-MT \cite{tiedemann-2020-opus-mt} is a project dedicated to creating and providing free resources for machine translation,providing over a thousand of PTMs of direct one-to-one translations.
+In contrast, mBART \cite{liu-2020-mbart}, M2M \cite{fan-2020-m2m100}, NLLB-200 \cite{nllb200-2020}, are massively multilingual models that support many-to-many translation language.
 
@@
 Therefore, they do not support easy evaluation for many-to-English translations.
-
-Furthermore, papers often use different sets and different versions of datasets to evaluate their model performance, incurring unstraightforward comparison.
+Furthermore, papers often use different datasets and versions to evaluate their model performance, making direct comparison challenging.
 
@@
 
-\subsection{Inference Details}
-
-OPUS-MT models are taken from the Helsinki-NLP repository.
-The 'mbart-large-50-many-to-many-mmt' is used, 'm2m100\_418M', and 'nllb-200-distilled-600M'.
-All models are taken using the transformers \cite{wolf-etal-2020-transformers} library from Hugging Face \cite{huggingface}.
-Batch size of 4 is used in all models.
-Translations are first generated and saved into CSV files, before calculating the metrics score.
-
 \subsection{Evaluation Metrics}
@@
 Furthermore, It is challenging to directly compare BLEU scores between paper \cite{post-2018-sacrebleu}.
-Thus, the author proposed a variation called SacreBLEU \cite{post-2018-sacrebleu}, which claimed to facilitate easy computation of BLEU scores that are shareable, comparable, and reproducible.
-This is done by including preprocessing steps and normalisation routines.
+Thus, the author proposed a standardised variant called SacreBLEU \cite{post-2018-sacrebleu}, designed to ensure easy, consistent, and comparable evaluation across different implementations and research studies.
 
@@
 
+\subsection{Inference Details}
+
+OPUS-MT models are taken from the Helsinki-NLP repository.
+The 'mbart-large-50-many-to-many-mmt' is used, 'm2m100\_418M', and 'nllb-200-distilled-600M'.
+All models are taken using the transformers \cite{wolf-etal-2020-transformers} library from Hugging Face \cite{huggingface}.
+Batch size of 4 is used in all models.
+Translations are first generated and saved into CSV files, before calculating the metrics score.
+
+The sacrebleu packages from pip is used to calculate SacreBLEU \cite{post-2018-sacrebleu}.
+For BLEU and METEOR, the NLTK \cite{bird-2009-natural} package is used.
+Corpus BLEU is implemented for both SacreBLEU and BLEU in the calculations.
 \section{Evaluation}
@@
 All models perform similarly on BLEU score across all languages, with NLLB-200 suffering a bit more on Japanese translation.
-The OPUS-MT model consistently shows high scores across all languages, with the highest SacreBLEU and METEOR scores overall, especially for Ukrainian and Italian.
-
-The OPUS-MT model Generally performs consistently across most languages but slightly lower compared to other models in a few cases.
+The OPUS-MT model consistently shows high scores across most languages, with the highest SacreBLEU and METEOR scores overall, especially for Ukrainian and Italian.
 The mBART-50 model achieved the best-performing multilingual model, showing competitive performance, often achieving the second-highest scores across several languages.
@@
 
-Dutch and Italian translations generally score higher across all metrics, suggesting that these languages are well-supported by the pre-trained models, while Japanese translations consistently receive lower scores across all models.
+Dutch and Italian translations generally score higher across all metrics, suggesting that these languages are well-supported by the pre-trained models.
+Japanese translations, however, pose significant challenges for all models, as indicated by the consistently low scores across all metrics.
 
@@
 
-Figure \ref{fig:lang_metrics} shows the SacreBLEU and METEOR score of each language by models.
+Figure \ref{fig:lang_metrics} shows the SacreBLEU and METEOR scores for each language by model.
 The OPUS-MT and mBART-50 models are the clear winners in almost all languages.
-Although M2M-100 performs competitively to mBART-50 in most languages and beating mBART-50 in Spanish-to-English translation.
-Again, NLLB-200 performs the weakest, with extremely low scores on German, Hebrew, and Japanese.
+Although M2M-100 performs competitively with mBART-50 in most languages, it surpasses mBART-50 in Spanish-to-English translation.
+On the other hand, NLLB-200 performs the weakest, with extremely low scores in German, Hebrew, and Japanese.
 
@@
 
-Specific languages model perform better than multilingual models in the said languages.
+This study findings show that while multilingual models have been shown to achieve higher performance compared to one-to-one translation models \cite{liu-2020-mbart}, this does not hold on short sentences, many-to-English translations across the 14 source languages.
+Additionally, mBART50 performed better than PTMs M2M-100 and NLLB-200 in these 14 languages.
 
 Limitations, only the first translation of the same phrases are taken.
+No fine-tuning is applied, which may affect the translation results, as fine-tuning models for specific languages can significantly improve models performance \cite{zhang-2023-fine-tuning}.
 
-We show that one-to-one PTMs outperform multilingual PTMs on sentence translation.
+\section{Future Works}
 
-mBART50 performed better than both M2M-100 and NLLB-200 in these 14 languages.
+Future works should be done by incorporating more Pre-Trained Models (PTMs) and incorporating fine-tuning for specific languages on languages-to-English translation.
+Evaluating with bigger datasets and longer texts can also be beneficial, as well as using language other than English as the target language.
 
-the OPUS-MT model appears to be the most robust and reliable across different languages, with consistently high scores.
-Japanese translations, however, pose significant challenges for all models, as indicated by the consistently low scores across all metrics.
-The variability observed in NLLB-200, especially for Japanese, suggests that this model might need further tuning or additional data for certain language pairs.
+Furthermore, it may be beneficial to consider more general multilingual PTMs such as mBERT \cite{wu-2020-mbert-are-all} and PolyLM \cite{wei-2023-polylm}, which are trained for a wide arrange of NLP tasks instead of just machine translation.
 

### 2024-08-14 paper/paper.tex
updated paper

@@
 
-Despite these advancements, pre-trained models are often evaluated using different set of benchmarks \cite{liu-2020-mbart,nllb200-2020,fan-2020-m2m100,wei-2023-polylm}, making it difficult to gauge their relative effectiveness across various languages.
+Despite these advancements, pre-trained models are often evaluated using different set of languages and benchmarks \cite{liu-2020-mbart,nllb200-2020,fan-2020-m2m100,wei-2023-polylm}, making it difficult to gauge their relative effectiveness across various languages.
 Nevertheless, there does not seem to be much work on comparing or benchmarking different pre-trained models in machine translation.
@@
     \centering
+    \large
     \begin{tabular}{|l|l|r|}
         \hline
-        \textbf{No.} & \textbf{Language} & \textbf{Sentence Pairs Count} \\
+        \textbf{No.} & \textbf{Language Pairs} & \textbf{Sentence Pairs Count} \\
         \hline
-        1            & Chinese           & 68,814                        \\
-        2            & Dutch             & 155,856                       \\
-        3            & Finnish           & 102,202                       \\
-        4            & French            & 405,088                       \\
-        5            & German            & 501,145                       \\
-        6            & Hebrew            & 172,082                       \\
-        7            & Hungarian         & 171,698                       \\
-        8            & Italian           & 624,160                       \\
-        9            & Japanese          & 270,116                       \\
-        10           & Polish            & 77,345                        \\
-        11           & Russian           & 722,837                       \\
-        12           & Spanish           & 265,253                       \\
-        13           & Turkish           & 710,279                       \\
-        14           & Ukrainian         & 214,244                       \\
+        1            & Chinese-English         & 68,814                        \\
+        2            & Dutch-English           & 155,856                       \\
+        3            & Finnish-English         & 102,202                       \\
+        4            & French-English          & 405,088                       \\
+        5            & German-English          & 501,145                       \\
+        6            & Hebrew-English          & 172,082                       \\
+        7            & Hungarian-English       & 171,698                       \\
+        8            & Italian-English         & 624,160                       \\
+        9            & Japanese-English        & 270,116                       \\
+        10           & Polish-English          & 77,345                        \\
+        11           & Russian-English         & 722,837                       \\
+        12           & Spanish-English         & 265,253                       \\
+        13           & Turkish-English         & 710,279                       \\
+        14           & Ukrainian-English       & 214,244                       \\
         \hline
@@
 The architecture is based on MARIAN-NMT \cite{mariannmt}, based on a standard transformer setup: 6 self-attentive layers in both the encoder and decoder networks, each with 8 attention heads per layer \cite{tiedemann-2020-opus-mt}.
-While OPUS-MT support both monolingual and multilingual PTMs, the multilingual models seem to lack proper documentation and therefore excluded from this experiment.
+
+While they provide both monolingual and multilingual PTMs, only the OPUS-MT monolingual models are used in this study.
 
@@
 
-Figure \ref{fig:mbart} shows the architecture of mBART's key features multilingual denoising pre-training and fine-tuning.
-It is shown that the model achieved consistent performance gains through pre-training in low-to-medium resource sentence level MT \cite{liu-2020-mbart}.
-The "denoising" here refers to a pre-training technique used to improve the model's ability to generate coherent and accurate translations by training it to reconstruct original sentences from corrupted (noisy) versions.
+The mBART model is built as a typical sequence-to-sequence transformer architecture \cite{vaswani-2017-attention}, using an embedding dimension of 1024 and 16 heads within 12 layers of encoder and 12 layers of decoder, totalling to around 680M parameters \cite{liu-2020-mbart}.
+
+Figure \ref{fig:mbart} shows  mBART's key features: multilingual denoising pre-training and fine-tuning.
+The dataset is first corrupted by a noising function, which is applied to 35\% of words of each span text \cite{liu-2020-mbart}.
+During pre-training, the model will be made to denoise the texts, improving the model's ability to generate coherent and accurate translations.
+After the pre-training step, the model is then able to be fine-tuned for downstream tasks.
+It is shown that the model achieved consistent performance gains through pre-training in low-to-medium resource sentence level MT \cite{liu-2020-mbart}
 
@@
 
-\textbf{M2M-100} \cite{fan-2020-m2m100} is designed to perform direct translation between 100 languages without relying on English as an intermediate language.
+\textbf{M2M-100} \cite{fan-2020-m2m100} is designed to perform direct translation between any pair of 100 languages without relying on English as an intermediate language.
+Similarly, the architecture is of a transformer sequence-to-sequence with a typical encoder that outputs embeddings of the same length, while the decoder is an autoregressive component that sequentially produce the target sentence one token at a time.
+The parameters include 12 encoder layers and 12 decoder layers, 8192 FFN size and 1024 embedding dimension, totalling to a 1.2B parameter count \cite{fan-2020-m2m100}.
 
@@
 The English-centric dataset (top left) includes training data exclusively involving translations to and from English, while the many-to-many multilingual setting (top right) involves direct translation data among multiple language pairs \cite{fan-2020-m2m100}.
-The M2M-100 model then integrates dense and sparse language-specific parameters to enable direct translation between languages (bottom part of Figure \ref{fig:m2m}).
+Finally, the model combines sparse language-specific parameters with dense parameters (bottom part of Figure \ref{fig:m2m}),
 
@@
 The model employs a Mixture of Experts (MoE) architecture and achieves state-of-the-art (SoTA) results across many language pairs, even surpassing Meta's previous model, M2M-100 \cite{nllb200-2020}.
+
+The 'nllb-200-distilled-600M' variant has a total of 600M parameters, distilled from originally 54B parameters \cite{nllb200-2020}
 
@@
 
+% the most assured way to determine the best MT system from competing alternatives is to have an assessment done by competent human translators.
+
 In this project, a variant of BLEU \cite{papieni-2002-bleu} called SacreBLEU \cite{post-2018-sacrebleu} and METEOR \cite{lavie-2007-meteor} will be used in this experiment due to their popularity, simplicity, and ease-of-use.
@@
 
-The n-gram precision, as presented in the original BLEU paper \cite{papieni-2002-bleu}, is calculated as:
-
-\begin{equation}
-    p_n = \frac{\sum_{C \in \{Candidates\}} \sum_{\text{n-gram} \in C} \text{Count}_{\text{clip}}(\text{n-gram})}{\sum_{C \in \{Candidates\}} \sum_{\text{n-gram} \in C} \text{Count}(\text{n-gram})}
-\end{equation}
-
-Where:
-\begin{itemize}
-    \item \( p_n \) is the precision for n-grams.
-    \item \( \sum_{C \in \{Candidates\}} \) denotes the summation over all candidate translations.
-    \item \( \sum_{\text{n-gram} \in C} \) denotes the summation over all n-grams in a candidate translation \( C \).
-    \item \( \text{Count}_{\text{clip}}(\text{n-gram}) \) is the clipped count of the n-gram, which is the count of the n-gram in the candidate translation limited by the maximum count of that n-gram in any reference translation.
-    \item \( \text{Count}(\text{n-gram}) \) is the count of the n-gram in the candidate translation.
-\end{itemize}
-
-Thus, the BLEU score is calculated as:
-
+The BLEU score is calculated as:
 \begin{equation}
@@
 \end{equation}
-
-Where:
-\begin{itemize}
-    \item \( BP \) is the brevity penalty.
-    \item \( p_n \) is the precision for n-grams.
-    \item \( w_n \) is the weight for each n-gram (often uniformly distributed, so \( w_n = \frac{1}{N} \)).
-\end{itemize}
+where \( BP \) is the brevity penalty, \( p_n \) is the precision for n-grams, and \( w_n \) is the weight for each n-gram (often uniformly distributed, so \( w_n = \frac{1}{N} \)).
 
 The brevity penalty (BP) is calculated as:
-
 \begin{equation}
@@
 \end{equation}
+where \( c \) is the length of the candidate translation and \( r \) is the effective reference length.
 
-Where:
-\begin{itemize}
-    \item \( c \) is the length of the candidate translation.
-    \item \( r \) is the effective reference length.
-\end{itemize}
+The n-gram precision, as presented in the original BLEU paper \cite{papieni-2002-bleu}, is calculated as:
+\begin{equation}
+    p_n = \frac{\sum_{C \in \{Candidates\}} \sum_{\text{n-gram} \in C} \text{Count}_{\text{clip}}(\text{n-gram})}{\sum_{C \in \{Candidates\}} \sum_{\text{n-gram} \in C} \text{Count}(\text{n-gram})}
+\end{equation}
+where \( p_n \) is the precision for n-grams, \( \sum_{C \in \{Candidates\}} \) denotes the summation over all candidate translations, \( \sum_{\text{n-gram} \in C} \) denotes the summation over all n-grams in a candidate translation \( C \).
+\( \text{Count}_{\text{clip}}(\text{n-gram}) \) is the clipped count of the n-gram, which is the count of the n-gram in the candidate translation limited by the maximum count of that n-gram in any reference translation.
+\( \text{Count}(\text{n-gram}) \) is the count of the n-gram in the candidate translation.
 
@@
 It is designed to address some limitations of the BLEU score, allowing matches between simple morphological variants and synonyms.
-
+The formula is defined as:
 \begin{equation}
@@
 \end{equation}
-
-Where:
-\begin{itemize}
-    \item \(P\) is the precision,
-    \item \(R\) is the recall,
-    \item \(\text{frag}\) is the fragmentation penalty,
-    \item \(\gamma\) is a parameter that controls the weight of the fragmentation penalty, commonly 0.5,
-    \item \(\alpha\) is a parameter that controls the balance between precision and recall, commonly 0.9,
-\end{itemize}
+where \(P\) is the precision, \(R\) is the recall, and \(\text{frag}\) is the fragmentation penalty.
+\(\gamma\) is a parameter that controls the weight of the fragmentation penalty, commonly set to 0.5, and \(\alpha\) is a parameter that controls the balance between precision and recall, commonly set to 0.9,
 
@@
 Corpus BLEU is implemented for both SacreBLEU and BLEU in the calculations.
+Additionally, smoothing method 7 is used for BLEU, which interpolates two other smoothing functions: smoothing method 4, which divides by \( \frac{1}{\ln(\text{len}(T))} \), where \( T \) is the length of the translation; and smoothing method 5, which averages the \( n-1 \), \( n \), and \( n+1 \) gram matched counts \cite{nltk-bleu-docs}.
+
+All codes are written in Python 3.12.2, run locally on MacOS Sonoma 14.4.1 with M2 chip and 16 GB memory. GitHub: \url{https://github.com/stefanliemawan/eval-machine-translation}
+
 \section{Evaluation}
@@
 
-Table \ref{table:mbart_result}, \ref{table:m2m_result}, \ref{table:nllb_result}, and \ref{table:opus_result} shows the result from all four PTMs.
-The BLEU score is very low in all cases due to the sentences being short (find cite).
-The mBART-50 model achieved best SacreBLEU and METEOR score in Italian-to-English translation, while suffers the most in Spanish-to-English (SacreBLEU) and Turkish-to-English (METEOR) translation.
-In addition, it slightly outperforms OPUS-MT in Polish-to-English translation on SacreBLEU score.
-Similarly, SacreBLEU for Italian-to-English translation outperforms other languages with the M2M-100 model, although Dutch beat Italian in METEOR score.
-Japanese received the worse score in both SacreBLEU and METEOR for this model.
-NLLB-200 scores much lower in both SacreBLEU and METEOR in most languages.
-It scores highest in Chinese-to-English translation in all metrics.
-However, Japanese-to-English translation suffers with a low score of 1.95 SacreBLEU and 0.09 METEOR.
-OPUS-MT model shows best performance in all languages, the highest being Ukrainian-to-English translation with 75.44 SacreBLEU and 0.86 METEOR.
-The lowest scores are within Chinese-to-English translation: 59.50 SacreBLEU and Japanese-to-English: 0.78 METEOR.
+Table \ref{table:opus_result}, \ref{table:mbart_result}, \ref{table:m2m_result}, and \ref{table:nllb_result}, and  shows the result from all four PTMs.
+All models perform similarly low on BLEU score across all languages, with NLLB-200 reaching 0.01 below more on Japanese translation.
+The BLEU scores are very low (below 10 BLEU) in all cases due to the sentences being short (find cite).
+This also highlights the superiority of SacreBLEU compared to the flawed, standard BLEU.
+
+% BLEU are notoriously unreliable for very short texts.
+The score will be disproportionately influenced by small differences if there isn't much text to begin with because there is a Brevity Penalty set in place to discourage the model from outputting fewer words and get a high score.
+
+OPUS-MT model shows the best performance in all languages, consistently reaching mostly above 60 SacreBLEU and 0.8 METEOR.
+The highest-rated languages are Ukrainian-to-English translation with 75.44 SacreBLEU and 0.86 METEOR, while the lowest scores are achieved on Chinese-to-English translation: 59.50 SacreBLEU and Japanese-to-English: 0.78 METEOR.
+
+The mBART-50 model is the second-best-performing PTMs and the best-performing multilingual model, achieving best SacreBLEU and METEOR score in Italian-to-English translation, while suffering the most in Spanish-to-English (SacreBLEU) and Turkish-to-English (METEOR) translation.
+Interestingly, Spanish received the worst score in SacreBLEU while it performs decently in METEOR, and Turkish performs well in SacreBLEU score but has the worst in METEOR.
+Additionally, the model slightly outperforms OPUS-MT in Polish-to-English translation on SacreBLEU score (0.85 higher).
+
+Similarly, SacreBLEU for Italian-to-English translation outperforms other languages with the M2M-100 model, although Dutch beat Italian by 0.01 in METEOR score.
+Japanese received the worst score in both SacreBLEU and METEOR for this model.
+M2M-100 performs worse in SacreBLEU compared to mBART-50, but comparably similar in METEOR scores.
+
+NLLB-200 scores much lower compared to other models in both SacreBLEU and METEOR in almost all languages.
+It scores highest in Chinese-to-English translation in all metrics, while showing significant variability among other languages.
+However, Japanese-to-English translation suffers with extremely low scores of 1.95 SacreBLEU and 0.09 METEOR.
+Additionally, German, Hebrew, and Ukrainian also have low SacreBLEU and METEOR scores, among the other languages.
 
@@
     \includegraphics[width=1\linewidth]{figures/metrics_bar.png}
-    \caption{BLEU, SacreBLEU, and METEOR scores of each model on each language.}
+    \caption{BLEU, SacreBLEU, and METEOR scores of each model on each language, divided by language}
     \label{fig:result_visual}
@@
 
-Figure \ref{fig:result_visual} shows the visualisation of all scores.
-
-All models perform similarly on BLEU score across all languages, with NLLB-200 suffering a bit more on Japanese translation.
-The OPUS-MT model consistently shows high scores across most languages, with the highest SacreBLEU and METEOR scores overall, especially for Ukrainian and Italian.
-The mBART-50 model achieved the best-performing multilingual model, showing competitive performance, often achieving the second-highest scores across several languages.
-M2M-100 performs worse in SacreBLEU compared to mBART-50, but comparably similar in METEOR scores.
-The NLLB-200 model shows significant variability, especially for Japanese, where the SacreBLEU and METEOR scores are extremely low.
+Figure \ref{fig:result_visual} shows the visualisation of all scores, categorised per language.
 
@@
 Japanese translations, however, pose significant challenges for all models, as indicated by the consistently low scores across all metrics.
+
+Japanese-to-English translation seem to have the worst score on average across the four models, but particularly bad on multilingual models.
+In contrast, Dutch and German, have consistently high results on all models except for NLLB-200, often reaching closely to the highest-language-score SacreBLEU and METEOR score.
 
@@
 Figure \ref{fig:lang_metrics} shows the SacreBLEU and METEOR scores for each language by model.
-The OPUS-MT and mBART-50 models are the clear winners in almost all languages.
-Although M2M-100 performs competitively with mBART-50 in most languages, it surpasses mBART-50 in Spanish-to-English translation.
-On the other hand, NLLB-200 performs the weakest, with extremely low scores in German, Hebrew, and Japanese.
+The OPUS-MT and mBART-50 models are again shown as the clear winners in almost all languages.
+M2M-100 also performs competitively with mBART-50 in some languages, surpassing mBART-50 in Spanish-to-English translation.
+All four models seem to perform consistently against each other in Chinese-to-English and Turkish-to-English translations.
+
+\begin{figure}[htbp]
+    \centering
+    \includegraphics[width=0.49\linewidth]{figures/average_metrics_bar.png}
+    \includegraphics[width=0.49\linewidth]{figures/average_metrics_lang_bar.png}
+    \caption{The average BLEU, SacreBLEU, and METEOR scores by model (left) and by language (right)}
+    \label{fig:average_bars}
+\end{figure}
+
+Figure \ref{fig:average_bars} visualise the average BLEU, SacreBLEU, and METEOR, for each model and each language, highlighting how the models and languages perform independently.
+It can be seen that mBART-50 perform more closely to OPUS-MT on METEOR score than SacreBLEU score.
+Language-wise, Italian, Polish, and Turkish achieves the best average SacreBLEU scores while Chinese, Polish and Turkish are shown as the top three in average METEOR scores.
+
+\begin{figure}[htbp]
+    \centering
+    \includegraphics[width=0.8\linewidth]{figures/opus_bleu_comparison.png}
+    \caption{Comparison between reported OPUS-MT models performance and performance in this experiment.}
+    \label{fig:opus_comparison}
+\end{figure}
+
+Figure \ref{fig:opus_comparison} visualises the difference between each OPUS-MT models reported BLEU scores compared to the SacreBLEU result of this study.
+The reported BLEU is evaluated on the Tatoeba dataset \cite{tiedemann-2020-opus-mt}.
+It can be seen that the scores are quite aligned, with Ukrainian and Italian reaching the two highest score while Chinese, Polish, and Japanese suffers.
+
+This study experiments also resulted in moderately higher overall SacreBLEU score, however, the dataset used here is also significantly less in size than the full Tatoeba challenge dataset \cite{tiedemann-2020-tatoeba-challenge}.
+
+\begin{table}[htbp]
+    \centering
+    \begin{tabular}{|l|r|r|}
+        \hline
+        \textbf{Language} & \textbf{SacreBLEU} & \textbf{Reported BLEU} \\
+        \hline
+        Chinese           & \textit{59.5}      & \textit{36.1}          \\
+        Dutch             & 69.8               & 60.9                   \\
+        Finnish           & 66.6               & 53.4                   \\
+        French            & 69.8               & 57.5                   \\
+        German            & 69.7               & 55.4                   \\
+        Hebrew            & 66.5               & 52.0                   \\
+        Italian           & 74.1               & 70.9                   \\
+        Japanese          & 63.1               & 41.7                   \\
+        Polish            & 61.9               & 54.9                   \\
+        Russian           & 66.7               & 61.1                   \\
+        Spanish           & 71.4               & 59.6                   \\
+        Turkish           & 72.6               & 63.5                   \\
+        Ukrainian         & \textbf{75.4}      & \textbf{64.1}          \\
+        \hline
+    \end{tabular}
+    \caption{OPUS-MT result compared with reported BLEU result}
+    \label{table:opus_comparison}
+\end{table}
+
+Additionally, Table \ref{table:opus_comparison} shows the full scores of both results SacreBLEU and reported BLEU from each OPUS-MT models.
+The reported BLEU scores are taken from each model cards in the HuggingFace repository \cite{huggingface}.
 
@@
 
-Figure \ref{fig:inference_timings} shows the time taken to translate for each language across experimented models.
+Lastly, Figure \ref{fig:inference_timings} shows the time taken to translate for each language across experimented models.
 The OPUS-MT model is the fastest among the other three multilingual models, which perform similarly in timings.
@@
 This study findings show that while multilingual models have been shown to achieve higher performance compared to one-to-one translation models \cite{liu-2020-mbart}, this does not hold on short sentences, many-to-English translations across the 14 source languages.
-Additionally, mBART50 performed better than PTMs M2M-100 and NLLB-200 in these 14 languages.
 
-Limitations, only the first translation of the same phrases are taken.
-No fine-tuning is applied, which may affect the translation results, as fine-tuning models for specific languages can significantly improve models performance \cite{zhang-2023-fine-tuning}.
+OPUS-MT achieves the best performance across all 14 languages.
+mBART-50 is the second-best-performing model, beating OPUS-MT only in Polish SacreBLEU.
+Nevertheless, both mBART-50 and M2M-100 performs generally well on all languages.
+NLLB-200, however, achieves much lower scores in both SacreBLEU and METEOR for all languages except for Chinese, slightly outperforming M2M-100 but still slightly below mBART-50.
 
-\section{Future Works}
+NLLB-200 should then be avoided for use without any further fine-tuning on these 14 languages.
 
-Future works should be done by incorporating more Pre-Trained Models (PTMs) and incorporating fine-tuning for specific languages on languages-to-English translation.
+Chinese, Italian, Polish, Turkish have the best average scores over all four models, Ukrainian by OPUS-MT has the highest SacreBLEU and METEOR score across all the other models and languages.
+
+Limitations, only the first translation of the same phrases are taken, which may disrupts the possibility for synonyms, especially for BLEU and SacreBLEU metrics.
+Additionally, no fine-tuning is applied in this experiment, which may affect the translation results, as fine-tuning models for specific languages can significantly improve models performance \cite{zhang-2023-fine-tuning}.
+
+\section{Discussion}
+
+Future works should be done by incorporating more Pre-Trained Models (PTMs) and implementing fine-tuning for specific languages on languages-to-English translation.
 Evaluating with bigger datasets and longer texts can also be beneficial, as well as using language other than English as the target language.

### 2024-08-14 analyse/analyse_opus_bleu_comp.py

@@ -1,0 +1,90 @@
+import matplotlib.pyplot as plt
+
+# Data from the table
+languages = [
+    "Chinese",
+    "Dutch",
+    "Finnish",
+    "French",
+    "German",
+    "Hebrew",
+    "Italian",
+    "Japanese",
+    "Polish",
+    "Russian",
+    "Spanish",
+    "Turkish",
+    "Ukrainian",
+]
+sacrebleu_scores = [
+    59.5,
+    69.8,
+    66.6,
+    69.8,
+    69.7,
+    66.5,
+    74.1,
+    63.1,
+    61.9,
+    66.7,
+    71.4,
+    72.6,
+    75.4,
+]
+reported_bleu_scores = [
+    36.1,
+    60.9,
+    53.4,
+    57.5,
+    55.4,
+    52.0,
+    70.9,
+    41.7,
+    54.9,
+    61.1,
+    59.6,
+    63.5,
+    64.1,
+]
+# Reversing the data so that Chinese is on top and Ukrainian is on the bottom
+languages_reversed = languages[::-1]
+sacrebleu_scores_reversed = sacrebleu_scores[::-1]
+reported_bleu_scores_reversed = reported_bleu_scores[::-1]
+
+# Plotting the data
+fig, ax = plt.subplots(figsize=(10, 7))
+
+# Plot SacreBLEU scores
+bars1 = ax.barh(
+    languages_reversed,
+    sacrebleu_scores_reversed,
+    color="skyblue",
+    edgecolor="black",
+    label="Result SacreBLEU",
+)
+
+# Plot Reported BLEU scores
+for i, score in enumerate(reported_bleu_scores_reversed):
+    if score is not None:
+        ax.barh(
+            languages_reversed[i],
+            score,
+            color="orange",
+            edgecolor="black",
+            alpha=0.6,
+            label="Reported BLEU" if i == 0 else "",
+        )
+
+# Adding labels and title
+ax.set_xlabel("BLEU Score")
+ax.set_title("Comparison of SacreBLEU and Reported BLEU Scores by Language")
+
+# Creating legend
+handles, labels = ax.get_legend_handles_labels()
+by_label = dict(zip(labels, handles))
+ax.legend(by_label.values(), by_label.keys(), loc="upper left")
+
+# Display the plot
+plt.tight_layout()
+
+plt.savefig(f"./figures/opus_bleu_comparison.png")

### 2024-08-14 analyse/analyse_results.py

@@ -31,7 +31,6 @@ MODELS = [
 
 BAR_COLOURS = ["tab:blue", "tab:orange", "tab:green", "tab:red"]
 
-
 with open("../inference/result/metrics.json", "r") as f:
     metrics = json.loads(f.read())
 
@@ -53,30 +52,27 @@ def plot_scores(dataframe, metric_name, ax, add_legend=False):
 
 bleu_scores = pd.DataFrame(index=LANGS)
 sacre_bleu_scores = pd.DataFrame(index=LANGS)
-meteor_bleu_scores = pd.DataFrame(index=LANGS)
+meteor_scores = pd.DataFrame(index=LANGS)
 
 for model in MODELS:
     bleu_scores[model] = [metrics[model][lang]["bleu"] for lang in LANGS]
     sacre_bleu_scores[model] = [metrics[model][lang]["sacre_bleu"] for lang in LANGS]
-    meteor_bleu_scores[model] = [metrics[model][lang]["meteor"] for lang in LANGS]
+    meteor_scores[model] = [metrics[model][lang]["meteor"] for lang in LANGS]
 
 fig, (ax1, ax2, ax3) = plt.subplots(3, 1, figsize=(12, 10), sharex=True)
 
 plot_scores(bleu_scores, "BLEU", ax1, add_legend=True)
-
 plot_scores(sacre_bleu_scores, "SacreBLEU", ax2)
-
-plot_scores(meteor_bleu_scores, "METEOR", ax3)
+plot_scores(meteor_scores, "METEOR", ax3)
 
 plt.tight_layout()
 plt.savefig(f"./figures/metrics_bar.png")
-
 plt.clf()
 
 # Compute average scores for each model
 average_bleu_scores = bleu_scores.mean()
 average_sacre_bleu_scores = sacre_bleu_scores.mean()
-average_meteor_scores = meteor_bleu_scores.mean()
+average_meteor_scores = meteor_scores.mean()
 
 # Create a DataFrame for average scores
 average_scores = pd.DataFrame(
@@ -87,35 +83,64 @@ average_scores = pd.DataFrame(
     }
 )
 
-fig, (ax4, ax5, ax6) = plt.subplots(3, 1, figsize=(8, 10))
+fig, (ax4, ax5) = plt.subplots(2, 1, figsize=(8, 10))
 
 bar_width = 0.8
 index = np.arange(len(average_scores.index))
 
-# Plot average BLEU scores
[…]

### 2024-08-14 inference/opus_multilingual.py

@@ -70,8 +70,8 @@ for src_lang, src_lang_code in LANGS.items():
     end = time.time()
     timing_df[f"translated_from_{src_lang}"] = [end - start]
 
-    translated_df.to_csv(f"result/opus.csv")
-    timing_df.to_csv(f"result/opus_timings.csv")
+    translated_df.to_csv(f"result/opus_mt.csv")
+    timing_df.to_csv(f"result/opus_mt_timings.csv")
 
 
 # not entirely sure how to use the mt model

### 2024-08-19 paper/paper.tex
slight update
updated, now 16 pages

@@
-\documentclass[a4paper]{article}
+\documentclass[a4paper, 11pt]{article}
 
@@
 The results indicate that OPUS-MT, a one-to-one PTM, outperforms multilingual models in one-to-English translation tasks, with mBART-50 leading among the multilingual PTMs.
-These findings highlight the strengths and limitations of both specific language and multilingual PTMs in machine translation.
+Chinese, Italian, Polish, and Turkish translations to English show the best performance on average across the four models used in this study, however, Ukrainian-to-English on the OPUS-MT model achieves the highest SacreBLEU and METEOR scores.
 \end{abstract}
@@
 
-According to Ethnologue \cite{ethnologue-2024}, 7,164 languages currently exist and in use today, with 40\% of them considered endangered.
-As of July 2024, 243 languages are supported by Google Translate (according to Wikipedia \cite{wikipedia-google-translate}).
-In modern times, the need for translation services has surged due to the growing exchange of information across different regions that speak various languages \cite{okpor-2014-machine-ta}.
+According to Ethnologue \cite{ethnologue-2024}, 7,164 languages currently exist and are in use today, with 40\% of them considered endangered.
+With so many languages around us, the language barrier has become a known issue, especially in the current globalised world where it is essential for individuals who speak different languages to be able to communicate effectively \cite{nakamura-2009-overcoming}.
+Thus, the need for translation services to facilitate communication across different languages has also surged \cite{okpor-2014-machine-ta}.
+As of August 2024, only 243 languages are supported by Google Translate (according to Wikipedia \cite{wikipedia-google-translate}).
 
-Machine translation (MT) is the task of automatically translating from one language to another.
-This can be done through text or audio.
-It can be traced back to 1949 \cite{weaver-1999}, with the first public demonstration of an MT system on January 7, 1954, in collaboration with IBM, where 49 Russian sentences were translated into English using a limited vocabulary of 250 words and 6 grammar rules \cite{hutchins-2006-first-mt}.
-However, over the next several decades, growth were limited for machine translation, with 1956-1966 considered the decade of high expectation and disillusion, and 1967-1976 dubbed 'the quiet decade' \cite{hutchins-2001-mt-50-years}.
-Then in 1989, the dominance of the rule-based approach has been challenged by the rise of new methods and strategies, collectively referred to as ‘corpus-based’ methods (data-driven) \cite{hutchins-1994-research-methods-mt,hutchins-1998-development-mt}.
+Machine translation (MT) refers to the task of automatically translating from one language to another, which typically can be done through text or audio.
+The research and practice can be traced back to 1949 \cite{weaver-1999}, with the first public demonstration of an MT system on January 7, 1954, in collaboration with IBM, where 49 Russian sentences were translated into English using a limited vocabulary of 250 words and 6 grammar rules \cite{hutchins-2006-first-mt}.
+However, over the next several decades, growth was limited for machine translation, with 1956-1966 considered the decade of high expectation and disillusion, and 1967-1976 dubbed 'the quiet decade' \cite{hutchins-2001-mt-50-years}.
+Then in 1989, the dominance of the rule-based approach was challenged by the rise of new methods and strategies, collectively referred to as ‘corpus-based’ methods (data-driven) \cite{hutchins-1994-research-methods-mt,hutchins-1998-development-mt}.
 Subsequently, statistics-based approaches for MT re-emerged, bolstered by the recent success of probabilistic techniques in speech recognition.
-Statistical machine translation \cite{lopez-2008-smt} dominated the domain between late 1990s through the early 2010s, before largely being surpassed by neural machine translation (NMT) \cite{cho-2014-properties,sutskever-2014-seq2seq}.
+Statistical machine translation \cite{lopez-2008-smt} dominated the domain between the late 1990s through the early 2010s, before largely being surpassed by neural machine translation (NMT) \cite{cho-2014-properties,sutskever-2014-seq2seq}.
 
@@
 The following years saw the birth of Large Language Models (LLMs) such as BERT \cite{devlin-2019-bert}, GPT \cite{openai-2024-gpt4}, and T5 \cite{raffel-2023-t5}, which revolutionised both MT and the whole field of NLP.
-Then in early 2020s, several pre-trained models (PTM) that are specifically designed for machine translation emerged, OPUS-MT \cite{tiedemann-2020-opus-mt} produced numerous PTMs designed for one-to-one translations.
-Additionally, multilingual models: mBART-50 \cite{liu-2020-mbart}, NLLB-200 \cite{nllb200-2020}, and M2M-100 \cite{fan-2020-m2m100} have also been published, growing popularity as multilinguality has been shown to allow for sharing information among languages \cite{garcia-2020-multilingual}.
-Most of these models are trained on diverse set of languages, allowing for many-to-many translation: able to translate between any of the supported pair of languages.
+Then in the early 2020s, several pre-trained models (PTM) that are specifically designed for machine translation emerged, OPUS-MT \cite{tiedemann-2020-opus-mt} produced numerous PTMs designed for one-to-one translations.
+Additionally, multilingual models: mBART-50 \cite{liu-2020-mbart}, NLLB-200 \cite{nllb200-2020}, and M2M-100 \cite{fan-2020-m2m100} have also been published, growing popularity as multilingualism has been shown to allow for sharing information among languages \cite{garcia-2020-multilingual}.
+Most of these models are trained on a diverse set of languages, allowing for many-to-many translation, which is able to translate between any of the supported pairs of languages.
 This allows the models to generalise over shared lexical and linguistic among languages, and have been shown to increase performance compared to one-to-one translation models \cite{liu-2020-mbart}.
 
-Despite these advancements, pre-trained models are often evaluated using different set of languages and benchmarks \cite{liu-2020-mbart,nllb200-2020,fan-2020-m2m100,wei-2023-polylm}, making it difficult to gauge their relative effectiveness across various languages.
+Despite these advancements, pre-trained models are often evaluated using a different set of languages and benchmarks \cite{liu-2020-mbart,nllb200-2020,fan-2020-m2m100,wei-2023-polylm}, making it difficult to gauge their relative effectiveness across various languages.
 Nevertheless, there does not seem to be much work on comparing or benchmarking different pre-trained models in machine translation.
-Intento published 'The State of Machine Translation 2024' \cite{intento-2020} providing an in-depth evaluation of popular MT engines and LLMs.
-However, the biggest drawback in this report is that the selected LLMs are general LLMs such as GPT, LLaMa, Mistral, instead of MT-specific LLMs.
+Intento recently published 'The State of Machine Translation 2024' \cite{intento-2020} providing an in-depth evaluation of 52 MT engines and LLMs across 11 language pairs.
+Intento reported a lack of growth in the number of supported languages and concluded that translation quality significantly varies depending on languages \cite{intento-2020}.
+However, the biggest drawback in this report is that the selected LLMs are general LLMs such as GPT-4 \cite{openai-2024-gpt4}, LLaMa2 \cite{touvron-2023-llama2}, and Mistral \cite{jiang-2023-mistral-7b}, instead of MT-specific models.
 
-Through simple experimentations, this paper aims to evaluate the performance of existing pre-trained models (PTMs) on many-to-English translation across 14 source languages.
-Although fine-tuning multilingual PTMs has been proven to increase model performance \cite{cooper-stickland-2021-recipes}, no pre-training or fine-tuning is performed in this study for simplicity reasons.
+Furthermore, BLEU \cite{papieni-2002-bleu} has been shown to be inconsistent depending on tokenisation and parameters that are used \cite{post-2018-sacrebleu}.
+While most researchers do report on these specifications explicitly, different BLEU parametrisations between different papers still make a direct comparison challenging to do.
+
+This study will lean towards evaluating MT-specific models rather than general LLMs.
+Through simple experimentations, this paper aims to evaluate the performance of existing and publicly available MT pre-trained models (PTMs) on many-to-English translation across 14 source languages.
+Although fine-tuning multilingual PTMs has been proven to increase translation performance \cite{cooper-stickland-2021-recipes}, no pre-training or fine-tuning is performed in this study due to simplicity reasons and resource limitations.
 A dataset is curated from the Tatoeba repository \cite{tatoeba}, containing 1,323 parallel sentence pairs across source and target languages.
@@
 
-The result indicates that OPUS-MT outperforms all other multilingual models in one-to-English translation, highlighting the superiority of specific languages PTMs compared to multilingual PTMs.
-mBART-50 shows the best performance among other multilingual PTMs: M2M-100 and NLLB-200.
+The result indicates that OPUS-MT outperforms all other multilingual models in one-to-English translation on both SacreBLEU and METEOR scores, highlighting the superiority of specific languages PTMs compared to unfine-tuned multilingual PTMs.
+Among the multilingual models, mBART-50 shows the best performance, while NLLB-200 performs the worst, by a significant margin.
+Ukrainian-to-English translation by OPUS-MT receives the highest scores, however, Chinese, Italian, Polish, and Turkish are better on average across the four models.
 
@@
 
-The encoder-decoder approach \cite{cho-2014-properties} remains as the foundation architecture for many sequence-to-sequence models in machine translation.
-Attention mechanism has been used for Neural Machine Translation (NMT) \cite{bahdanau-2016-nmt-jointly, luong-2015-effective-attention} before the release of the transformer \cite{vaswani-2017-attention}
-
-Currently, Pre-Trained-Models (PTMs) have become a cornerstone in machine translation, leveraging large datasets and powerful architectures to achieve state-of-the-art performance \cite{han-2021-ptms}.
+Currently, pre-trained models (PTMs) have become a cornerstone in machine translation, leveraging large datasets and powerful architectures to achieve state-of-the-art performance \cite{han-2021-ptms}.
+Attention mechanism has been used for Neural Machine Translation (NMT) \cite{bahdanau-2016-nmt-jointly, luong-2015-effective-attention} even before the introduction of the transformer \cite{vaswani-2017-attention}.
+The transformer, with its encoder-decoder approach \cite{cho-2014-properties}, remains the foundational architecture for many sequence-to-sequence models in machine translation.
 Most PTMs support two kinds of translations: one-to-one translation and many-to-many translation (multilingual models).
 
-One-to-One Translation refers to a translation approach where a model is specifically trained to translate between one source language and one target language.
+One-to-one translation refers to a translation approach where a model is specifically trained to translate between one source language and one target language.
 This setup is characterised by having a dedicated model for each unique language pair.
-OPUS-MT \cite{tiedemann-2020-opus-mt} and Facebook Fairseq \cite{ott-2019-fairseq} are examples of PTMs that support one-to-one translation.
-Multilingual models have also gained significant prominence, enabling many-to-many translation, where a single model can translate between multiple source and target languages \cite{aharoni-2019-massively-multilingual}.
+OPUS-MT \cite{tiedemann-2020-opus-mt} is a project dedicated to creating and providing free resources for machine translation, providing over a thousand PTMs of direct one-to-one translations.
+Additionally, Facebook Fairseq \cite{ott-2019-fairseq} similarly offers one-to-one PTMs, although they are only limited to translations between English, French, German, and Russian.
 
-OPUS-MT \cite{tiedemann-2020-opus-mt} is a project dedicated to creating and providing free resources for machine translation,providing over a thousand of PTMs of direct one-to-one translations.
-In contrast, mBART \cite{liu-2020-mbart}, M2M \cite{fan-2020-m2m100}, NLLB-200 \cite{nllb200-2020}, are massively multilingual models that support many-to-many translation language.
+Many-to-many translation allows for a translation between multiple pairs of source and target languages.
+This is done by training a single multilingual model, typically trained on multilingual datasets.
+Multilingual models have recently gained significant prominence, shown to be simpler and efficient \cite{aharoni-2019-massively-multilingual}.
+mBART \cite{liu-2020-mbart}, M2M \cite{fan-2020-m2m100}, NLLB-200 \cite{nllb200-2020}, are examples of massively multilingual models that support many-to-many translation language.
 
@@
 Thus, parallel corpora are defined as sets of texts in a given source language along with their translations in another target language \cite{lefer-2020-parallel-corpora}.
-It can be bilingual or multilingual, and are crucial for training and evaluating machine translation.
-Languages without or with limited parallel corpora are referred as \textbf{low-resources language}.
+It can be bilingual or multilingual and is crucial for training and evaluating machine translation.
+Languages without or with limited parallel corpora are referred to as low-resource languages.
 
 Since the early 2000s, English has often been used as an intermediary language, meaning texts originally written in languages are first translated into English, and then from English into other languages, resulting in most target texts being translations of translations \cite{lefer-2020-parallel-corpora}.
-This is called indirect translation, and posed for many problems in translations as terms can be ambiguous with varying or similar meanings \cite{rosa-2017-indirect-translation-problems}.
+This is called indirect translation and poses many problems in translations as terms can be ambiguous with varying or similar meanings \cite{rosa-2017-indirect-translation-problems}.
 
@@
 As of July 2024, it contains 12,186,207 sentences over 423 supported languages, growing daily through volunteer contributions.
+There are currently
+86,915 members contributed through the Tatoeba website.
 
 Several datasets are often used for evaluation in machine translation.
-The Tatoeba Challenge \cite{tiedemann-2020-tatoeba-challenge} covers 487 languages in 4,024 language pairs, including 657 test sets sourced from Tatoeba website, covering 138 languages.
+The Tatoeba Challenge \cite{tiedemann-2020-tatoeba-challenge} covers 487 languages in 4,024 language pairs, including 657 test sets sourced from the Tatoeba website, covering 138 languages.
 The TED Talks dataset \cite{ye-2018-word-embeddings-ted} contains transcripts from TED talks for more than 50 languages.
 IWSLT \cite{agarwal-iwstl-2023} also contains TED talks data, but paired with English translations.
-Finally, WMT \cite{barrault-2020-wmt} is an annual event that organises tasks for machine translation and provides a collection of datasets for benchmarking and evaluating translation systems.
+Finally, WMT (Conference on Machine Translation) \cite{barrault-2020-wmt}, previously named Workshop on Statistical Machine Translation, is an annual event that organises tasks for machine translation and provides a collection of datasets for benchmarking and evaluating translation systems.
 
-While these datasets can be used to evaluate translation performance between any language pairs, they do not contain parallel corpora between different languages.
-Corpus between English and French for example, contains different text to the corpus between English and Spanish.
+While these datasets can be used to evaluate translation performance between any language pairs, they do not contain parallel corpora between multiple different languages.
+A corpus between English and French, for example, contains different sets of text than a corpus between English and Spanish.
 Therefore, they do not support easy evaluation for many-to-English translations.
-Furthermore, papers often use different datasets and versions to evaluate their model performance, making direct comparison challenging.
-
+Furthermore, papers often use different datasets and versions to evaluate their model performance, making direct comparisons challenging.
 mBart \cite{liu-2020-mbart} uses WMT19 for English-German translation and TED15 for Chinese-English translation.
-M2M-100 \cite{fan-2020-m2m100} uses 7 different datasets shared across languages.
-NLLB-200  \cite{nllb200-2020} uses FLORES dataset, designed for low-resources languages.
-OPUS-MT \cite{tiedemann-2023-democratizing,tiedemann-2020-opus-mt} models use datasets from Tatoeba challenges.
+M2M-100 \cite{fan-2020-m2m100} uses seven different datasets shared across languages.
+NLLB-200  \cite{nllb200-2020} uses the FLORES dataset, designed for low-resources languages.
+OPUS-MT \cite{tiedemann-2023-democratizing,tiedemann-2020-opus-mt} models are evaluated using the datasets from Tatoeba challenges.
 
@@
 
-Tatoeba is a vast, continuously expanding database consisting sentences and their translations, built through the contributions of thousands of volunteers, offering a tool that allows users to see examples of how words are used in sentences \cite{tatoeba}.
-They currently have 12,132,349 sentences and 423 supported languages, with around one to two thousand new sentences added daily, on average.
-The English sentence dataset contains 1,905,089 sentences, the largest one in their repository, with Russian in the second place with 1,066,633 sentences.
-Some languages supported on the website is shown in Figure \ref{fig:tatoeba_languages} and Figure \ref{fig:tatoeba_top_bottom_languages}, sorted from the biggest corpus.
+The Tatoeba website \cite{tatoeba} includes a vast, continuously expanding database consisting of sentences and their translations, built through the contributions of thousands of volunteers.
+They currently have 12,132,349 sentences of over 423 supported languages, with around one to two thousand new sentences added daily, on average.
+The English sentence dataset contains 1,905,089 sentences, the largest one in their repository, with Russian in second place with 1,066,633 sentences.
+Figure \ref{fig:tatoeba_languages} shows the repository, sorted from the biggest corpus.
 
@@
     \centering
-    \includegraphics[width=0.9\linewidth]{images/tatoeba_languages.png}
-    \caption{Tatoeba's languages repository with 10,000+ sentences and 100,000+ sentences \cite{tatoeba}}
+    \includegraphics[width=0.65\linewidth]{images/tatoeba_languages.png}
+    \includegraphics[width=0.34\linewidth]{images/tatoeba_top_20_lang.png}
+    \caption{Tatoeba's repository by languages, sorted by sentences count \cite{tatoeba}}
     \label{fig:tatoeba_languages}
-\end{figure}
-
-\begin{figure}[htbp]
-    \centering
-    \includegraphics[width=0.5\linewidth]{images/tatoeba_top_20_lang.png}
-    \includegraphics[width=0.46\linewidth]{images/tatoeba_bottom_20_lang.png}
-    \caption{Tatoeba top 20 and bottom 20 languages based on sentences count \cite{tatoeba}}
-    \label{fig:tatoeba_top_bottom_languages}
 \end{figure}
@@
     \large
-    \begin{tabular}{|l|l|r|}
+    \begin{tabular}{llr}
         \hline
@@
         \hline
+                     & Total                   & 4,461,119                     \\
+        \hline
     \end{tabular}
@@
 
-Table \ref{table:eval_languages} show the 14 languages selected for this project.
-Languages are chosen based on its resources' availability in Tatoeba, as well as considering supported languages in most PTMs models.
-Thus, the languages chosen here can be considered as moderate to high resources languages.
+Table \ref{table:eval_languages} shows the 14 language pairs selected for this study.
+Languages are chosen based on the resources' availability in Tatoeba and supported languages in most PTMs.
+Thus, the languages chosen here can be considered moderate to high-resource languages.
 
-To build the dataset, sentences in English are first downloaded, containing 1,898,494 sentences (it is unclear why it is less than the number stated in the Tatoeba website).
-Then for each language, sentence pairs between English and source languages are downloaded individually and compiled.
-The result is a single Dataframe containing 1,323 parallel sentences in all 14 languages, this will be treated as a test set to evaluate the models performance on each language.
-Note that in the case of multiple reference translations are available for the same sentence in English, only the first sentence is taken while the rest is discarded.
-Therefore, there is only one reference for each candidate sentence.
+To build the dataset, 1,898,494 English sentences were initially downloaded, though the number is less than what is listed on the Tatoeba website for reasons that remain unclear.
+For each of the 14 source languages, sentence pairs in English were individually downloaded and compiled.
+The English corpus was then merged with each compiled source language corpus to create parallel sentences across 14 languages.
+In each corpus, multiple translations for the same English phrase were dropped, retaining only the first translation to ensure that each English sentence had only a single translation reference.
+The final dataset contains 1,323 parallel sentences in English and all 14 source languages.
+This corpus will be used as a test set to evaluate the models' performance in each language.
 
@@
 People kill people' can be found inside the corpus.
-A few of them also include human names, 'Compare your answer with Tom's', 'Muiriel is 20 now'.
-All of the sentences are straightforward and literal, without the use of linguistic features such as metaphors or sarcasm.
-Therefore, machine translation process should be straightforward on this level.
-Figure \ref{table:parallel_sentence} shows two examples of parallel sentences in the final dataset, across all 14 languages and the original English sentence.
+A few of them also include human names, 'Compare your answer with Tom's', and 'Muiriel is 20 now'.
+All the sentences seem to be straightforward and literal, without any use of linguistic features such as metaphors or sarcasm.
+Figure \ref{table:parallel_sentence} shows a couple of examples of parallel sentences in the final dataset, across all 14 languages and the original English sentence.
 
@@
         \centering
-        \begin{tabular}{|l|l|l|}
+        \begin{tabular}{lrr}
             \hline
@@
 
-Figure \ref{fig:word_count_box} shows a box plot of sentences word count.
-The median for Dutch, Finnish, French, German, Hebrew, Italian, Polish, Russian, Spanish, Turkish, Ukrainian seems relatively low, with a moderate spread.
-A few to moderate count of outlier exist in all languages, suggesting occasional longer sentences, with the highest variability seen in Japanese.
-Note that for Chinese and Japanese the count is based on each Chinese/Japanese letter, which explains their higher interquartile range (IQR) and larger numbers of outliers.
+Figure \ref{fig:word_count_box} shows a box plot of each language sentence's word count.
+Note that for Chinese and Japanese the word count is counted for each Chinese/Japanese letter instead, which explains their higher range and larger numbers of outliers.
+
+Most of the languages' interquartile range (IQR) is quite low (around two) with a centre median, except for Chinese, Italian, and Japanese.
+Finnish and Hebrew have the lowest median, while Polish, Russian, Spanish, Turkish, and Ukrainian have a median closely similar to each other.
+English, Dutch, French, German, and Italian have a slightly higher median, with Italian also having a wider IQR.
+Lastly, Japanese and Chinese have the highest median, the widest range, and significantly more outliers.
 
@@
 
-\textbf{OPUS-MT} \cite{tiedemann-2023-democratizing,tiedemann-2020-opus-mt} provides over 1,000 pre-trained models for translation between numerous language pairs.
+OPUS-MT \cite{tiedemann-2023-democratizing,tiedemann-2020-opus-mt} provides over 1,000 pre-trained models for translation between numerous language pairs.
 The architecture is based on MARIAN-NMT \cite{mariannmt}, based on a standard transformer setup: 6 self-attentive layers in both the encoder and decoder networks, each with 8 attention heads per layer \cite{tiedemann-2020-opus-mt}.
 
-While they provide both monolingual and multilingual PTMs, only the OPUS-MT monolingual models are used in this study.
+While the framework provides both monolingual and multilingual PTMs, only the OPUS-MT monolingual models are used in this study.
+This is partly due to the lack of documentation on the multilingual models.
 
@@
 
-\textbf{mBART} \cite{liu-2020-mbart} is a sequence-to-sequence denoising auto-encoder model specifically designed for multilingual tasks.
+mBART \cite{liu-2020-mbart} is a sequence-to-sequence denoising auto-encoder model specifically designed for multilingual tasks.
 The mBART-50 variant supports many-to-many translations for over 50 languages.
@@
 
-The mBART model is built as a typical sequence-to-sequence transformer architecture \cite{vaswani-2017-attention}, using an embedding dimension of 1024 and 16 heads within 12 layers of encoder and 12 layers of decoder, totalling to around 680M parameters \cite{liu-2020-mbart}.
+The mBART model is built as a typical sequence-to-sequence transformer architecture \cite{vaswani-2017-attention}, using an embedding dimension of 1024 and 16 heads within 12 layers of encoder and 12 layers of decoder, totalling around 680M parameters \cite{liu-2020-mbart}.
 
@@
 After the pre-training step, the model is then able to be fine-tuned for downstream tasks.
-It is shown that the model achieved consistent performance gains through pre-training in low-to-medium resource sentence level MT \cite{liu-2020-mbart}
+It is shown that the model achieved consistent performance gains through pre-training in low-to-medium resource sentence-level translation \cite{liu-2020-mbart}
 
@@
 
-\textbf{M2M-100} \cite{fan-2020-m2m100} is designed to perform direct translation between any pair of 100 languages without relying on English as an intermediate language.
-Similarly, the architecture is of a transformer sequence-to-sequence with a typical encoder that outputs embeddings of the same length, while the decoder is an autoregressive component that sequentially produce the target sentence one token at a time.
-The parameters include 12 encoder layers and 12 decoder layers, 8192 FFN size and 1024 embedding dimension, totalling to a 1.2B parameter count \cite{fan-2020-m2m100}.
+M2M-100 \cite{fan-2020-m2m100} is designed to perform direct translation between any pair of 100 languages without relying on English as an intermediate language.
+Similarly, the architecture is of a transformer sequence-to-sequence with a typical encoder that outputs embeddings of the same length, while the decoder is an autoregressive component that sequentially produces the target sentence one token at a time.
+The parameters include 12 encoder layers and 12 decoder layers, 8192 hidden layer sizes, and 1024 embedding dimensions, totalling a 1.2B parameter count \cite{fan-2020-m2m100}.
 
@@
 
-Figure \ref{fig:m2m} show the inner working of the M2M model.
+Figure \ref{fig:m2m} shows the inner workings of the M2M model.
 The English-centric dataset (top left) includes training data exclusively involving translations to and from English, while the many-to-many multilingual setting (top right) involves direct translation data among multiple language pairs \cite{fan-2020-m2m100}.
@@
 
-\textbf{NLLB-200} \cite{nllb200-2020} is built to handle translation tasks across a broad spectrum of languages, including many that are low-resource or underrepresented in existing datasets.
+NLLB-200 \cite{nllb200-2020} is built to handle translation tasks across a broad spectrum of languages, including many that are low-resource or underrepresented in existing datasets.
 It supports translations for 200 languages, including numerous underrepresented languages, and is currently one of the most extensive multilingual machine translation models.
@@
 
-% the most assured way to determine the best MT system from competing alternatives is to have an assessment done by competent human translators.
-
-In this project, a variant of BLEU \cite{papieni-2002-bleu} called SacreBLEU \cite{post-2018-sacrebleu} and METEOR \cite{lavie-2007-meteor} will be used in this experiment due to their popularity, simplicity, and ease-of-use.
-
+In this study, BLEU \cite{papieni-2002-bleu}, SacreBLEU \cite{post-2018-sacrebleu}, and METEOR \cite{lavie-2007-meteor} are applied as evaluation metrics due to their popularity, simplicity, and ease-of-use.
 While other metrics such as Crosslingual Optimized Metric for Evaluation of Translation (COMET) \cite{rei-2020-comet} and BERTSCore \cite{zhang-2020-bertscore} exist, they involve using a deep neural network or a transformer to evaluate, which naturally increase computational cost and slow down calculation greatly.
@@
 
-Bilingual Evaluation Understudy (BLEU) \cite{papieni-2002-bleu} is the most commonly used metrics for machine translation (MT).
+Bilingual Evaluation Understudy (BLEU) \cite{papieni-2002-bleu} is the most commonly used metric for machine translation (MT).
 It assesses how well a candidate translation matches the reference translation using precision metrics for n-grams and incorporates a brevity penalty to prevent overly short translations from achieving high scores.
@@
 
-The machine translation community's rely heavily BLEU score, however, it has several drawbacks.
+The machine translation community relies heavily upon the BLEU score, despite its several drawbacks.
 The metric has been reported to not correlate strongly with human judgement, showing variations in translation that could mean that a higher BLEU score does not necessarily indicate a true enhancement in translation quality \cite{callison-burch-2006-reevaluating-bleu}.
-
-Furthermore, It is challenging to directly compare BLEU scores between paper \cite{post-2018-sacrebleu}.
-Thus, the author proposed a standardised variant called SacreBLEU \cite{post-2018-sacrebleu}, designed to ensure easy, consistent, and comparable evaluation across different implementations and research studies.
+Furthermore, the metric is tokenisation-dependant, making it challenging to directly compare BLEU scores between paper \cite{post-2018-sacrebleu}.
+Thus, the author proposed a standardised variant called SacreBLEU \cite{post-2018-sacrebleu}, which is essentially a Python script that applies its own metric-internal preprocessing and produces the same values as WMT \cite{barrault-2020-wmt}, utilising the WMT (2008–2018) and IWSLT 2017 \cite{cettolo-etal-2017-iwslt} test sets.
 
@@
 
-OPUS-MT models are taken from the Helsinki-NLP repository.
-The 'mbart-large-50-many-to-many-mmt' is used, 'm2m100\_418M', and 'nllb-200-distilled-600M'.
+OPUS-MT models are taken from the Helsinki-NLP repository with the exception of 'tiedeman/opus-mt-he-en' for Hebrew-to-English translation.
+The 'mbart-large-50-many-to-many-mmt', the 'm2m100\_418M', and the 'nllb-200-distilled-600M' models are used for mBART-50 \cite{liu-2020-mbart}, M2M-100 \cite{fan-2020-m2m100}, and NLLB-200 \cite{nllb200-2020}, respectively.
 All models are taken using the transformers \cite{wolf-etal-2020-transformers} library from Hugging Face \cite{huggingface}.
-Batch size of 4 is used in all models.
+A batch size of 4 is used in all models.
 Translations are first generated and saved into CSV files, before calculating the metrics score.
 
-The sacrebleu packages from pip is used to calculate SacreBLEU \cite{post-2018-sacrebleu}.
-For BLEU and METEOR, the NLTK \cite{bird-2009-natural} package is used.
-Corpus BLEU is implemented for both SacreBLEU and BLEU in the calculations.
-Additionally, smoothing method 7 is used for BLEU, which interpolates two other smoothing functions: smoothing method 4, which divides by \( \frac{1}{\ln(\text{len}(T))} \), where \( T \) is the length of the translation; and smoothing method 5, which averages the \( n-1 \), \( n \), and \( n+1 \) gram matched counts \cite{nltk-bleu-docs}.
+For BLEU and METEOR, tokenisation and metrics calculation use the NLTK \cite{bird-2009-natural} package, specifically the \texttt{nltk.word\_tokenize}, \texttt{nltk.translate.meteor\_score}, and \texttt{nltk.translate.bleu\_score.corpus\_bleu}.
+Additionally, smoothing method 7 is used for the BLEU calculation, which interpolates two other smoothing functions: smoothing method 4, which divides by \( \frac{1}{\ln(\text{len}(T))} \), where \( T \) is the length of the translation; and smoothing method 5, which averages the \( n-1 \), \( n \), and \( n+1 \) gram matched counts \cite{nltk-bleu-docs}.
+The \texttt{sacrebleu} \cite{post-2018-sacrebleu} package from pip is used to calculate SacreBLEU \cite{post-2018-sacrebleu}.
 
-All codes are written in Python 3.12.2, run locally on MacOS Sonoma 14.4.1 with M2 chip and 16 GB memory. GitHub: \url{https://github.com/stefanliemawan/eval-machine-translation}
+All codes are written in Python 3.12.2 and run locally on MacOS Sonoma 14.4.1 with an M2 chip and 16 GB memory.
+Datasets and scripts will be available on GitHub: \url{https://github.com/stefanliemawan/eval-machine-translation}
 
@@
         \footnotesize
-        \begin{tabular}{|l|r|r|r|}
+        \begin{tabular}{lrrr}
             \hline
@@
 Table \ref{table:opus_result}, \ref{table:mbart_result}, \ref{table:m2m_result}, and \ref{table:nllb_result}, and  shows the result from all four PTMs.
-All models perform similarly low on BLEU score across all languages, with NLLB-200 reaching 0.01 below more on Japanese translation.
+All models perform similarly low on BLEU score across all languages, with NLLB-200 reaching 0.01 below on Japanese translation.
 The BLEU scores are very low (below 10 BLEU) in all cases due to the sentences being short (find cite).
@@
 
-% BLEU are notoriously unreliable for very short texts.
-The score will be disproportionately influenced by small differences if there isn't much text to begin with because there is a Brevity Penalty set in place to discourage the model from outputting fewer words and get a high score.
+% BLEU is notoriously unreliable for very short texts.
+The score will be disproportionately influenced by small differences if there isn't much text to begin with because there is a Brevity Penalty set in place to discourage the model from outputting fewer words and getting a high score.
 
@@
 
-The mBART-50 model is the second-best-performing PTMs and the best-performing multilingual model, achieving best SacreBLEU and METEOR score in Italian-to-English translation, while suffering the most in Spanish-to-English (SacreBLEU) and Turkish-to-English (METEOR) translation.
-Interestingly, Spanish received the worst score in SacreBLEU while it performs decently in METEOR, and Turkish performs well in SacreBLEU score but has the worst in METEOR.
+The mBART-50 model is the second-best-performing PTM and the best-performing multilingual model, achieving the best SacreBLEU and METEOR score in Italian-to-English translation while suffering the most in Spanish-to-English (SacreBLEU) and Turkish-to-English (METEOR) translation.
+Interestingly, Spanish receives the worst score in SacreBLEU while it performs decently in METEOR, and Turkish performs well in SacreBLEU score but has the worst in METEOR.
 Additionally, the model slightly outperforms OPUS-MT in Polish-to-English translation on SacreBLEU score (0.85 higher).
@@
     \includegraphics[width=1\linewidth]{figures/metrics_bar.png}
-    \caption{BLEU, SacreBLEU, and METEOR scores of each model on each language, divided by language}
+    \caption{BLEU, SacreBLEU, and METEOR scores of each model on each language}
     \label{fig:result_visual}
@@
 
-Figure \ref{fig:result_visual} shows the visualisation of all scores, categorised per language.
+Figure \ref{fig:result_visual} visualises all scores gained by each model, categorised by language.
 
-Dutch and Italian translations generally score higher across all metrics, suggesting that these languages are well-supported by the pre-trained models.
+Dutch, Italian, and Ukrainian translations generally score higher across all metrics, suggesting that these languages are well-supported by the pre-trained models.
 Japanese translations, however, pose significant challenges for all models, as indicated by the consistently low scores across all metrics.
 
-Japanese-to-English translation seem to have the worst score on average across the four models, but particularly bad on multilingual models.
-In contrast, Dutch and German, have consistently high results on all models except for NLLB-200, often reaching closely to the highest-language-score SacreBLEU and METEOR score.
+Japanese-to-English translation seems to have the worst score on average across the four models, but particularly bad on multilingual models.
+In contrast, Dutch and German have consistently high results on all models except for NLLB-200, often reaching close to the highest language SacreBLEU and METEOR scores.
 
@@
 
-Figure \ref{fig:lang_metrics} shows the SacreBLEU and METEOR scores for each language by model.
+Figure \ref{fig:lang_metrics} shows the SacreBLEU and METEOR scores for each language, categorised by model.
 The OPUS-MT and mBART-50 models are again shown as the clear winners in almost all languages.
-M2M-100 also performs competitively with mBART-50 in some languages, surpassing mBART-50 in Spanish-to-English translation.
+M2M-100 also performs competitively with mBART-50 in most languages, surpassing mBART-50 in Spanish-to-English translation.
+NLLB-200 is clearly shown to suffer greatly across all languages besides Chinese.
 All four models seem to perform consistently against each other in Chinese-to-English and Turkish-to-English translations.
@@
 Figure \ref{fig:average_bars} visualise the average BLEU, SacreBLEU, and METEOR, for each model and each language, highlighting how the models and languages perform independently.
-It can be seen that mBART-50 perform more closely to OPUS-MT on METEOR score than SacreBLEU score.
-Language-wise, Italian, Polish, and Turkish achieves the best average SacreBLEU scores while Chinese, Polish and Turkish are shown as the top three in average METEOR scores.
+It can be seen that mBART-50 performs more closely to OPUS-MT on the METEOR score than the SacreBLEU score.
+Language-wise, Italian, Polish, and Turkish achieve the best average SacreBLEU scores while Chinese, Polish, and Turkish are shown as the top three in average METEOR scores.
 
@@
 
-Figure \ref{fig:opus_comparison} visualises the difference between each OPUS-MT models reported BLEU scores compared to the SacreBLEU result of this study.
+Figure \ref{fig:opus_comparison} visualises the difference between each OPUS-MT model's reported BLEU scores compared to the SacreBLEU result of this study.
 The reported BLEU is evaluated on the Tatoeba dataset \cite{tiedemann-2020-opus-mt}.
-It can be seen that the scores are quite aligned, with Ukrainian and Italian reaching the two highest score while Chinese, Polish, and Japanese suffers.
+It can be seen that the scores are quite aligned, with Ukrainian and Italian reaching the two highest scores while Chinese, Polish, and Japanese suffer.
 
-This study experiments also resulted in moderately higher overall SacreBLEU score, however, the dataset used here is also significantly less in size than the full Tatoeba challenge dataset \cite{tiedemann-2020-tatoeba-challenge}.
+This study's experiments also resulted in a moderately higher overall SacreBLEU score, however, the dataset used here is also significantly less in size than the full Tatoeba challenge dataset \cite{tiedemann-2020-tatoeba-challenge}.
 
@@
     \centering
-    \begin{tabular}{|l|r|r|}
+    \begin{tabular}{l r r}
         \hline
@@
 
-Additionally, Table \ref{table:opus_comparison} shows the full scores of both results SacreBLEU and reported BLEU from each OPUS-MT models.
-The reported BLEU scores are taken from each model cards in the HuggingFace repository \cite{huggingface}.
+Additionally, Table \ref{table:opus_comparison} shows the full scores of both results SacreBLEU and reported BLEU from each OPUS-MT model \cite{tiedemann-2020-opus-mt}.
+The reported BLEU scores are taken from each model card in the HuggingFace repository \cite{huggingface}.
+Unfortunately, other models are not benchmarked on the Tatoeba dataset, not evaluated on the source languages used in this study, and use different metrics \cite{liu-2020-mbart,fan-2020-m2m100,nllb200-2020}, and therefore cannot be directly compared to the results of this study.
 
@@
 
-Lastly, Figure \ref{fig:inference_timings} shows the time taken to translate for each language across experimented models.
-The OPUS-MT model is the fastest among the other three multilingual models, which perform similarly in timings.
-It is quite unclear why mBART50 Spanish-to-English translation suffers from a significantly higher timing.
+Lastly, Figure \ref{fig:inference_timings} shows the time taken to translate all 1,323 sentences for each language across experimented models.
+The OPUS-MT model is the fastest among the other three multilingual models.
+M2M-100 and NLLB-200 achieve similar timings, while mBart-50 has slightly longer timings on average.
+It is unknown why mBART50 suffers from a significantly higher timing on the Spanish-to-English translation.
 Re-running the experiment shows that this timing is consistent and not an anomaly.
@@
 
-This study findings show that while multilingual models have been shown to achieve higher performance compared to one-to-one translation models \cite{liu-2020-mbart}, this does not hold on short sentences, many-to-English translations across the 14 source languages.
+This study's findings show that while multilingual models have been shown to achieve higher performance compared to one-to-one translation models \cite{liu-2020-mbart}, this does not hold on short sentences, many-to-English translations across the 14 source languages, with regard to unfine-tuned multilingual models in this study.
 
@@
 mBART-50 is the second-best-performing model, beating OPUS-MT only in Polish SacreBLEU.
-Nevertheless, both mBART-50 and M2M-100 performs generally well on all languages.
+Nevertheless, both mBART-50 and M2M-100 perform generally well in all languages.
 NLLB-200, however, achieves much lower scores in both SacreBLEU and METEOR for all languages except for Chinese, slightly outperforming M2M-100 but still slightly below mBART-50.
+NLLB-200 should then be avoided for use without any further fine-tuning on these languages.
+Bear in mind that the distilled version is used in this study instead of the larger 3.3 billion parameters model variant.
 
-NLLB-200 should then be avoided for use without any further fine-tuning on these 14 languages.
-
-Chinese, Italian, Polish, Turkish have the best average scores over all four models, Ukrainian by OPUS-MT has the highest SacreBLEU and METEOR score across all the other models and languages.
-
-Limitations, only the first translation of the same phrases are taken, which may disrupts the possibility for synonyms, especially for BLEU and SacreBLEU metrics.
-Additionally, no fine-tuning is applied in this experiment, which may affect the translation results, as fine-tuning models for specific languages can significantly improve models performance \cite{zhang-2023-fine-tuning}.
+Chinese, Italian, Polish, and Turkish have the best average scores over all four models.
+Ukrainian by OPUS-MT has the highest SacreBLEU and METEOR scores across all the other models and languages.
+These results show that these languages are well-supported on the four pre-trained models (PTMs) evaluated in this study, achieving decent results without any fine-tuning.
 
@@
 
-Future works should be done by incorporating more Pre-Trained Models (PTMs) and implementing fine-tuning for specific languages on languages-to-English translation.
-Evaluating with bigger datasets and longer texts can also be beneficial, as well as using language other than English as the target language.
+When the dataset is curated, only the first translation of the same English phrases is taken, the others are discarded.
+This ignores several possible translation references which may include synonyms or different phrasings, thus affecting the evaluation score, especially for BLEU and SacreBLEU metrics.
 
-Furthermore, it may be beneficial to consider more general multilingual PTMs such as mBERT \cite{wu-2020-mbert-are-all} and PolyLM \cite{wei-2023-polylm}, which are trained for a wide arrange of NLP tasks instead of just machine translation.
+The dataset consists of mostly short sentences.
+Whether the result of this study stays consistent for longer sentences or documents is not known and should be further looked into.
+Moreover, the linguistic depth of the sentences is considered low and does not require deep understanding.
+
+As Tatoeba \cite{tatoeba} is run mostly by volunteers, there are also several limitations with its corpus.
+The quality of sentences can vary significantly depending on the contributors' language proficiency and the level of community engagement for specific languages.
+Moreover, not all sentences are reviewed or verified by native speakers or trusted users.
+
+The languages chosen are mostly high-resource languages, while models such as NLLB-200 \cite{nllb200-2020} are specifically designed to address the problem with low-resource languages.
+Data leakage is also a possibility, as sentences included in this dataset are fairly common phrases, it is highly likely that the model has seen these during training.
+
+Future works should incorporate more Pre-Trained Models (PTMs) and implement fine-tuning multilingual models for specific languages on languages-to-English translation as it can significantly improve performance \cite{zhang-2023-fine-tuning}.
+Furthermore, it may be beneficial to consider more general multilingual PTMs such as mBERT \cite{wu-2020-mbert-are-all} and PolyLM \cite{wei-2023-polylm}, which are trained for a wide range of NLP tasks instead of just machine translation.
 

### 2024-08-19 analyse/analyse_sentences.py

@@ -65,7 +65,7 @@ def analyse_sentence_length(df):
 
 def analyse_sentence_length_2(df):
     plt.clf()
-    plt.figure(figsize=(15, 6))
+    plt.figure(figsize=(16, 8))
 
     # Prepare data for box plot
     lengths = []
@@ -88,6 +88,8 @@ def analyse_sentence_length_2(df):
     plt.title("Word Count Distribution")
     plt.ylabel("Word Count")
     plt.xlabel(None)
+
+    plt.tight_layout()
 
     plt.savefig("./figures/word_count_box.png")
 
@@ -143,6 +145,6 @@ def analyse_embedding(df):
 
 
 # analyse_sentence_length(sentences_df)
-# analyse_sentence_length_2(sentences_df)
+analyse_sentence_length_2(sentences_df)
 # generate_word_cloud(sentences_df)
-analyse_embedding(sentences_df)
+# analyse_embedding(sentences_df)

### 2024-08-19 paper/paper.tex
updated, now 16 pages

@@
 
-This study's experiments also resulted in a moderately higher overall SacreBLEU score, however, the dataset used here is also significantly less in size than the full Tatoeba challenge dataset \cite{tiedemann-2020-tatoeba-challenge}.
+While this study's experiments resulted in a moderately higher overall SacreBLEU, the dataset used here is also significantly smaller than the full Tatoeba challenge dataset \cite{tiedemann-2020-tatoeba-challenge} that are used in the OPUS-MT reported BLEU.
+In contrast, this study employs the same set of parallel sentences for all languages, which should lead to more consistent results.
+Nevertheless, a direct comparison between these results is not fully pertinent.
 
@@
 
-The dataset consists of mostly short sentences.
+The dataset consists of mostly short sentences, with little or none linguistic depth.
 Whether the result of this study stays consistent for longer sentences or documents is not known and should be further looked into.
-Moreover, the linguistic depth of the sentences is considered low and does not require deep understanding.
+The Tatoeba dataset \cite{tiedemann-2020-tatoeba-challenge} consisting of short sentences has been reported to be too optimistic when evaluated with other datasets that are more realistic \cite{tiedemann-2020-opus-mt}.
 

### 2024-08-20 paper/paper.tex
updated

@@
 
-\title{Evaluation of Pre-Trained Models for Many-to-English Translation}
+\title{Evaluation of Machine Translation Pre-Trained Models for Many-to-English Translation}
 \author{Stefan Liemawan Adji}
@@
     Machine translation (MT) has become an essential tool for overcoming language barriers in an increasingly globalised world.
-Today, there are pre-trained models (PTMs) specifically designed for automatic machine translation.
-This paper evaluates the performance of various pre-trained models (PTMs) for many-to-English translation.
+Today, there are pre-trained models specifically designed for automatic machine translation.
+This paper evaluates the performance of various machine translation pre-trained models (MT-PTMs) for many-to-English translation.
 The models assessed include OPUS-MT \cite{tiedemann-2020-opus-mt}, mBART-50 \cite{liu-2020-mbart}, NLLB-200 \cite{nllb200-2020}, and M2M-100 \cite{fan-2020-m2m100}.
 Using a dataset curated from the Tatoeba repository, comprising 1,323 parallel sentence pairs across 14 source languages, translation quality is measured using BLEU, SacreBLEU, and METEOR scores.
-The results indicate that OPUS-MT, a one-to-one PTM, outperforms multilingual models in one-to-English translation tasks, with mBART-50 leading among the multilingual PTMs.
+The results indicate that OPUS-MT, a one-to-one PTM, outperforms multilingual models in one-to-English translation tasks, with mBART-50 leading among the multilingual models.
 Chinese, Italian, Polish, and Turkish translations to English show the best performance on average across the four models used in this study, however, Ukrainian-to-English on the OPUS-MT model achieves the highest SacreBLEU and METEOR scores.
@@
 This study will lean towards evaluating MT-specific models rather than general LLMs.
-Through simple experimentations, this paper aims to evaluate the performance of existing and publicly available MT pre-trained models (PTMs) on many-to-English translation across 14 source languages.
-Although fine-tuning multilingual PTMs has been proven to increase translation performance \cite{cooper-stickland-2021-recipes}, no pre-training or fine-tuning is performed in this study due to simplicity reasons and resource limitations.
+Through simple experimentations, this paper aims to evaluate the performance of existing and publicly available MT pre-trained models (MT-PTMs) on many-to-English translation across 14 source languages.
+Although fine-tuning multilingual MT-PTMs has been proven to increase translation performance \cite{cooper-stickland-2021-recipes}, no pre-training or fine-tuning is performed in this study due to simplicity reasons and resource limitations.
 A dataset is curated from the Tatoeba repository \cite{tatoeba}, containing 1,323 parallel sentence pairs across source and target languages.
-The models include a one-to-one PTMs: OPUS-MT \cite{tiedemann-2023-democratizing,tiedemann-2020-opus-mt}, and multilingual PTMs: such as mBART-50 \cite{liu-2020-mbart}, NLLB-200 \cite{nllb200-2020}, and M2M-100 \cite{fan-2020-m2m100}.
+The models include a one-to-one PTMs: OPUS-MT \cite{tiedemann-2023-democratizing,tiedemann-2020-opus-mt}, and multilingual MT-PTMs: such as mBART-50 \cite{liu-2020-mbart}, NLLB-200 \cite{nllb200-2020}, and M2M-100 \cite{fan-2020-m2m100}.
 The performance of these models is evaluated using the BLEU \cite{papieni-2002-bleu}, SacreBLEU \cite{post-2018-sacrebleu}, and METEOR \cite{lavie-2007-meteor}.
 
-The result indicates that OPUS-MT outperforms all other multilingual models in one-to-English translation on both SacreBLEU and METEOR scores, highlighting the superiority of specific languages PTMs compared to unfine-tuned multilingual PTMs.
+The result indicates that OPUS-MT outperforms all other multilingual models in one-to-English translation on both SacreBLEU and METEOR scores, highlighting the superiority of specific languages MT-PTMs compared to unfine-tuned multilingual MT-PTMs.
 Among the multilingual models, mBART-50 shows the best performance, while NLLB-200 performs the worst, by a significant margin.
@@
 
-Currently, pre-trained models (PTMs) have become a cornerstone in machine translation, leveraging large datasets and powerful architectures to achieve state-of-the-art performance \cite{han-2021-ptms}.
 Attention mechanism has been used for Neural Machine Translation (NMT) \cite{bahdanau-2016-nmt-jointly, luong-2015-effective-attention} even before the introduction of the transformer \cite{vaswani-2017-attention}.
-The transformer, with its encoder-decoder approach \cite{cho-2014-properties}, remains the foundational architecture for many sequence-to-sequence models in machine translation.
-Most PTMs support two kinds of translations: one-to-one translation and many-to-many translation (multilingual models).
+The Transformer, with its encoder-decoder architecture \cite{vaswani-2017-attention}, has since become the foundational framework for many sequence-to-sequence models in machine translation \cite{cho-2014-properties}.
+By combining this architecture with training on large datasets, models were able to capture and retain vast amounts of linguistic information, leading to the development of pre-trained models.
+
+Currently, pre-trained models have become a cornerstone in machine translation, achieving state-of-the-art performances \cite{han-2021-ptms}.
+While general Large Language Models (LLMs) such as BERT \cite{devlin-2019-bert}, GPT \cite{openai-2024-gpt4}, or LLAMA \cite{touvron-2023-llama2} can be used for machine translation tasks, in this study the focus will be towards MT-specific PTMs (MT-PTMs) which are specifically designed, built, and trained for machine translation tasks.
+Most MT-PTMs typically support two kinds of translations: one-to-one translation and many-to-many translation (multilingual models).
 
@@
 This setup is characterised by having a dedicated model for each unique language pair.
-OPUS-MT \cite{tiedemann-2020-opus-mt} is a project dedicated to creating and providing free resources for machine translation, providing over a thousand PTMs of direct one-to-one translations.
-Additionally, Facebook Fairseq \cite{ott-2019-fairseq} similarly offers one-to-one PTMs, although they are only limited to translations between English, French, German, and Russian.
+OPUS-MT \cite{tiedemann-2020-opus-mt} is a project dedicated to creating and providing free resources for machine translation, providing over a thousand MT-PTMs of direct one-to-one translations.
+Additionally, Facebook Fairseq \cite{ott-2019-fairseq} similarly offers one-to-one MT-PTMs, although they are only limited to translations between English, French, German, and Russian.
 
@@
 mBART \cite{liu-2020-mbart}, M2M \cite{fan-2020-m2m100}, NLLB-200 \cite{nllb200-2020}, are examples of massively multilingual models that support many-to-many translation language.
+
+With a lot of
 
@@
 Table \ref{table:eval_languages} shows the 14 language pairs selected for this study.
-Languages are chosen based on the resources' availability in Tatoeba and supported languages in most PTMs.
+Languages are chosen based on the resources' availability in Tatoeba and supported languages in most MT-PTMs.
 Thus, the languages chosen here can be considered moderate to high-resource languages.
@@
 The architecture is based on MARIAN-NMT \cite{mariannmt}, based on a standard transformer setup: 6 self-attentive layers in both the encoder and decoder networks, each with 8 attention heads per layer \cite{tiedemann-2020-opus-mt}.
+The models were trained for up to 72 hours on 1 or 4 GPUs, noting that not all of them fully converged within this time frame.
 
-While the framework provides both monolingual and multilingual PTMs, only the OPUS-MT monolingual models are used in this study.
+While the framework provides both monolingual and multilingual models, only the OPUS-MT monolingual models are used in this study.
 This is partly due to the lack of documentation on the multilingual models.
@@
 
-OPUS-MT models are taken from the Helsinki-NLP repository with the exception of 'tiedeman/opus-mt-he-en' for Hebrew-to-English translation.
-The 'mbart-large-50-many-to-many-mmt', the 'm2m100\_418M', and the 'nllb-200-distilled-600M' models are used for mBART-50 \cite{liu-2020-mbart}, M2M-100 \cite{fan-2020-m2m100}, and NLLB-200 \cite{nllb200-2020}, respectively.
-All models are taken using the transformers \cite{wolf-etal-2020-transformers} library from Hugging Face \cite{huggingface}.
+OPUS-MT models are taken from the Helsinki-NLP repository through the Hugging Face website \cite{huggingface,zhang-2023-fine-tuning} with the exception of Hebrew-to-English models taken directly from 'tiedeman/opus-mt-he-en'.
+The 'mbart-large-50-many-to-many-mmt', the 'm2m100\_418M', and the 'nllb-200-distilled-600M' model variants are used for mBART-50 \cite{liu-2020-mbart}, M2M-100 \cite{fan-2020-m2m100}, and NLLB-200 \cite{nllb200-2020}, respectively.
+Similarly, all multilingual models are downloaded using the transformers \cite{wolf-2020-transformers} library from Hugging Face \cite{huggingface}.
 A batch size of 4 is used in all models.
-Translations are first generated and saved into CSV files, before calculating the metrics score.
+Translations from each source language to english are first generated through inference and saved into CSV files, before calculating the metrics score.
 
-For BLEU and METEOR, tokenisation and metrics calculation use the NLTK \cite{bird-2009-natural} package, specifically the \texttt{nltk.word\_tokenize}, \texttt{nltk.translate.meteor\_score}, and \texttt{nltk.translate.bleu\_score.corpus\_bleu}.
-Additionally, smoothing method 7 is used for the BLEU calculation, which interpolates two other smoothing functions: smoothing method 4, which divides by \( \frac{1}{\ln(\text{len}(T))} \), where \( T \) is the length of the translation; and smoothing method 5, which averages the \( n-1 \), \( n \), and \( n+1 \) gram matched counts \cite{nltk-bleu-docs}.
+For BLEU and METEOR, tokenisation and metrics calculation use the NLTK \cite{bird-2009-natural} package, specifically:
+\begin{itemize}
+    \item \texttt{nltk.word\_tokenize} \\
+    \item \texttt{nltk.translate.meteor\_score} \\
+    \item \texttt{nltk.translate.bleu\_score.corpus\_bleu}
+\end{itemize}
+Additionally, the smoothing method 7 is used for the BLEU calculation, which interpolates two other smoothing functions: smoothing method 4, which divides by \( \frac{1}{\ln(\text{len}(T))} \), where \( T \) is the length of the translation; and smoothing method 5, which averages the \( n-1 \), \( n \), and \( n+1 \) gram matched counts \cite{nltk-bleu-docs}.
 The \texttt{sacrebleu} \cite{post-2018-sacrebleu} package from pip is used to calculate SacreBLEU \cite{post-2018-sacrebleu}.

### 2024-08-20 paper/paper.tex
updated

@@
 The following years saw the birth of Large Language Models (LLMs) such as BERT \cite{devlin-2019-bert}, GPT \cite{openai-2024-gpt4}, and T5 \cite{raffel-2023-t5}, which revolutionised both MT and the whole field of NLP.
-Then in the early 2020s, several pre-trained models (PTM) that are specifically designed for machine translation emerged, OPUS-MT \cite{tiedemann-2020-opus-mt} produced numerous PTMs designed for one-to-one translations.
+Then in the early 2020s, several pre-trained models that are specifically designed for machine translation emerged, these will be referred in this paper as Machine Translation Pre-Trained Models (MT-PTMs).
+OPUS-MT \cite{tiedemann-2020-opus-mt} produced numerous models designed for one-to-one translations.
 Additionally, multilingual models: mBART-50 \cite{liu-2020-mbart}, NLLB-200 \cite{nllb200-2020}, and M2M-100 \cite{fan-2020-m2m100} have also been published, growing popularity as multilingualism has been shown to allow for sharing information among languages \cite{garcia-2020-multilingual}.
@@
 Despite these advancements, pre-trained models are often evaluated using a different set of languages and benchmarks \cite{liu-2020-mbart,nllb200-2020,fan-2020-m2m100,wei-2023-polylm}, making it difficult to gauge their relative effectiveness across various languages.
-Nevertheless, there does not seem to be much work on comparing or benchmarking different pre-trained models in machine translation.
-Intento recently published 'The State of Machine Translation 2024' \cite{intento-2020} providing an in-depth evaluation of 52 MT engines and LLMs across 11 language pairs.
-Intento reported a lack of growth in the number of supported languages and concluded that translation quality significantly varies depending on languages \cite{intento-2020}.
+Multilingual datasets often do not contain parallel texts across different language pairs; instead, each language pair typically has its own unique set of texts that may not overlap with the texts available in other languages \cite{tiedemann-2020-tatoeba-challenge,cettolo-2012-wit3}.
+
+Nevertheless, there does not seem to be much work on comparing or benchmarking different MT-PTMs in machine translation.
+Intento recently published 'The State of Machine Translation 2024' \cite{intento-2024} providing an in-depth evaluation of 52 MT engines and LLMs across 11 language pairs.
+Figure \ref{fig:intento_quality} shows the reported translation quality across different domains and language pairs, showing that the results significantly varies based on these parameters.
+They also reported a lack of growth in the number of supported languages and concluded that translation quality significantly varies depending on languages \cite{intento-2024}.
 However, the biggest drawback in this report is that the selected LLMs are general LLMs such as GPT-4 \cite{openai-2024-gpt4}, LLaMa2 \cite{touvron-2023-llama2}, and Mistral \cite{jiang-2023-mistral-7b}, instead of MT-specific models.
+
+\begin{figure}[htbp]
+    \centering
+    \includegraphics[width=1\linewidth]{images/intento_quality.png}
+    \caption{Quality of MT engines and LLMs by domains (y-axis) and language pairs \cite{intento-2024}}
+    \label{fig:intento_quality}
+\end{figure}
 
@@
 
-With a lot of
-
 \subsection{Parallel Corpora}
@@
 Furthermore, papers often use different datasets and versions to evaluate their model performance, making direct comparisons challenging.
-mBart \cite{liu-2020-mbart} uses WMT19 for English-German translation and TED15 for Chinese-English translation.
+mBART \cite{liu-2020-mbart} uses WMT19 for English-German translation and TED15 for Chinese-English translation.
 M2M-100 \cite{fan-2020-m2m100} uses seven different datasets shared across languages.
@@
 
-Similarly, SacreBLEU for Italian-to-English translation outperforms other languages with the M2M-100 model, although Dutch beat Italian by 0.01 in METEOR score.
+Similarly, SacreBLEU for Italian-to-English translation outperforms other languages with the M2M-100 model, although Dutch-to-English surpasses Italian-to-English by 0.01 in METEOR score.
 Japanese received the worst score in both SacreBLEU and METEOR for this model.
@@
     \includegraphics[width=1\linewidth]{figures/metrics_bar.png}
-    \caption{BLEU, SacreBLEU, and METEOR scores of each model on each language}
+    \caption{BLEU, SacreBLEU, and METEOR scores of each model on each language pair.}
     \label{fig:result_visual}
@@
 
-Figure \ref{fig:result_visual} visualises all scores gained by each model, categorised by language.
-
+The BLEU, SacreBLEU, and METEOR scores gained by each model, categorised by every language pairs are illustrated in Figure \ref{fig:result_visual}.
 Dutch, Italian, and Ukrainian translations generally score higher across all metrics, suggesting that these languages are well-supported by the pre-trained models.
 Japanese translations, however, pose significant challenges for all models, as indicated by the consistently low scores across all metrics.
-
 Japanese-to-English translation seems to have the worst score on average across the four models, but particularly bad on multilingual models.
@@
 
-Figure \ref{fig:average_bars} visualise the average BLEU, SacreBLEU, and METEOR, for each model and each language, highlighting how the models and languages perform independently.
-It can be seen that mBART-50 performs more closely to OPUS-MT on the METEOR score than the SacreBLEU score.
-Language-wise, Italian, Polish, and Turkish achieve the best average SacreBLEU scores while Chinese, Polish, and Turkish are shown as the top three in average METEOR scores.
+Visualisations of the average SacreBLEU and METEOR, categorised by model (left figures) and language (right figures) is shown in Figure \ref{fig:average_bars}, providing a closer insight on how each model and language perform independently.
+It can be seen that when averaged across all languages, mBART-50 performs more closely to OPUS-MT on the METEOR score than the SacreBLEU score.
+Language-wise, Italian, Polish, and Turkish achieve the best average SacreBLEU scores while Chinese, Polish, and Turkish are the top three languages with the highest average METEOR scores.
 
@@
 
-Figure \ref{fig:opus_comparison} visualises the difference between each OPUS-MT model's reported BLEU scores compared to the SacreBLEU result of this study.
+Figure \ref{fig:opus_comparison} highlights the difference between each OPUS-MT model's reported BLEU scores compared to the SacreBLEU result of this study.
 The reported BLEU is evaluated on the Tatoeba dataset \cite{tiedemann-2020-opus-mt}.
@@
 The OPUS-MT model is the fastest among the other three multilingual models.
-M2M-100 and NLLB-200 achieve similar timings, while mBart-50 has slightly longer timings on average.
+M2M-100 and NLLB-200 achieve similar timings, while mBART-50 has slightly longer timings on average.
 It is unknown why mBART50 suffers from a significantly higher timing on the Spanish-to-English translation.
@@
 mBART-50 is the second-best-performing model, beating OPUS-MT only in Polish SacreBLEU.
-Nevertheless, both mBART-50 and M2M-100 perform generally well in all languages.
+Nevertheless, both the mBART-50 and the M2M-100 models perform generally well in all languages.
 NLLB-200, however, achieves much lower scores in both SacreBLEU and METEOR for all languages except for Chinese, slightly outperforming M2M-100 but still slightly below mBART-50.
 NLLB-200 should then be avoided for use without any further fine-tuning on these languages.
-Bear in mind that the distilled version is used in this study instead of the larger 3.3 billion parameters model variant.
+Note that that the distilled version is used in this study for NLLB-200 instead of the larger 3.3 billion parameters model variant.
 
@@
 Ukrainian by OPUS-MT has the highest SacreBLEU and METEOR scores across all the other models and languages.
-These results show that these languages are well-supported on the four pre-trained models (PTMs) evaluated in this study, achieving decent results without any fine-tuning.
+These results show that these languages are well-supported within the four (MT-PTMs) evaluated in this study, achieving decent results without any fine-tuning.
 

### 2024-08-22 paper/paper.tex
updated

@@
 The following years saw the birth of Large Language Models (LLMs) such as BERT \cite{devlin-2019-bert}, GPT \cite{openai-2024-gpt4}, and T5 \cite{raffel-2023-t5}, which revolutionised both MT and the whole field of NLP.
-Then in the early 2020s, several pre-trained models that are specifically designed for machine translation emerged, these will be referred in this paper as Machine Translation Pre-Trained Models (MT-PTMs).
+Then in the early 2020s, several pre-trained models that are specifically designed for machine translation emerged, these will be referred to in this paper as Machine Translation Pre-Trained Models (MT-PTMs).
 OPUS-MT \cite{tiedemann-2020-opus-mt} produced numerous models designed for one-to-one translations.
@@
 Intento recently published 'The State of Machine Translation 2024' \cite{intento-2024} providing an in-depth evaluation of 52 MT engines and LLMs across 11 language pairs.
-Figure \ref{fig:intento_quality} shows the reported translation quality across different domains and language pairs, showing that the results significantly varies based on these parameters.
+Figure \ref{fig:intento_quality} shows the reported translation quality across different domains and language pairs, showing that the results significantly vary based on these parameters.
 They also reported a lack of growth in the number of supported languages and concluded that translation quality significantly varies depending on languages \cite{intento-2024}.
@@
 
-Corpora are large and structured sets of texts used for linguistic research and analysis.
+Corpora simply means a collection of texts, which in this case are used for the purpose of training models.
 Thus, parallel corpora are defined as sets of texts in a given source language along with their translations in another target language \cite{lefer-2020-parallel-corpora}.
-It can be bilingual or multilingual and is crucial for training and evaluating machine translation.
+They can be bilingual or multilingual and this is crucial for training and evaluating machine translation.
 Languages without or with limited parallel corpora are referred to as low-resource languages.
 
-Since the early 2000s, English has often been used as an intermediary language, meaning texts originally written in languages are first translated into English, and then from English into other languages, resulting in most target texts being translations of translations \cite{lefer-2020-parallel-corpora}.
+Particularly for these low-resource languages, where there are not enough parallel corpora, English has often been used as an intermediary or 'pivot' language.
+This means texts originally written in languages are first translated into English, and then from English into other languages, resulting in most target texts being translations of translations \cite{lefer-2020-parallel-corpora}.
 This is called indirect translation and poses many problems in translations as terms can be ambiguous with varying or similar meanings \cite{rosa-2017-indirect-translation-problems}.
 
-Neural Machine Translation (NMT) systems require vast amounts of training data, and thus the availability of parallel corpora is crucial for building effective models \cite{koehn-2017-challenges}.
-The lack of extensive parallel corpora, especially for low-resource languages, leads to suboptimal performance in NMT techniques compared to their high-resource counterparts \cite{ranathunga-2023-nmt-low-res}.
-OPUS \cite{opus} is a comprehensive collection of open-source parallel corpora used extensively in the field of machine translation (MT).
+Since Neural Machine Translation (NMT) systems require vast amounts of training data, the availability of parallel corpora is crucial for building effective models \cite{koehn-2017-challenges}.
+The lack of extensive parallel corpora leads to suboptimal performance in NMT techniques \cite{ranathunga-2023-nmt-low-res}.
+Today, numerous websites, repositories, and datasets exist containing large amounts of parallel texts between specific languages.
+
+OPUS \cite{tiedemann-2012-opus-parallel} provides a comprehensive collection of open-source parallel corpora dedicated to machine translation (MT) tasks.
 It includes corpora for 744 languages and contains over 1,210 different datasets, amassing a total of 45,945,946,108 sentence pairs.
@@
 
-Several datasets are often used for evaluation in machine translation.
+A number of renowned datasets are commonly used for evaluation in machine translation.
 The Tatoeba Challenge \cite{tiedemann-2020-tatoeba-challenge} covers 487 languages in 4,024 language pairs, including 657 test sets sourced from the Tatoeba website, covering 138 languages.
@@
 
-While these datasets can be used to evaluate translation performance between any language pairs, they do not contain parallel corpora between multiple different languages.
+While these datasets can be used to evaluate translation performance between any language pairs, they mostly do not contain parallel corpora across different languages.
 A corpus between English and French, for example, contains different sets of text than a corpus between English and Spanish.
 Therefore, they do not support easy evaluation for many-to-English translations.
-Furthermore, papers often use different datasets and versions to evaluate their model performance, making direct comparisons challenging.
+Furthermore, research papers often use different datasets and versions to evaluate their model performance, making direct comparisons even more challenging.
 mBART \cite{liu-2020-mbart} uses WMT19 for English-German translation and TED15 for Chinese-English translation.
@@
 
-% The original paper is 192 pages long and impossible to be dissected in just one section.
-
 \subsection{Evaluation Metrics}
@@
 Bilingual Evaluation Understudy (BLEU) \cite{papieni-2002-bleu} is the most commonly used metric for machine translation (MT).
-It assesses how well a candidate translation matches the reference translation using precision metrics for n-grams and incorporates a brevity penalty to prevent overly short translations from achieving high scores.
+It assesses how well a candidate translation matches the reference translation by using precision metrics for n-grams and a brevity penalty to ensure a good match in regard to length, word choice, and word order.
 
@@
 
-The machine translation community relies heavily upon the BLEU score, despite its several drawbacks.
+The machine translation community relies heavily upon the BLEU score, despite its several drawbacks \cite{lommel-2016-blues}.
 The metric has been reported to not correlate strongly with human judgement, showing variations in translation that could mean that a higher BLEU score does not necessarily indicate a true enhancement in translation quality \cite{callison-burch-2006-reevaluating-bleu}.
@@
 where \(P\) is the precision, \(R\) is the recall, and \(\text{frag}\) is the fragmentation penalty.
-\(\gamma\) is a parameter that controls the weight of the fragmentation penalty, commonly set to 0.5, and \(\alpha\) is a parameter that controls the balance between precision and recall, commonly set to 0.9,
+\(\gamma\) is a parameter that controls the weight of the fragmentation penalty, commonly set to 0.5, and \(\alpha\) is a parameter that controls the balance between precision and recall, commonly set to 0.9.
+
+Although the latest official version of METEOR (METEOR-1.5) was released in 2014 \cite{denkowski-lavie-2014-meteor-universal}, the NLTK library \cite{bird-2009-natural} still implements the older version based on the 2007 paper by Lavie and Agarwal \cite{lavie-2007-meteor}.
+Therefore, this study will use the older version of METEOR, as the calculations will rely on the NLTK library.
 
@@
 A batch size of 4 is used in all models.
-Translations from each source language to english are first generated through inference and saved into CSV files, before calculating the metrics score.
+Translations from each source language to English are first generated through inference and saved into CSV files, before calculating the metrics score.
 
@@
 Table \ref{table:opus_result}, \ref{table:mbart_result}, \ref{table:m2m_result}, and \ref{table:nllb_result}, and  shows the result from all four PTMs.
-All models perform similarly low on BLEU score across all languages, with NLLB-200 reaching 0.01 below on Japanese translation.
-The BLEU scores are very low (below 10 BLEU) in all cases due to the sentences being short (find cite).
-This also highlights the superiority of SacreBLEU compared to the flawed, standard BLEU.
-
-% BLEU is notoriously unreliable for very short texts.
-The score will be disproportionately influenced by small differences if there isn't much text to begin with because there is a Brevity Penalty set in place to discourage the model from outputting fewer words and getting a high score.
+All models perform similarly low on BLEU score (below 10) across all languages, with NLLB-200 reaching 0.01 below on Japanese translation.
+As BLEU is notoriously unreliable for very short texts, these results are not surprising.
+The BLEU score can be disproportionately affected by small differences in shorter texts due to the limited number of n-grams available for comparison, as well as the brevity penalty \cite{chatgpt-2024-bleu,papieni-2002-bleu}.
 
@@
 
-The BLEU, SacreBLEU, and METEOR scores gained by each model, categorised by every language pairs are illustrated in Figure \ref{fig:result_visual}.
+The BLEU, SacreBLEU, and METEOR scores gained by each model, categorised by every language pair are illustrated in Figure \ref{fig:result_visual}.
 Dutch, Italian, and Ukrainian translations generally score higher across all metrics, suggesting that these languages are well-supported by the pre-trained models.
@@
 
-Visualisations of the average SacreBLEU and METEOR, categorised by model (left figures) and language (right figures) is shown in Figure \ref{fig:average_bars}, providing a closer insight on how each model and language perform independently.
+Visualisations of the average SacreBLEU and METEOR, categorised by the model (left figures) and language (right figures) are shown in Figure \ref{fig:average_bars}, providing a closer insight into how each model and language perform independently.
 It can be seen that when averaged across all languages, mBART-50 performs more closely to OPUS-MT on the METEOR score than the SacreBLEU score.
@@
     \centering
-    \includegraphics[width=0.8\linewidth]{figures/opus_bleu_comparison.png}
+    \includegraphics[width=0.9\linewidth]{figures/opus_bleu_comparison.png}
     \caption{Comparison between reported OPUS-MT models performance and performance in this experiment.}
@@
 
-While this study's experiments resulted in a moderately higher overall SacreBLEU, the dataset used here is also significantly smaller than the full Tatoeba challenge dataset \cite{tiedemann-2020-tatoeba-challenge} that are used in the OPUS-MT reported BLEU.
+While this study's experiments resulted in a moderately higher overall SacreBLEU, the dataset used here is also significantly smaller than the full Tatoeba challenge dataset \cite{tiedemann-2020-tatoeba-challenge} that is used in the OPUS-MT reported BLEU.
 In contrast, this study employs the same set of parallel sentences for all languages, which should lead to more consistent results.
@@
 The reported BLEU scores are taken from each model card in the HuggingFace repository \cite{huggingface}.
-Unfortunately, other models are not benchmarked on the Tatoeba dataset, not evaluated on the source languages used in this study, and use different metrics \cite{liu-2020-mbart,fan-2020-m2m100,nllb200-2020}, and therefore cannot be directly compared to the results of this study.
+Unfortunately, other models are either not benchmarked on the Tatoeba dataset, not evaluated on the source languages used in this study, or use different metrics \cite{liu-2020-mbart,fan-2020-m2m100,nllb200-2020}, and therefore cannot be directly compared to the results of this study.
 
@@
     \centering
-    \includegraphics[width=0.9\linewidth]{figures/inference_timings.png}
+    \includegraphics[width=1\linewidth]{figures/inference_timings.png}
     \caption{Inference timings for every model.}
@@
 
-This study's findings show that while multilingual models have been shown to achieve higher performance compared to one-to-one translation models \cite{liu-2020-mbart}, this does not hold on short sentences, many-to-English translations across the 14 source languages, with regard to unfine-tuned multilingual models in this study.
+This study created a simple dataset containing 1,323 parallel sentence pairs in 14 moderate-to-high-resource languages and evaluated several monolingual and multilingual Machine Translation Pre-Trained Models (MT-PTMs).
+The findings suggest that while multilingual models have been shown to outperform one-to-one translation models \cite{liu-2020-mbart}, this premise does not necessarily hold in our experiments.
+Specifically, with a small dataset consisting of short sentences, many-to-English translations across 14 languages with monolingual MT-PTMs (OPUS-MT) outperform unfine-tuned multilingual MT-PTMs (mBART-50, M2M-100, NLLB-200).
 
@@
 NLLB-200 should then be avoided for use without any further fine-tuning on these languages.
-Note that that the distilled version is used in this study for NLLB-200 instead of the larger 3.3 billion parameters model variant.
+Note that the distilled version is used in this study for NLLB-200 instead of the larger 3.3 billion parameters model variant.
 
@@
 
-When the dataset is curated, only the first translation of the same English phrases is taken, the others are discarded.
-This ignores several possible translation references which may include synonyms or different phrasings, thus affecting the evaluation score, especially for BLEU and SacreBLEU metrics.
+In curating the dataset, only the first translation of the same English phrases was taken, while additional translations were discarded.
+This ignores several possible translation references which might include synonyms or different phrasings, thus potentially affecting the evaluation score, especially for BLEU and SacreBLEU metrics.
+Moreover, the dataset mainly consists of short sentences with limited linguistic depth.
+Whether the result of this study would stay consistent with longer sentences or more complex documents should be further explored.
+The Tatoeba dataset \cite{tiedemann-2020-tatoeba-challenge} consisting of short sentences has been reported to be too optimistic when evaluated with other datasets that are more realistic \cite{tiedemann-2020-opus-mt}.
 
-The dataset consists of mostly short sentences, with little or none linguistic depth.
-Whether the result of this study stays consistent for longer sentences or documents is not known and should be further looked into.
-The Tatoeba dataset \cite{tiedemann-2020-tatoeba-challenge} consisting of short sentences has been reported to be too optimistic when evaluated with other datasets that are more realistic \cite{tiedemann-2020-opus-mt}.
+The METEOR implementation in this study through the NLTK library is based on the original METEOR paper \cite{lavie-2007-meteor}, which is described as "relatively simple and naive." A newer version, METEOR-1.5 \cite{denkowski-lavie-2014-meteor-universal} offers improved features and has been reported to produce significantly different results \cite{tuetschek-2021-meteor-issue-nltk}.
+Therefore, reevaluating translation models using the updated metric could be beneficial.
 

### 2024-08-22 paper/paper.tex
finalised

@@
 All the sentences seem to be straightforward and literal, without any use of linguistic features such as metaphors or sarcasm.
-Figure \ref{table:parallel_sentence} shows a couple of examples of parallel sentences in the final dataset, across all 14 languages and the original English sentence.
+A word cloud is illustrated in Figure \ref{fig:wordcloud}, while Figure \ref{table:parallel_sentence} shows a couple of examples of parallel sentences in the final dataset, across all 14 languages and the original English sentence.
+
+\begin{figure}[htbp]
+    \centering
+    \includegraphics[width=0.9\linewidth]{figures/wordcloud.png}
+    \caption{Dataset word cloud illustrated.}
+    \label{fig:wordcloud}
+\end{figure}
 
@@
 The METEOR implementation in this study through the NLTK library is based on the original METEOR paper \cite{lavie-2007-meteor}, which is described as "relatively simple and naive." A newer version, METEOR-1.5 \cite{denkowski-lavie-2014-meteor-universal} offers improved features and has been reported to produce significantly different results \cite{tuetschek-2021-meteor-issue-nltk}.
-Therefore, reevaluating translation models using the updated metric could be beneficial.
+Therefore, re-evaluating translation models using the updated metric might be beneficial.
+Experimenting with other metrics such as chrF \cite{popovic-2015-chrf}, which is widely used by Meta AI \cite{nllb200-2020} could also provide a more comprehensive evaluation.
 
